---
title: "Neemias Queta: Flourishing in TD Garden (2025–26)"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
params:
  season: "2025-26"
  player_name: "Neemias Queta"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center"
)
```

```{r libraries}
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(showtext)
library(sysfonts)
library(ggrepel)
library(glue)
library(hoopR)
library(ggimage)
library(janitor)
library(ragg)
```

```{r constants-and-parameters}
# Report parameters
season_used <- params$season
target_name <- params$player_name

# Celtics palette
celtics_green <- "#007A33"
celtics_gold  <- "#BA9653"
celtics_black <- "#000000"

# Theme constants
bg_color   <- "#f6f4ef"
text_dark  <- "#141414"
grid_color <- "#dedad3"

font_add_google("Poppins", "poppins")
showtext_auto()
```

```{r helper-functions}
theme_caleb_celtics <- function() {
  theme_minimal(base_family = "poppins") +
    theme(
      plot.background  = element_rect(fill = bg_color, color = NA),
      panel.background = element_rect(fill = bg_color, color = NA),
      panel.grid.major = element_line(color = grid_color, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(
        size = 18,
        face = "bold",
        hjust = 0.5,
        color = text_dark
      ),
      plot.subtitle    = element_text(
        size = 10,
        hjust = 0.5,
        color = "#2f2f2f",
        margin = margin(b = 6)
      ),
      axis.title       = element_text(size = 12, face = "bold", color = text_dark),
      axis.text        = element_text(size = 10, color = "#333333"),
      legend.position  = "none",
      plot.caption     = element_text(
        size = 8,
        color = "#6e6e6e",
        hjust = 0,
        margin = margin(t = 6)
      ),
      plot.margin      = margin(10, 12, 14, 12)
    )
}

style_latex_table <- function(table_obj, font_size = 9, scale_down = FALSE) {
  latex_options <- c("striped", "hold_position")

  if (scale_down) {
    latex_options <- c(latex_options, "scale_down")
  }

  table_obj %>%
    kableExtra::kable_styling(
      latex_options = latex_options,
      font_size = font_size
    )
}
```

# Executive Summary

Neemias Queta has undergone a dramatic role shift this season. After being released by Sacramento in late 2023, he found a home in Boston on a two-way contract, where his 2024–25 campaign showed flashes of value as a rim protector, screener, and play finisher in limited minutes. At the time, his impact was constrained by frontcourt depth behind Kristaps Porziņģis, Al Horford, and Luke Kornet, and by inconsistencies that capped his usage.

That context changed in 2025–26. With all three veterans departing, Queta was pushed into a starting role and asked to absorb real minutes, tougher matchups, and sustained responsibility. Rather than being exposed by the jump in role, he has met the demands placed on him. In a season defined by frontcourt uncertainty and selective trust, Queta has earned consistent minutes while other options have cycled through short stints and DNPs.

His value is rooted in play finishing, vertical rim pressure, and rim protection outcomes that hold up at starter usage. That skill set also scales cleanly: it works if he remains the starting center, and it remains valuable if Boston pursues a higher-end option at the position. Either way, Queta has established himself as a reliable, winning piece rather than a stopgap solution.

```{r data-loading}
# Game-level net rating and minutes.
# These values are entered manually from game-level on/off data.
queta_games <- tibble(
  game = 1:33,
  minutes = c(
    25, 22, 27, 22, 26, 20, 16, 26, 24, 30, 26, 29, 20, 34, 23, 29,
    5, NA, 29, NA, 21, 21, 26, 34, 25, 35, 25, 22, 24, 21, 29, 29, 23
  ),
  net_rtg = c(
     8.9,  0.4, 19.5, 27.9, 44.2, -7.3, -22.6, 31.3, 47.8, 14.8,
    -2.4, -2.1, 62.6, 18.6, 18.5, -2.7,  -6.8,  NA,  -2.0,  NA,
   -26.8, 63.8, 15.1, 21.6, -23.4, 14.1,   1.8, -25.1, -6.5, -3.8,
     3.3, -10.7, -4.3
  )
) %>%
  mutate(
    status = case_when(
      is.na(minutes) ~ "DNP",
      game == 17 ~ "Left early",
      TRUE ~ "Played"
    )
  )

# Offensive play types from Synergy / SportsRadar.
pt_off <- tibble::tribble(
  ~playtype, ~poss, ~pct_time, ~pct_time_rank, ~pts, ~ppp, ~ppp_rank, ~ppp_rating,
  ~fg_att, ~fg_made, ~fg_miss, ~fg_pct, ~efg_pct, ~ts_pct, ~tsa, ~to_pct, ~pct_ft,
  "Cut",          75, 29.0, 93, 106, 1.413, 70, "Very Good", 68, 47, 21, 69.1, 69.1, 69.8, 76, 0.0, 14.7,
  "P&R Roll Man", 66, 25.5, 91,  88, 1.333, 79, "Very Good", 58, 42, 16, 72.4, 72.4, 71.0, 62, 6.1,  7.6,
  "Putbacks",     51, 19.7, 86,  64, 1.255, 68, "Very Good", 42, 27, 15, 64.3, 64.3, 64.1, 50, 2.0, 19.6,
  "Transition",   19,  7.3,  2,  28, 1.474, 96, "Excellent", 16, 12,  4, 75.0, 75.0, 73.4, 19, 0.0, 21.1
)

# Defensive play types from Synergy / SportsRadar.
# These are opponent outcomes against Queta.
pt_def <- tibble::tribble(
  ~playtype, ~poss, ~pct_time, ~pct_time_rank, ~pts, ~ppp, ~ppp_rank, ~ppp_rating,
  ~fg_att, ~fg_made, ~fg_miss, ~fg_pct, ~efg_pct, ~ts_pct, ~tsa, ~to_pct,
  "P&R Ball Handler", 176, 53.5, 94, 155, 0.881, 50, "Average",       134, 48, 86, 35.8, 44.4, 50.5, 153, 12.5,
  "Spot Up",           74, 22.5, 20,  90, 1.216, 19, "Below Average",  70, 33, 37, 47.1, 62.9, 63.5,  71,  4.1,
  "P&R Roll Man",      34, 10.3, 63,  34, 1.000, 74, "Very Good",      28, 14, 14, 50.0, 53.6, 55.5,  31, 11.8,
  "Isolation",         30,  9.1, 54,  21, 0.700, 85, "Excellent",      26,  9, 17, 34.6, 34.6, 37.2,  28,  6.7,
  "Post-Up",           10,  3.0,  7,   4, 0.400, 99, "Excellent",      10,  2,  8, 20.0, 20.0, 20.0,  10,  0.0
)

# Databallr WOWY, low-leverage possessions removed.
wowy <- tibble::tribble(
  ~split, ~min, ~off_rtg, ~def_rtg, ~net_rtg, ~off_ts, ~off_tov, ~off_orb, ~def_ts, ~def_tov, ~def_orb,
  "On",   700, 121.3, 111.8,  9.4, 58.5, 12.4, 31.1, 56.3, 16.2, 29.5,
  "Off",  673, 121.1, 119.0,  2.2, 59.5, 12.8, 27.5, 58.3, 14.1, 27.6
) %>%
  bind_rows(
    tibble::tibble(
      split = "On - Off",
      min = NA_real_,
      off_rtg = 0.1,
      def_rtg = -7.1,
      net_rtg = 7.2,
      off_ts = -1.0,
      off_tov = -0.4,
      off_orb = 3.6,
      def_ts = -2.0,
      def_tov = 2.1,
      def_orb = 1.9
    )
  )

# NBA.com defense dashboard.
nba_def <- tibble::tibble(
  player = "Neemias Queta",
  team = "BOS",
  age = 26,
  pos = "C",
  gp = 29,
  games_guarded = 29,
  freq_pct = 100,
  dfgm = 5.8,
  dfga = 12.2,
  dfg_pct = 47.5,
  fg_pct = 49.9,
  diff_pct = -2.4
)

# Rebounding environment by contest level.
reb_contest <- tibble::tribble(
  ~range, ~gp, ~reb_freq, ~oreb, ~dreb, ~reb,
  ~cont_oreb, ~cont_dreb, ~cont_reb, ~cont_reb_pct,
  ~uncont_oreb, ~uncont_dreb, ~uncont_reb, ~uncont_reb_pct,
  "0 Contesting Rebounders",  30, 43.3, 0.9, 2.4, 3.3, 0.0, 0.0, 0.0,   0, 0.9, 2.4, 3.3, 100,
  "1 Contesting Rebounder",   30, 49.4, 1.7, 2.1, 3.8, 1.7, 2.1, 3.8, 100, 0.0, 0.0, 0.0,   0,
  "2+ Contesting Rebounders", 30,  7.4, 0.4, 0.2, 0.6, 0.4, 0.2, 0.6, 100, 0.0, 0.0, 0.0,   0
)

# NBA.com advanced profile.
nba_adv <- tibble::tibble(
  season = "2025-26",
  team = "BOS",
  gp = 31,
  min = 24.7,
  offrtg = 118.5,
  defrtg = 108.7,
  netrtg = 9.7,
  ast_pct = 8.2,
  ast_to = 1.70,
  ast_ratio = 14.9,
  oreb_pct = 11.4,
  dreb_pct = 18.5,
  reb_pct = 14.9
)

# NBA.com traditional profile.
nba_trad <- tibble::tibble(
  season = "2025-26",
  team = "BOS",
  gp = 31,
  min = 24.7,
  pts = 10.1,
  fgm = 4.4,
  fga = 6.6,
  fg_pct = 66.7,
  fg3m = 0.0,
  fg3a = 0.1,
  fg3_pct = 0.0,
  ftm = 1.4,
  fta = 2.3,
  ft_pct = 60.0,
  oreb = 2.9,
  dreb = 4.7,
  reb = 7.6,
  ast = 1.5,
  tov = 0.9,
  stl = 0.8,
  blk = 1.3,
  pf = 2.9
)

# CraftedNBA defensive impact metrics.
crafted_def <- tibble::tribble(
  ~metric,                         ~value, ~percentile,
  "Deflections",                    2.2,     41,
  "raDTOV (Rim-Adjusted D-TOV)",     1.0,     97,
  "Defensive Rebounds (DRB)",       -1.0,     31,
  "Rim Defense (pts saved)",        -6.3,     74,
  "Rim Frequency Defended",         44.4,     86,
  "Block Rate (BLK%)",              5.3,     94,
  "Defensive Versatility",          31.0,     35,
  "Relative Personal Fouls (rPF)",  1.9,      4,
  "Crafted DPM",                    1.6,     87
)
```

# Net Impact Across Minutes

```{r net-impact-vs-minutes, fig.cap="Net impact vs minutes played"}
ggplot(queta_games, aes(x = minutes, y = net_rtg)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_point(color = celtics_green, alpha = 0.85, size = 2.6) +
  geom_smooth(method = "loess", se = FALSE, color = celtics_gold, linewidth = 1) +
  labs(
    title = "Queta’s Minutes Align With Positive Impact",
    subtitle = "Game-level net rating vs minutes played",
    x = "Minutes played",
    y = "Net rating",
    caption = "Source: Game-level on/off data | Viz: @Rambzee_"
  ) +
  theme_caleb_celtics()
```

When Queta’s minutes increase, Boston’s net rating does not deteriorate. Results are often positive across his starter-minute range, and while individual games vary by matchup and context, there is no indication that extended minutes introduce defensive slippage or lineup stress. That stability reinforces Queta’s growth and impact as a starter.

# The Numbers

```{r table-role}
role_tbl <- nba_trad %>%
  transmute(
    Season = season,
    GP     = gp,
    MPG    = min,
    PTS    = pts,
    REB    = reb,
    OREB   = oreb,
    DREB   = dreb,
    AST    = ast,
    BLK    = blk,
    TOV    = tov,
    `FG%`  = fg_pct,
    `FT%`  = ft_pct
  )

knitr::kable(
  role_tbl,
  format = "latex",
  booktabs = TRUE,
  caption = "Traditional production (NBA.com)"
) %>%
  style_latex_table(font_size = 10)
```

```{r table-advanced-profile}
adv_tbl <- nba_adv %>%
  transmute(
    Season   = season,
    OffRtg   = offrtg,
    DefRtg   = defrtg,
    NetRtg   = netrtg,
    `OREB%`  = oreb_pct,
    `DREB%`  = dreb_pct,
    `REB%`   = reb_pct,
    `AST%`   = ast_pct,
    `AST/TO` = ast_to
  )

knitr::kable(
  adv_tbl,
  format = "latex",
  booktabs = TRUE,
  caption = "Advanced profile (NBA.com)"
) %>%
  style_latex_table(font_size = 10)
```

# Offensive Role: Play Finishing

Queta’s offensive role is built around finishing plays rather than creating them. His highest-volume scoring possessions come from:

- Cuts through timing and rim runs
- Roll man possessions that create vertical spacing
- Putbacks that turn glass pressure into second-chance points

```{r table-offensive-playtypes}
pt_off_tbl <- pt_off %>%
  mutate(
    pct_time = pct_time / 100,
    fg_pct   = fg_pct / 100,
    efg_pct  = efg_pct / 100,
    ts_pct   = ts_pct / 100,
    to_pct   = to_pct / 100,
    pct_ft   = pct_ft / 100
  ) %>%
  transmute(
    Play           = playtype,
    Poss           = poss,
    `Time%`        = pct_time,
    PPP            = round(ppp, 3),
    Rating         = ppp_rating,
    `TS%`          = ts_pct,
    `TO%`          = to_pct,
    `FT% of plays` = pct_ft
  ) %>%
  arrange(desc(`Time%`))

knitr::kable(
  pt_off_tbl,
  format = "latex",
  booktabs = TRUE,
  caption = "Offensive play types (Synergy/SportsRadar)"
) %>%
  style_latex_table(font_size = 9, scale_down = TRUE) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::column_spec(4, bold = TRUE)
```

## Efficiency by Play Type

```{r offensive-ppp-plot, fig.cap="PPP by offensive play type"}
weighted_off_ppp <- weighted.mean(pt_off$ppp, pt_off$poss, na.rm = TRUE)

ggplot(pt_off, aes(x = ppp, y = reorder(playtype, ppp))) +
  geom_point(
    aes(size = poss),
    color = celtics_green,
    alpha = 0.9
  ) +
  geom_vline(
    xintercept = weighted_off_ppp,
    linetype = "dashed",
    color = celtics_gold,
    linewidth = 0.9
  ) +
  scale_size(range = c(2, 8)) +
  labs(
    title = "Efficiency by Play Type",
    subtitle = glue(
      "Size: Volume. Dashed line: weighted average PPP ({round(weighted_off_ppp, 3)})."
    ),
    x = "PPP",
    y = NULL,
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb_celtics()
```

## Interpretation

Queta’s offensive value shows up in how defenses are forced to react to him. As a vertical spacer, he consistently draws weakside help and collapses coverage in actions Boston already prioritizes. His cuts and rolls convert ball pressure and spacing into efficient points without changing the structure of the offense.

That pressure carries over to the glass. Queta’s offensive rebounding follows naturally from the attention he commands at the rim. His second-chance scoring, supported by 2.9 offensive rebounds per game and 64.3% shooting on putbacks, raises the cost of helping off him and extends possessions without requiring touches or play calls.

# Defensive Impact

```{r table-wowy}
knitr::kable(
  wowy,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  caption = "Databallr WOWY (low leverage removed): ON vs OFF"
) %>%
  style_latex_table(font_size = 9, scale_down = TRUE) %>%
  kableExtra::column_spec(1, bold = TRUE)
```

Queta’s defensive impact is substantial and supported by multiple lenses. With him on the floor, Boston allows 7.1 fewer points per 100 possessions, and opponent true shooting drops by approximately two percentage points. The resulting +7.2 net rating reflects impact driven primarily by defensive outcomes rather than offensive usage.

```{r table-craftednba-defense}
crafted_def_tbl <- crafted_def %>%
  transmute(
    Metric = metric,
    Value = value,
    Percentile = percentile
  )

knitr::kable(
  crafted_def_tbl,
  format = "latex",
  booktabs = TRUE,
  caption = "Defensive impact metrics (CraftedNBA)"
) %>%
  style_latex_table(font_size = 9, scale_down = TRUE) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::column_spec(3, bold = TRUE)
```

CraftedNBA metrics help explain how that defensive impact is produced. Queta contests a large share of opponent rim attempts, with a 44.4% rim frequency defended mark that sits in the 86th percentile. When he is the closest defender at the rim, opponents shoot 6.3 percentage points worse than their baseline, a 74th-percentile outcome. He pairs that deterrence with strong event creation, ranking in the 97th percentile in rim-adjusted defensive turnovers forced and the 94th percentile in block rate.

Queta’s defensive profile is concentrated rather than broad. He is most effective around the basket, offers limited value guarding in space, and does not consistently disrupt passing lanes. His foul rate places a ceiling on minutes, and his defensive rebounding is solid but not a defining strength. Within a rim-focused role, however, the results are clear: Queta reliably limits shot quality at the basket and generates defensive events in the areas Boston prioritizes most.

## Rim Defense

```{r rim-defense-data}
# NBA Stats API calls can fail if the endpoint is unavailable or if network access is blocked.
# The filters and calculations below match the original rim defense visual logic.
def_dash <- hoopR::nba_leaguedashptdefend(
  season = season_used,
  season_type = "Regular Season"
)$LeagueDashPTDefend %>%
  janitor::clean_names()

rim <- def_dash %>%
  mutate(
    workload  = as.numeric(d_fga),
    dfg_pct   = as.numeric(d_fg_pct),
    base_pct  = as.numeric(normal_fg_pct),
    fg_change = dfg_pct - base_pct
  ) %>%
  filter(!is.na(workload), workload >= 75, !is.na(fg_change)) %>%
  filter(player_position %in% c("C", "PF", "F-C", "C-F", "PF-C"))

players <- hoopR::nba_leaguedashplayerstats(
  season = season_used,
  season_type = "Regular Season",
  measure_type = "Base"
)$LeagueDashPlayerStats %>%
  janitor::clean_names() %>%
  select(player_id, player_name)

rim <- rim %>%
  left_join(players, by = "player_name") %>%
  mutate(
    headshot = if_else(
      !is.na(player_id),
      paste0(
        "https://cdn.nba.com/headshots/nba/latest/1040x760/",
        player_id,
        ".png"
      ),
      NA_character_
    )
  )

rim_target <- rim %>%
  filter(player_name == target_name)

rim_anchors <- rim %>%
  slice_max(workload, n = 150) %>%
  bind_rows(rim_target) %>%
  distinct(player_name, .keep_all = TRUE)

rim_other <- rim %>%
  filter(!player_name %in% rim_anchors$player_name)
```

```{r rim-defense-plot-build}
p_rim <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_point(
    data = rim_other,
    aes(x = workload, y = fg_change),
    color = "#6b6b6b",
    alpha = 0.15,
    size = 1.2
  ) +
  ggimage::geom_image(
    data = rim_anchors,
    aes(x = workload, y = fg_change, image = headshot),
    size = 0.075,
    asp = 1.1
  ) +
  ggimage::geom_image(
    data = rim_target,
    aes(x = workload, y = fg_change, image = headshot),
    size = 0.09,
    asp = 1.1
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Rim Defense in League Context",
    subtitle = "Defended FG% relative to opponent norm. Volume on X-axis.",
    x = "Shots defended",
    y = "Change in opponent FG%",
    caption = "Source: NBA Stats (via hoopR) | Viz: @Rambzee_"
  )
```

```{r rim-defense-save-and-include, fig.cap="Rim Defense in League Context", out.width="\\linewidth"}
rim_output_file <- "fig_rim_def_context.png"

p_rim <- p_rim +
  theme(
    plot.background  = element_rect(fill = bg_color, color = NA),
    panel.background = element_rect(fill = bg_color, color = NA),
    panel.grid.major = element_line(color = grid_color, linewidth = 0.35),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(size = 26, face = "bold", hjust = 0.5, color = text_dark),
    plot.subtitle    = element_text(size = 16, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 6)),
    axis.title       = element_text(size = 22, face = "bold", color = text_dark),
    axis.text        = element_text(size = 18, color = "#333333"),
    legend.position  = "none",
    plot.caption     = element_text(size = 14, color = "#6e6e6e", hjust = 0, margin = margin(t = 6)),
    plot.margin      = margin(10, 12, 14, 12)
  )

ggsave(
  filename = rim_output_file,
  plot = p_rim,
  width = 7.2,
  height = 4.6,
  units = "in",
  dpi = 320,
  bg = bg_color,
  device = ragg::agg_png
)

knitr::include_graphics(rim_output_file)
```

Across all shot types, Queta defends a high volume of shots while holding opponents roughly three percentage points below their normal efficiency.

## Interpretation

Opponent results back up the film and the tracking data. When Queta is the primary defender, shot efficiency drops below baseline, consistent with real rim protection rather than scheme-driven results. Those outcomes align with the on/off data, where Boston defends at a considerably higher level during his minutes.

On/off numbers should always be analyzed in context, but the direction matters. With Queta on the floor, the Celtics are not simply holding ground defensively; they are gaining it. His minutes consistently align with defensive improvement and positive net results, reinforcing his impact.

## Rebounding Environment

```{r rebounding-environment-plot, fig.cap="Rebounding environment by contest level"}
reb_long <- reb_contest %>%
  mutate(reb_freq = reb_freq / 100) %>%
  select(range, reb_freq)

ggplot(reb_long, aes(x = reorder(range, reb_freq), y = reb_freq)) +
  geom_col(fill = celtics_green, alpha = 0.95) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Rebounding Environment",
    subtitle = "Share of rebounds by number of contesting rebounders",
    x = NULL,
    y = "Frequency",
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb_celtics()
```

Queta’s rebounding profile reflects his role within Boston’s defensive structure. Most of his rebounds come with zero or one contesting rebounder, with limited 2+ contested volume. That distribution is typical for a center operating behind strong point-of-attack defense and emphasizes positioning and timing rather than pure box-out dominance.

The takeaway is not that Queta is an elite defensive rebounder. It is that his rebounding fits the Celtics ecosystem. He secures possessions efficiently without compromising rim protection or defensive coverage.

# Closing Summary

Neemias Queta has taken a real step forward this season, marked by improved execution and consistency on both ends. Offensively, he has become a reliable play finisher, turning Boston’s creation into points through cuts, rolls, putbacks, and strong on- and off-ball screening. Defensively, his rim protection has stabilized, and Boston’s lineups have consistently performed better with him on the floor.

In a season defined by frontcourt uncertainty, Queta has provided the Celtics with minutes they can trust. He has handled starter-level responsibility, sustained physicality, and tougher matchups without forcing schematic compromises or disrupting offensive flow.

If this version of Queta holds, center minutes are no longer something Boston needs to manage around. They have a dependable big who brings athleticism, energy, and physicality that align directly with the Celtics’ identity.
