---
title: "Anthony Davis in Washington: A Buy-Low Bet on Top-75 Talent"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
params:
  playtypes_xlsx: "AD Playtypes (Last 5 Seasons).xlsx"
  zones_xlsx: "AD Scoring Zones (Last 5 Seasons).xlsx"
  years_back: 5
  season_keep: ["2021-22","2022-23","2023-24","2024-25","2025-26"]
  mix_playtypes: ["P&R Roll Man","Post-Up","Putbacks","Cut","Isolation","Spot Up","Transition"]
  focus_playtypes: ["P&R Roll Man","Cut","Post-Up","Isolation"]
  fig_w: 7.2
  fig_h: 4.6
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{ragged2e}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  fig.width = params$fig_w,
  fig.height = params$fig_h,
  dpi = 300,
  out.width = "100%"
)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)
library(readr)
library(forcats)
library(janitor)
library(readxl)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(showtext)
library(sysfonts)
library(lubridate)
library(glue)
library(hoopR)
```

```{r constants-and-parameters}
# Theme constants
bg_color   <- "#f6f4ef"
text_dark  <- "#141414"
grid_color <- "#dedad3"

# Wizards palette
wiz_navy <- "#002B5C"
wiz_red  <- "#E31837"
wiz_silv <- "#C4CED4"

# Report parameters
season_keep <- params$season_keep
mix_playtypes <- params$mix_playtypes
focus_playtypes <- params$focus_playtypes

font_add_google("Inter", "inter")
showtext_auto()
```

```{r helper-functions}
theme_caleb <- function() {
  theme_minimal(base_family = "inter") +
    theme(
      plot.background  = element_rect(fill = bg_color, color = NA),
      panel.background = element_rect(fill = bg_color, color = NA),
      panel.grid.major = element_line(color = grid_color, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 18, face = "bold", hjust = 0.5, color = text_dark),
      plot.subtitle    = element_text(size = 10, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 8)),
      axis.title       = element_text(size = 12, face = "bold", color = text_dark),
      axis.text        = element_text(size = 10, color = "#333333"),
      plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 0, margin = margin(t = 8)),
      plot.margin      = margin(10, 12, 14, 12),
      legend.position  = "top",
      legend.text      = element_text(size = 9)
    )
}

to_num <- function(x) {
  readr::parse_number(as.character(x))
}

extract_start_year <- function(season_chr) {
  start_year <- str_extract(as.character(season_chr), "(19|20)[0-9]{2}")
  suppressWarnings(as.numeric(start_year))
}

infer_col <- function(df, candidates) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

infer_season_col <- function(df) {
  infer_col(df, c("season", "year", "season_label", "season_yr", "seasons"))
}

weighted_mode <- function(x, w) {
  x <- as.character(x)
  keep <- !is.na(x) & !is.na(w)
  x <- x[keep]
  w <- w[keep]
  if (length(x) == 0) return(NA_character_)
  weights_by_value <- tapply(w, x, sum)
  names(weights_by_value)[which.max(weights_by_value)]
}

kbl_caleb <- function(df, caption = NULL, font_size = 9, scale_down = TRUE) {
  table_obj <- kableExtra::kbl(df, format = "latex", booktabs = TRUE, caption = caption)
  latex_options <- c("striped", "hold_position")
  if (scale_down) latex_options <- c(latex_options, "scale_down")
  table_obj %>% kableExtra::kable_styling(latex_options = latex_options, font_size = font_size)
}
```

```{r league-reference-tables}
league_ts <- tribble(
  ~season, ~lg_ts,
  "2020-21", 0.572,
  "2021-22", 0.566,
  "2022-23", 0.581,
  "2023-24", 0.580,
  "2024-25", 0.576,
  "2025-26", 0.580
)

league_zone_fg <- tribble(
  ~zone, ~lg_fg,
  "Restricted Area", 0.664,
  "In The Paint (Non-RA)", 0.444,
  "Midrange", 0.413,
  "Threes", 0.359
)
```

# Executive Summary

Anthony Davis is a top-75 player of all time and, when healthy, has still played at a borderline top-10 level over the past half-decade. Availability remains the concern: he has appeared in just 279 of 456 possible games (61.2%) over the last six seasons, including 71 of 134 (53%) over the last two years.

Injuries, some unlucky and some cumulative, have pushed Davis to the lowest market value of his career, compounded by unrealistic expectations that asked him to carry a young roster and replace a franchise’s former all-time great.

In Washington, Davis finally slots into his preferred role at the four alongside Alex Sarr, forming a high-upside defensive frontcourt behind a young core and a Trae Young-led backcourt. He still provides elite rim pressure, defensive versatility, and play-finishing value when used to amplify strengths rather than force self-creation.

For the Wizards, Davis’s value lies in accelerating their competitive timeline. He adds reliable rim pressure, creates defensive havoc through elite disruption, and brings structure to a roster that will naturally be volatile. If it works, Washington gains an elite two-way presence. If it does not, the downside can be washed away relatively quickly.

```{r data-loading}
# Excel paths are parameterized in the YAML.
# Keep the two files in the project root or update params$playtypes_xlsx and params$zones_xlsx.
pt_raw <- read_excel(params$playtypes_xlsx) %>%
  clean_names()

zn_raw <- read_excel(params$zones_xlsx) %>%
  clean_names()
```

```{r season-filtering}
pt_season_col <- infer_season_col(pt_raw)
zn_season_col <- infer_season_col(zn_raw)

if (is.na(pt_season_col)) {
  stop("Playtypes: could not infer season column. Add or rename a 'season' or 'year' column.")
}

if (is.na(zn_season_col)) {
  stop("Zones: could not infer season column. Add or rename a 'season' or 'year' column.")
}

pt <- pt_raw %>% mutate(season = .data[[pt_season_col]])
zn <- zn_raw %>% mutate(season = .data[[zn_season_col]])

if (!is.null(season_keep)) {
  keep <- season_keep
} else {
  keep <- pt %>%
    distinct(season) %>%
    mutate(start_year = extract_start_year(season)) %>%
    arrange(desc(start_year)) %>%
    slice_head(n = params$years_back) %>%
    pull(season)
}

keep_levels <- tibble(season = keep) %>%
  mutate(start_year = extract_start_year(season)) %>%
  arrange(start_year) %>%
  pull(season)

pt <- pt %>%
  filter(season %in% keep_levels) %>%
  mutate(season = factor(season, levels = keep_levels))

zn <- zn %>%
  filter(season %in% keep_levels) %>%
  mutate(season = factor(season, levels = keep_levels))
```

```{r playtype-cleaning}
col_playtype <- infer_col(pt, c("playtype", "play_type", "type"))
col_poss <- infer_col(pt, c("possessions", "poss"))
col_ppp <- infer_col(pt, c("ppp", "avg_ppp", "points_per_possession"))
col_time <- infer_col(pt, c("percent_time", "pct_time", "time", "share"))

if (any(is.na(c(col_playtype, col_poss, col_ppp)))) {
  stop("Missing required playtype columns. Update mapping in this chunk.")
}

pt_clean <- pt %>%
  transmute(
    season = season,
    play_type = .data[[col_playtype]],
    poss = to_num(.data[[col_poss]]),
    pct_time = if (!is.na(col_time)) to_num(.data[[col_time]]) else NA_real_,
    ppp = to_num(.data[[col_ppp]])
  ) %>%
  filter(!is.na(play_type), poss > 0, !is.na(ppp)) %>%
  mutate(
    play_type = case_when(
      str_to_lower(play_type) %in% c("iso", "isolation") ~ "Isolation",
      TRUE ~ as.character(play_type)
    )
  )

pt_meta <- pt %>%
  transmute(
    season = season,
    play_type = case_when(
      str_to_lower(.data[[col_playtype]]) %in% c("iso", "isolation") ~ "Isolation",
      TRUE ~ as.character(.data[[col_playtype]])
    ),
    poss = to_num(.data[[col_poss]]),
    ppp_rank = if ("ppp_rank" %in% names(pt)) to_num(ppp_rank) else NA_real_,
    ppp_rating = if ("ppp_rating" %in% names(pt)) as.character(ppp_rating) else NA_character_
  )
```

```{r zone-cleaning}
col_zone <- infer_col(zn, c("zone", "shot_zone", "area", "range"))
col_fgm <- infer_col(zn, c("fgm", "makes", "made"))
col_fga <- infer_col(zn, c("fga", "attempts", "shots"))

if (any(is.na(c(col_zone, col_fgm, col_fga)))) {
  stop("Zones: need zone + fgm + fga columns. Update mapping in this chunk.")
}

zone_order <- c("Restricted Area", "In The Paint (Non-RA)", "Midrange", "Threes")

zn_clean <- zn %>%
  transmute(
    season = season,
    zone = as.character(.data[[col_zone]]),
    fgm = to_num(.data[[col_fgm]]),
    fga = to_num(.data[[col_fga]])
  ) %>%
  filter(!is.na(zone), !is.na(season), fga > 0) %>%
  mutate(
    zone = factor(zone, levels = zone_order),
    fg_pct = fgm / fga
  )
```

# Overview

```{r ad-season-profile-table, results='asis'}
ad_seasons <- tribble(
  ~season, ~gp, ~mpg,
  ~fgm, ~fga, ~fg_pct,
  ~tpm, ~tpa, ~tp_pct,
  ~ftm, ~fta, ~ft_pct,
  ~reb, ~ast, ~stl, ~blk, ~pf, ~tov, ~pts,
  "2020-21", 36, 32.3, 8.4, 17.0, 0.491, 0.7, 2.8, 0.260, 4.4, 5.9, 0.738, 7.9, 3.1, 1.3, 1.6, 1.7, 2.1, 21.8,
  "2021-22", 40, 35.1, 9.3, 17.4, 0.532, 0.3, 1.8, 0.186, 4.4, 6.1, 0.713, 9.9, 3.1, 1.2, 2.3, 2.4, 2.1, 23.2,
  "2022-23", 56, 34.0, 9.7, 17.2, 0.563, 0.3, 1.3, 0.257, 6.2, 7.9, 0.784, 12.5, 2.6, 1.1, 2.0, 2.6, 2.2, 25.9,
  "2023-24", 76, 35.5, 9.4, 16.9, 0.556, 0.4, 1.4, 0.271, 5.5, 6.8, 0.816, 12.6, 3.5, 1.2, 2.3, 2.3, 2.1, 24.7,
  "2024-25", 51, 33.4, 9.2, 17.8, 0.516, 0.7, 2.4, 0.282, 5.6, 7.2, 0.775, 11.6, 3.5, 1.2, 2.2, 1.9, 2.2, 24.7,
  "2025-26", 20, 31.3, 8.5, 16.7, 0.506, 0.5, 1.9, 0.270, 3.0, 4.1, 0.728, 11.1, 2.8, 1.1, 1.7, 2.1, 2.1, 20.4
) %>%
  mutate(ts = pts / (2 * (fga + 0.44 * fta))) %>%
  left_join(league_ts, by = "season") %>%
  mutate(rts = ts - lg_ts)

ad_table <- ad_seasons %>%
  mutate(
    `FGM-FGA` = sprintf("%.1f-%.1f", fgm, fga),
    `3PM-3PA` = sprintf("%.1f-%.1f", tpm, tpa),
    `FTM-FTA` = sprintf("%.1f-%.1f", ftm, fta)
  ) %>%
  transmute(
    Season = season,
    GP = gp,
    MPG = round(mpg, 1),
    `FGM-FGA` = `FGM-FGA`,
    `FG%` = percent(fg_pct, accuracy = 0.1),
    `3PM-3PA` = `3PM-3PA`,
    `3P%` = percent(tp_pct, accuracy = 0.1),
    `FTM-FTA` = `FTM-FTA`,
    `FT%` = percent(ft_pct, accuracy = 0.1),
    REB = round(reb, 1),
    AST = round(ast, 1),
    STL = round(stl, 1),
    BLK = round(blk, 1),
    PF = round(pf, 1),
    TOV = round(tov, 1),
    PTS = round(pts, 1),
    `TS%` = percent(ts, accuracy = 0.1),
    `League TS%` = percent(lg_ts, accuracy = 0.1),
    `rTS%` = percent(rts, accuracy = 0.1)
  )

kbl_caleb(
  ad_table,
  caption = "Anthony Davis: Season splits (2020-21 to 2025–26) - TS% and rTS% vs league",
  font_size = 8,
  scale_down = TRUE
) %>%
  column_spec(1, bold = TRUE)
```

```{=latex}
\par\justifying
```

Over the past five seasons, Davis has hovered around 24 points per game on great efficiency with strong rebounding and elite defensive playmaking, but the year-to-year story shows how much his production can still be shaped by minutes and health. That volatility shows up in swings in both scoring and efficiency relative to league-average TS%. Davis is most effective when he is attacking the rim and drawing fouls, and that pattern shows up across his best recent seasons.

# Offense: Playtypes

## Playtype Table

```{r playtype-table}
pt_table <- pt_clean %>%
  left_join(pt_meta, by = c("season", "play_type", "poss")) %>%
  filter(play_type %in% unique(c(mix_playtypes, focus_playtypes))) %>%
  group_by(play_type) %>%
  summarise(
    Seasons = n_distinct(season),
    Poss = sum(poss, na.rm = TRUE),
    `Avg %Time` = if (all(is.na(pct_time))) NA_real_ else weighted.mean(pct_time, w = poss, na.rm = TRUE),
    `Avg PPP` = weighted.mean(ppp, w = poss, na.rm = TRUE),
    `Avg PPP Rank` = if (all(is.na(ppp_rank))) NA_real_ else weighted.mean(ppp_rank, w = poss, na.rm = TRUE),
    `Avg PPP Rating` = if (all(is.na(ppp_rating))) NA_character_ else weighted_mode(ppp_rating, poss),
    .groups = "drop"
  ) %>%
  arrange(desc(Poss)) %>%
  mutate(
    `Avg %Time` = ifelse(is.na(`Avg %Time`), NA, percent(`Avg %Time`, accuracy = 0.1)),
    `Avg PPP` = round(`Avg PPP`, 3),
    `Avg PPP Rank` = ifelse(is.na(`Avg PPP Rank`), NA, round(100 * `Avg PPP Rank`, 0))
  )

kbl_caleb(
  pt_table,
  caption = "Anthony Davis: Featured playtypes (possession-weighted summary)",
  font_size = 9
) %>%
  column_spec(1, bold = TRUE)
```

Davis’ offense is at its best in traditional big-man roles. As a roller, cutter, offensive rebounder, and post-up threat, he still produces strong volume and efficiency, averaging over 1.0 PPP across those actions. His offensive rebounding, in particular, remains a major advantage.

His limitations show up more as a perimeter scorer. Davis ranks in just the 17th percentile on spot-ups, struggles to consistently punish closeouts, and is good but not great in isolation outside the post. That profile fits Washington well because it does not require him to be a primary creator. His value comes from being a pressure point through screens, seals, short rolls, and mismatches.

## Usage Mix

```{r playtype-usage-plot}
playtype_colors_wiz <- c(
  "P&R Roll Man" = wiz_navy,
  "Post-Up" = wiz_red,
  "Putbacks" = wiz_silv,
  "Cut" = "#7BDFF2",
  "Isolation" = "#8E7CC3",
  "Spot Up" = "#2A9D8F",
  "Transition" = "#264653"
)

pt_share <- pt_clean %>%
  filter(play_type %in% mix_playtypes) %>%
  group_by(season) %>%
  mutate(pct_use = poss / sum(poss, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(play_type = factor(play_type, levels = mix_playtypes))

ggplot(pt_share, aes(season, pct_use, fill = play_type)) +
  geom_col(width = 0.78, alpha = 0.95) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = playtype_colors_wiz) +
  labs(
    title = "AD: Playtype Usage Mix",
    subtitle = "Seven-playtype profile across 2021-22 to 2025-26 seasons",
    x = NULL,
    y = "% of possessions",
    fill = NULL,
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb()
```

## PPP Trends

```{r playtype-ppp-plot}
pt_plot4 <- pt_clean %>%
  filter(play_type %in% focus_playtypes) %>%
  mutate(play_type = factor(play_type, levels = focus_playtypes))

pt_baseline <- pt_plot4 %>%
  group_by(play_type) %>%
  summarise(ppp_base = weighted.mean(ppp, w = poss, na.rm = TRUE), .groups = "drop")

ggplot(pt_plot4, aes(season, ppp, group = 1)) +
  geom_hline(data = pt_baseline, aes(yintercept = ppp_base), linetype = "dashed", linewidth = 0.6, color = "#3a3a3a", alpha = 0.55, inherit.aes = FALSE) +
  geom_line(linewidth = 1.2, color = wiz_navy) +
  geom_point(aes(size = poss), color = wiz_red, alpha = 0.95) +
  facet_wrap(~ play_type, ncol = 2) +
  scale_size(range = c(1.5, 4), guide = "none") +
  coord_cartesian(ylim = c(0.6, 1.48)) +
  labs(
    title = "Anthony Davis: Playtype Efficiency Trends",
    subtitle = "Dot size = possessions | Dashed line = AD's average PPP",
    x = NULL,
    y = "PPP",
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb() +
  theme(panel.grid.major = element_line(color = "#e3dfd8", linewidth = 0.3), axis.text = element_text(size = 8, color = "#333333"))
```

Davis is most consistent as a cutter and roll man, where his efficiency remains strong year to year. His post-up scoring has softened somewhat, and isolation dipped this season in limited time. When his offense is driven by movement and rim pressure it is at its best. When it shifts toward self-creation, it becomes less efficient, though he remains a good scorer overall.

# Shot Zones: AD vs League FG%

```{r shot-zone-plot}
zn_sum <- zn_clean %>%
  group_by(zone) %>%
  summarise(fgm = sum(fgm, na.rm = TRUE), fga = sum(fga, na.rm = TRUE), fg_pct = fgm / fga, .groups = "drop") %>%
  left_join(league_zone_fg %>% mutate(zone = factor(zone, levels = zone_order)), by = "zone") %>%
  mutate(lbl = paste0(percent(fg_pct, accuracy = 0.1), " (", comma(fga), ")"))

zone_colors <- c(
  "Restricted Area" = wiz_navy,
  "In The Paint (Non-RA)" = "#1D4E89",
  "Midrange" = wiz_silv,
  "Threes" = wiz_red
)

ggplot(zn_sum, aes(zone, fg_pct, fill = zone)) +
  geom_col(width = 0.72, alpha = 0.95) +
  geom_segment(aes(x = as.numeric(zone) - 0.36, xend = as.numeric(zone) + 0.36, y = lg_fg, yend = lg_fg), linewidth = 0.6, linetype = "dotted", color = "#636363", alpha = 0.9) +
  geom_text(aes(y = fg_pct * 0.5, label = lbl), size = 2.8, family = "inter", color = "white", fontface = "bold") +
  scale_fill_manual(values = zone_colors, guide = "none") +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.8), expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "Anthony Davis: Shot Zone FG% vs League",
    subtitle = "2021-22 to 2025-26 | League FG% shown as dotted reference",
    x = NULL,
    y = "FG%",
    caption = "Data: NBA.com & Basketball-Reference.com | Viz: @Rambzee_"
  ) +
  theme_caleb()
```

Davis is still a top-tier finisher at the rim, and he remains effective in the paint as a counter when he cannot get all the way to the basket. As his shot diet drifts to the midrange or beyond the arc, the efficiency drop becomes clear. He can take those shots when needed, but his offense is at its best when he is living near the rim and forcing defenses to respond to his size and touch rather than settling for jumpers.

# Defensive Profile

```{r defensive-api-functions}
options(timeout = 120)

seasons_def <- c("2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
ad_player_id <- 203076
min_gp <- 20
min_mpg <- 15

safe_ldps <- function(season) {
  tryCatch(
    hoopR::nba_leaguedashplayerstats(season = season, season_type = "Regular Season", per_mode = "PerGame"),
    error = function(e) {
      message("LDPS failed for ", season, ": ", e$message)
      NULL
    }
  )
}

safe_shotdef <- function(season, player_id) {
  tryCatch(
    hoopR::nba_playerdashptshotdefend(season = season, season_type = "Regular Season", player_id = player_id),
    error = function(e) {
      message("ShotDef failed for ", season, ": ", e$message)
      NULL
    }
  )
}

extract_first_df <- function(obj, preferred_name = NULL) {
  if (is.null(obj)) return(NULL)
  if (!is.null(preferred_name) && preferred_name %in% names(obj) && is.data.frame(obj[[preferred_name]])) return(obj[[preferred_name]])
  tables <- obj[vapply(obj, is.data.frame, logical(1))]
  if (length(tables) == 0) return(NULL)
  tables[[1]]
}

stocks_season_tbl <- function(season) {
  raw <- extract_first_df(safe_ldps(season), preferred_name = "LeagueDashPlayerStats")
  if (is.null(raw) || nrow(raw) == 0) return(NULL)

  df <- raw %>% clean_names()
  nm_player <- intersect(c("player_name", "player"), names(df))[1]
  nm_team <- intersect(c("team_abbreviation", "team"), names(df))[1]

  df %>%
    transmute(
      season = season,
      player_id = as.integer(player_id),
      player = .data[[nm_player]],
      team = if (!is.na(nm_team)) .data[[nm_team]] else NA_character_,
      gp = as.numeric(gp),
      mpg = as.numeric(min),
      stl = as.numeric(stl),
      blk = as.numeric(blk),
      stocks = stl + blk
    ) %>%
    filter(is.finite(gp), is.finite(mpg), gp >= min_gp, mpg >= min_mpg) %>%
    mutate(stocks_rank = rank(-stocks, ties.method = "min"), n_players = n())
}

clean_shotdef_tbl <- function(obj, season_label) {
  shot_tbl <- extract_first_df(obj, preferred_name = "DefendingShots")
  if (is.null(shot_tbl) || nrow(shot_tbl) == 0) return(NULL)

  df <- shot_tbl %>% clean_names()
  nm_cat <- intersect(c("defense_category", "shot_distance_range", "distance_range"), names(df))[1]
  nm_dfg <- intersect(c("d_fg_pct", "defended_fg_pct", "fg_pct"), names(df))[1]
  nm_exp <- intersect(c("normal_fg_pct", "expected_fg_pct", "fg_pct_expected"), names(df))[1]
  nm_att <- intersect(c("d_fga", "fga", "defended_fga"), names(df))[1]

  if (any(is.na(c(nm_cat, nm_dfg, nm_exp, nm_att)))) return(NULL)

  df %>%
    transmute(
      season = season_label,
      defense_category = as.character(.data[[nm_cat]]),
      d_fga = as.numeric(.data[[nm_att]]),
      d_fg_pct = as.numeric(.data[[nm_dfg]]),
      exp_fg_pct = as.numeric(.data[[nm_exp]])
    )
}
```

```{r stocks-ranks-last5, results='asis'}
stocks_all <- map_dfr(seasons_def, stocks_season_tbl)

if (nrow(stocks_all) == 0) {
  kbl_caleb(
    tibble(Note = "No league player stats returned. Most likely: API call failed or hoopR not loaded."),
    caption = "Stocks per game ranks (STL+BLK) — last 5 seasons",
    font_size = 9,
    scale_down = FALSE
  )
} else {
  ad_stocks <- stocks_all %>%
    filter(player_id == ad_player_id) %>%
    transmute(
      Season = season,
      Team = team,
      GP = gp,
      MPG = round(mpg, 1),
      STL = round(stl, 1),
      BLK = round(blk, 1),
      `Stocks/G` = round(stocks, 2),
      Rank = paste0(stocks_rank, " / ", n_players),
      Pctile = percent(1 - (stocks_rank - 1) / n_players, accuracy = 0.1)
    )

  kbl_caleb(
    ad_stocks,
    caption = glue("Anthony Davis: Stocks per game ranks (STL+BLK) — min {min_gp} GP, min {min_mpg} MPG"),
    font_size = 9,
    scale_down = FALSE
  ) %>%
    column_spec(1, bold = TRUE)
}
```

Stocks per game is a simple but effective way to capture defensive event creation, and it highlights one of Davis’s most dependable strengths. When healthy, he has consistently ranked near the top of the league in combined steals and blocks, using elite hands and defensive awareness to create extra possessions.

For a young Washington team, the margins matter even more. Defensive events fuel transition opportunities, cover for mistakes, and reduce the need for perfect half-court execution. Pairing Davis with Alex Sarr, who brings similarly strong stocks production (3.7 per game), gives the Wizards a frontcourt capable of turning disruption into easy points.

```{=latex}
\par\justifying
```

```{r stocks-leaderboard}
stocks_leaderboard <- tribble(
  ~Rank, ~Player, ~`Stocks/G`, ~`BLK/G`, ~`STL/G`, ~GP, ~MPG, ~PPG, ~RPG, ~APG, ~`TS%`,
  1, "Victor Wembanyama", 4.57, 3.4, 1.1, 155, 30.5, 22.9, 10.9, 3.5, 0.587,
  2, "Anthony Davis", 3.31, 2.2, 1.2, 243, 34.3, 24.4, 11.8, 3.2, 0.603,
  3, "Jaren Jackson Jr.", 3.10, 2.0, 1.1, 327, 29.5, 19.7, 5.9, 1.6, 0.572,
  4, "Walker Kessler", 2.85, 2.4, 0.5, 201, 25.3, 9.5, 9.3, 1.2, 0.679,
  5, "Chet Holmgren", 2.83, 2.2, 0.6, 160, 29.0, 16.6, 8.2, 2.1, 0.632
) %>%
  mutate(`TS%` = percent(`TS%`, accuracy = 0.1))

kbl_caleb(
  stocks_leaderboard,
  caption = "NBA stocks leaders (STL + BLK per game), last 5 seasons",
  font_size = 8,
  scale_down = TRUE
) %>%
  row_spec(2, bold = TRUE)
```

Since 2021, Davis ranks second in the NBA in stocks per game, highlighting how rare his defensive impact still is even as he moves into the later stages of his career. Built on elite instincts, timing, and still-high-level athleticism, his value shows up not just in rim protection but in consistently driving strong team defense.

```{r rim-protection-league-rank, results='asis'}
ad_rim <- map_dfr(seasons_def, function(season) clean_shotdef_tbl(safe_shotdef(season, ad_player_id), season)) %>%
  filter(
    is.finite(d_fga), d_fga > 0,
    str_detect(defense_category, "6") & str_detect(str_to_lower(defense_category), "less")
  ) %>%
  mutate(diff = d_fg_pct - exp_fg_pct)

if (nrow(ad_rim) == 0) {
  kbl_caleb(
    tibble(Note = "No <6 ft shot defense rows returned. Most likely: category label mismatch or API returned different schema."),
    caption = "Rim protection (AD): DFG% vs Expected - last 5 seasons",
    font_size = 9,
    scale_down = FALSE
  )
} else {
  ad_rim_out <- ad_rim %>%
    transmute(
      Season = season,
      `Defended Att` = as.integer(round(d_fga)),
      `DFG%` = percent(d_fg_pct, accuracy = 0.1),
      `Exp%` = percent(exp_fg_pct, accuracy = 0.1),
      Diff = percent(diff, accuracy = 0.1)
    )

  kbl_caleb(
    ad_rim_out,
    caption = "Anthony Davis: Rim protection at the rim (<6 ft) - DFG% vs Expected",
    font_size = 9,
    scale_down = FALSE
  ) %>%
    column_spec(1, bold = TRUE)
}
```

```{=latex}
\par\justifying
```

The rim data highlights Davis’s value as an anchor. Opponents consistently shoot worse than expected at the rim when he is the primary defender. Pairing him with Alex Sarr only amplifies that impact. Sarr brings his own rim protection, giving Washington real flexibility. Davis can handle bigger centers while Sarr roams as a backline defender, or the two can switch against quicker bigs. Together, they give the Wizards a frontcourt that can erase mistakes and control the paint.

# Fit in Washington

```{r washington-core-table}
wiz_core <- tribble(
  ~Player, ~Age, ~Pos, ~MPG, ~PTS, ~TRB, ~AST, ~`3P%`, ~`3PA`, ~STL, ~BLK, ~TOV,
  "Trae Young", 27, "PG", 34.3, 25.2, 3.5, 9.8, 0.351, 7.5, 1.0, 0.2, 4.2,
  "Kyshawn George", 22, "SF", 30.6, 15.1, 5.5, 4.9, 0.376, 5.6, 1.1, 0.9, 2.8,
  "Alex Sarr", 20, "C", 28.3, 17.3, 7.7, 2.8, 0.339, 3.0, 0.8, 2.1, 1.7,
  "Bub Carrington", 20, "PG", 28.2, 10.1, 3.6, 4.6, 0.402, 5.3, 0.6, 0.2, 2.4,
  "Bilal Coulibaly", 21, "SG", 26.6, 9.7, 4.6, 2.6, 0.262, 3.3, 1.4, 0.9, 1.5,
  "Tre Johnson", 19, "SG", 25.1, 12.9, 2.7, 2.2, 0.393, 5.5, 0.6, 0.3, 1.4
)

kbl_caleb(
  wiz_core %>% mutate(`3P%` = percent(`3P%`, accuracy = 0.1)),
  caption = "Washington young core (as provided) — context for AD fit",
  font_size = 8,
  scale_down = TRUE
)
```

```{=latex}
\par\justifying
```

## Defense: Raising the Floor and Aiding the Young Guys

With a young core and a shaky defensive backcourt, you are going to have blown coverages and late help. AD reduces the cost of those mistakes by turning rim attempts into tougher finishes and by creating extra possessions through blocks and steals. That gives Washington margin while their defensive habits develop.

**With Sarr**

This is the most interesting pairing. Sarr’s tools, including length, mobility, and rim protection, allow him to play as a roaming eraser, while Davis can take the more physical matchups and serve as the primary anchor. That dynamic is flexible. Against bruising centers, AD handles the strength while Sarr operates as an elite weak-side deterrent. Against quicker bigs and five-out lineups, the two can switch more actions and still recover at the rim. Either way, it is a frontcourt built to cover ground and mask perimeter breakdowns, especially when Trae is targeted defensively.

**With Kyshawn + Bilal**

These two are promising defensive wings. If Kyshawn continues to grow as a bigger, switchable wing and Bilal maintains his disruptive point-of-attack defense, the AD-Sarr pairing gives them a strong backline to play more aggressively. That support allows the wings to press, gamble selectively, and use their length to create advantages. Whether Washington reaches a top-15 defense will likely hinge on how well those wings, and the rest of the perimeter group, can consistently hold up.

**With Trae + Bub**

Trae will always be a defensive stress point. Even with effort, the small-guard tax is hard to escape, and Bub faces similar issues as he continues learning physicality and positioning. This is where AD’s size and mobility matter most, reducing how often guards are left on an island. When they are the primary defender, the task becomes funneling drives, buying time, and fighting hard enough to turn rim attempts into pull-ups instead of easy finishes.

## Offense: Simplify, Add Stability, and Relieve Pressure

Washington does not need AD to carry the offense or create everything himself. They need him to be a steady interior force that bends the defense. With four of their six core players shooting 35% or better from three and Sarr close behind at 34%, there is enough shooting to punish help and keep the floor functional. There is still room for growth, but the surrounding spacing should be sufficient to maximize Davis’s impact in the paint.

**Trae + AD**

Trae’s gravity paired with AD’s screening and rolling should quickly become the backbone of Washington’s offense. Even without elite spacing, that two-man game forces rotations, creates rim pressure, and opens kick-out reads for young wings. AD also serves as a dependable release valve through lobs, short rolls, seals, and post touches so possessions do not stall when defenses switch or navigate screens well.

**Sarr’s development**

If Sarr develops into a credible floor-spacer, it unlocks AD at the four in his preferred role and keeps the lane cleaner for rim pressure. Early on, the two can be staggered to maintain spacing and rim presence. Long term, Sarr’s shooting is what makes the pairing seamless. Even if the window is only a few seasons, playing alongside an elite interior scorer like AD could accelerate Sarr’s offensive development.

**Kyshawn + Tre**

Both benefit from the attention Davis commands. Kyshawn should see more catch-and-shoot looks and cleaner driving lanes, letting him lean into his growth as a secondary creator against tilted defenses. Tre gains easier shot opportunities as a movement shooter and a reliable screen partner if his on-ball reps increase in Trae-out minutes. Beyond spacing, AD’s screening is also impactful. Across staggers, flares, zipper cuts, and other off-ball actions, Davis’s screening strength forces defenders to navigate real contact rather than slip cleanly over the top. That physical toll creates consistent separation, turning marginal windows into clean catches and freeing shooters for a handful of extra quality looks each game.

## Overall: A Buy-Low Bet

Anthony Davis is an expensive name, but the actual cost was relatively modest. Washington gave up players and picks they had minimal long-term regard for and, in return, added a borderline top-10 talent without tying their future to him. If the fit works, they gain elite two-way production. If it does not, they can move on without damaging the core, either by letting him walk or absorbing a short late-career overpay.

The move fits the Trae Young approach. Buy low on elite talent, reduce expectations, and see what a new environment can unlock alongside a young roster. This may not be the future Davis once imagined, but it gives him a clear role as a stabilizing veteran and tone-setter, similar to the bridge LeBron once provided for him.
