---
title: "Rudy Gobert: Defensive Impact & DPOY Case (2025–26)"
author: "Caleb Ramsey"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: false
params:
  season_label: "2025–26"
  player_name: "Rudy Gobert"
  team_label: "Minnesota Timberwolves"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)
library(janitor)
library(readxl)
library(stringr)
library(forcats)

# Timberwolves palette (simple)

MIN_NAVY  <- "#0C2340"
MIN_LAKE  <- "#236192"
MIN_GREEN <- "#78BE20"

theme_caleb <- function() {
theme_minimal() +
theme(
plot.background = element_rect(fill = "#f6f4ef", color = NA),
panel.background = element_rect(fill = "#f6f4ef", color = NA),
panel.grid.major = element_line(color = "#dedad3", linewidth = 0.35),
panel.grid.minor = element_blank(),
plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
plot.subtitle = element_text(size = 10, hjust = 0.5, margin = margin(b = 8)),
axis.title = element_text(size = 12, face = "bold"),
axis.text  = element_text(size = 10),
plot.caption = element_text(size = 8, hjust = 0),
legend.position = "top"
)
}

knitr::opts_chunk$set(
echo = FALSE, message = FALSE, warning = FALSE,
fig.pos = "H", fig.align = "center",
fig.width = 7.2, fig.height = 4.6,
dpi = 300, out.width = "100%"
)

to_num <- function(x) {
x <- as.character(x)
x <- str_replace_all(x, ",", "")
x <- str_replace_all(x, "%", "")
suppressWarnings(as.numeric(str_replace_all(x, "[^0-9\\.-]", "")))
}

as_rate01 <- function(x)
{
x <- suppressWarnings(as.numeric(x))
ifelse(!is.na(x) & x > 1, x / 100, x)
}

kbl_pdf <- function(df) {
  df2 <- df
  names(df2) <- gsub("_", " ", names(df2))

  knitr::kable(
    df2,
    format   = "latex",
    booktabs = TRUE,
    longtable = FALSE,
    escape   = TRUE
  )
}

tbl_float <- function(tabular_tex, caption) {
  knitr::asis_output(
    paste0(
      "\\begin{table}[H]\n\\centering\n",
      "\\caption{", caption, "}\n",
      tabular_tex, "\n",
      "\\end{table}\n"
    )
  )
}

```

# Executive Summary

Rudy Gobert is still the backbone of Minnesota’s defense. When he’s on the floor the rim shrinks; when he sits the defense collapses. The Wolves still orbit him.

Offensively his role hasn’t changed, but he’s finishing at a career-best level inside. A four-time DPOY, he continues to carry a massive defensive workload and tilt games the way he has for nearly a decade.

# Offense
```{r off-data}
# update path if needed

fin_path <- "C:/Users/caleb/Downloads/RudyGobert_RimAndPaintFinishing_Career.xlsx"

fin_raw <- read_excel(fin_path) %>%
clean_names()

# normalize year colnames: "2014.0" -> "2014"

fix_year_cols <- function(nm) {
nm %>%
str_replace("^x", "") %>%
str_replace("\\.0$", "") %>%
str_replace("_0$", "")
}
names(fin_raw) <- fix_year_cols(names(fin_raw))

metric_col <- names(fin_raw)[1]  # first column is the metric label

fin_long <- fin_raw %>%
pivot_longer(
cols = -all_of(metric_col),
names_to = "season",
values_to = "value"
) %>%
mutate(
metric = str_to_lower(as.character(.data[[metric_col]])),
season = suppressWarnings(as.integer(season)),
value  = to_num(value)
)

# IMPORTANT: do NOT match "ra" (it hits "non-ra")

fin_long2 <- fin_long %>%
mutate(
zone = case_when(
str_detect(metric, "restricted") ~ "Restricted Area",
str_detect(metric, "paint")      ~ "Paint (Non-RA)",
TRUE                             ~ NA_character_
),
measure = case_when(
str_detect(metric, "fga$") ~ "FGA",
str_detect(metric, "fgm$") ~ "FGM",
str_detect(metric, "fg%|fg_percent|fg$") ~ "FG%",
TRUE ~ NA_character_
),
fg_pct = if_else(measure == "FG%", as_rate01(value), NA_real_)
) %>%
filter(!is.na(zone), zone %in% c("Restricted Area", "Paint (Non-RA)")) %>%
filter(measure %in% c("FGA", "FGM", "FG%"))
```

## Career Finishing Snapshot
```{r career-finishing, results='asis'}
fin_fg <- fin_long2 %>%
filter(measure == "FG%") %>%
group_by(season, zone) %>%
summarise(fg_pct = mean(fg_pct, na.rm = TRUE), .groups = "drop") %>%
arrange(season)

fin_tbl <- fin_long2 %>%
mutate(out = if_else(measure == "FG%", fg_pct, value)) %>%
group_by(season, zone, measure) %>%
summarise(
out = case_when(
measure %in% c("FGA", "FGM") ~ sum(out, na.rm = TRUE),
measure == "FG%"            ~ mean(out, na.rm = TRUE),
TRUE                        ~ NA_real_
),
.groups = "drop"
) %>%
pivot_wider(
names_from  = c(zone, measure),
values_from = out,
values_fn   = list(out = ~ mean(.x, na.rm = TRUE))
) %>%
mutate(season_num = suppressWarnings(as.integer(season))) %>%
arrange(season_num) %>%
select(-season_num) %>%
rename(Year = season) %>%
mutate(
`Restricted Area_FG%` = percent(`Restricted Area_FG%`, accuracy = 1),
`Paint (Non-RA)_FG%`  = percent(`Paint (Non-RA)_FG%`, accuracy = 1)
)

fin_tbl_disp <- fin_tbl %>%
  rename(
    `RA FG%`        = `Restricted Area_FG%`,
    `RA FGA`        = `Restricted Area_FGA`,
    `RA FGM`        = `Restricted Area_FGM`,
    `Paint FG%`     = `Paint (Non-RA)_FG%`,
    `Paint FGA`     = `Paint (Non-RA)_FGA`,
    `Paint FGM`     = `Paint (Non-RA)_FGM`
  )

tbl_float(as.character(kbl_pdf(fin_tbl_disp)),
          "Career finishing by zone (Restricted Area vs Non-RA Paint).")
```

## Career Finishing Trend
```{r career-finishing-trend, fig.cap = "Rudy Gobert's Finishing Across His Career"}
fin_fg2 <- fin_fg %>%
  mutate(
    is_latest = season == max(season, na.rm = TRUE)
  )

ggplot(fin_fg2, aes(x = season, y = fg_pct, color = zone, group = zone)) +
  geom_line(linewidth = 1.15) +
  geom_point(aes(size = is_latest), show.legend = FALSE) +
  scale_size_manual(values = c("FALSE" = 2.2, "TRUE" = 4.2)) +
  scale_color_manual(
    values = c(
      "Restricted Area" = MIN_LAKE,
      "Paint (Non-RA)"  = MIN_NAVY
    )
  ) +
  scale_x_continuous(breaks = seq(min(fin_fg2$season), max(fin_fg2$season), by = 2)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  facet_wrap(~ zone, scales = "free_y") +
  labs(
    title = "Gobert Finishing Efficiency by Zone",
    subtitle = "Restricted Area vs Non-RA Paint",
    x = NULL, y = "FG%",
    color = NULL,
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

Gobert’s shot diet has always been simple, but Figure 1 shows a steady climb in restricted-area efficiency. Over his career, his finishing at the rim has gradually risen into the mid-70s, while his non–restricted-area paint efficiency (historically shaky) has been significantly better in 2025–26. He's improved a bit with his touch, even at this stage of his career.

## Other Impacts
He is averaging 3.9 offensive rebounds per game, accounting for 34.5% of Minnesota’s total (11.3), a major source of extra possessions. He also creates advantages as a screener, averaging 4.2 screen assists per game, second only to Jusuf Nurkić (4.6). These plays offset some of his limitations on the margins.

# Defense
## Team Impact
```{r on-off-data}
onoff <- tribble(
~split, ~min, ~off_rtg, ~def_rtg, ~net_rtg,
"Gobert ON",  1231, 116.8, 108.1,  8.7,
"Gobert OFF",  637, 122.6, 123.4, -0.8
) %>%
mutate(
`OffRtg` = sprintf("%.1f", off_rtg),
`DefRtg` = sprintf("%.1f", def_rtg),
`Net`    = sprintf("%+.1f", net_rtg)
) %>%
select(Split = split, MIN = min, OffRtg = `OffRtg`, DefRtg = `DefRtg`, Net = `Net`)


tbl_float(as.character(kbl_pdf(onoff)), "Minnesota on/off with Gobert (ratings per 100 possessions).")
```

## DFG% Impact
```{r dfg-impact, results='asis'}
# ---------- 1) Total DFG% diff leaderboard (min 25 GP) ----------

dfg_total <- tribble(
~player, ~team, ~age, ~pos, ~gp, ~g, ~freq_pct, ~dfgm, ~dfga, ~dfg_pct, ~fg_pct, ~diff_pp,
"Victor Wembanyama","SAS",22,"F-C",28,28,100,174,440,39.5,49.5,-10.0,
"Duop Reath","POR",29,"C",26,26,100,48,125,38.4,47.3,-8.9,
"Ariel Hukporti","NYK",23,"C",25,25,100,53,133,39.8,48.5,-8.6,
"Jonathan Isaac","ORL",28,"F",29,29,100,44,113,38.9,47.2,-8.2,
"Ajay Mitchell","OKC",23,"G",41,41,100,125,335,37.3,45.3,-8.0,
"Chet Holmgren","OKC",23,"C-F",37,37,100,216,530,40.8,47.6,-6.9,
"Steven Adams","HOU",32,"C",31,31,100,158,369,42.8,49.4,-6.6,
"Paul Reed","DET",26,"F",29,29,100,76,181,42.0,48.4,-6.4,
"Brook Lopez","LAC",37,"C",34,34,100,131,306,42.8,49.1,-6.3,
"Jaren Jackson Jr.","MEM",26,"F-C",38,38,100,241,578,41.7,47.9,-6.2,
"Rudy Gobert","MIN",33,"C",41,41,100,307,708,43.4,49.5,-6.1,
"Jericho Sims","MIL",27,"C",27,27,100,59,145,40.7,46.7,-6.0,
"Tyrese Proctor","CLE",21,"G",26,26,100,41,103,39.8,45.6,-5.8,
"Julian Phillips","CHI",22,"F",25,25,100,39,99,39.4,45.2,-5.8,
"Evan Mobley","CLE",24,"C",37,37,100,276,639,43.2,48.7,-5.5
) %>%
mutate(
dfg_pct = as_rate01(dfg_pct),
fg_pct  = as_rate01(fg_pct)
) %>%
arrange(diff_pp)

gobert_total_rank <- dfg_total %>%
mutate(rank = row_number()) %>%
filter(player == "Rudy Gobert") %>%
pull(rank)

dfg_total_tbl <- dfg_total %>%
transmute(
Rank = row_number(),
Player = player,
Team = team,
GP = gp,
`DFGM/DFGA` = paste0(dfgm, "/", dfga),
`Opp FG%` = percent(dfg_pct, accuracy = 0.1),
`Expected FG%` = percent(fg_pct, accuracy = 0.1),
`Diff (pp)` = sprintf("%+.1f", diff_pp)
)

cap_total <- paste0(
  "Total defensive FG% impact leaderboard (min 25 GP). Gobert ranks #",
  gobert_total_rank, " of 15 (more negative = better)."
)

tbl_float(as.character(kbl_pdf(dfg_total_tbl)), cap_total)

# ---------- 2) Rim-only DFG% diff leaderboard (min 25 GP, >=100 defended) ----------

dfg_rim <- tribble(
~player, ~team, ~age, ~pos, ~gp, ~g, ~freq_pct, ~dfgm, ~dfga, ~dfg_pct, ~fg_pct, ~diff_pp,
"Isaiah Stewart","DET",24,"F-C",36,36,36.7,72,158,45.6,62.7,-17.2,
"Chet Holmgren","OKC",23,"C-F",37,37,44.2,113,234,48.3,63.3,-15.0,
"Kevin Durant","HOU",37,"F",37,34,35.1,76,154,49.4,63.7,-14.3,
"Trayce Jackson-Davis","GSW",25,"F",29,24,47.6,50,101,49.5,62.7,-13.2,
"Dominick Barlow","PHI",22,"F",30,30,38.3,70,139,50.4,63.1,-12.8,
"Adem Bona","PHI",22,"F",31,29,41.2,58,112,51.8,64.6,-12.8,
"Brook Lopez","LAC",37,"C",34,32,42.5,64,130,49.2,61.9,-12.7,
"Brandon Miller","CHA",23,"F",25,24,43.6,73,142,51.4,63.7,-12.3,
"Jaden McDaniels","MIN",25,"F",41,40,38.1,117,228,51.3,63.5,-12.2,
"Rudy Gobert","MIN",33,"C",41,41,36.4,132,258,51.2,63.3,-12.1,
"Neemias Queta","BOS",26,"C",38,38,42.6,103,197,52.3,63.6,-11.3,
"Nic Claxton","BKN",26,"C",37,37,48.2,129,245,52.7,63.6,-10.9,
"Evan Mobley","CLE",24,"C",37,37,44.9,153,287,53.3,64.0,-10.7,
"Ivica Zubac","LAC",28,"C",35,35,37.1,109,202,54.0,64.6,-10.6,
"Quinten Post","GSW",25,"C",43,40,35.3,82,153,53.6,63.5,-9.9
) %>%
mutate(
freq_pct = as_rate01(freq_pct),
dfg_pct  = as_rate01(dfg_pct),
fg_pct   = as_rate01(fg_pct)
) %>%
arrange(diff_pp)

gobert_rim_rank <- dfg_rim %>%
mutate(rank = row_number()) %>%
filter(player == "Rudy Gobert") %>%
pull(rank)

dfg_rim_tbl <- dfg_rim %>%
transmute(
Rank = row_number(),
Player = player,
Team = team,
GP = gp,
`DFGM/DFGA` = paste0(dfgm, "/", dfga),
`Rim Opp FG%` = percent(dfg_pct, accuracy = 0.1),
`Rim Expected FG%` = percent(fg_pct, accuracy = 0.1),
`Diff (pp)` = sprintf("%+.1f", diff_pp)
)

cap_rim <- paste0(
  "Rim-only defensive FG% impact leaderboard (min 25 GP, ",
  "\\geq 100 defended). Gobert ranks #", gobert_rim_rank, " of 15."
)

tbl_float(as.character(kbl_pdf(dfg_rim_tbl)), cap_rim)
```

```{=latex}
\par\raggedright
```

## Total DFG% Leaderboard
```{r tot-dfg-leaderboard, fig.cap="Defensive FG% impact (all shots). Diff = Opp FG% minus Expected FG% in percentage points. Lower is better."}
plot_total <- dfg_total %>%
  mutate(
    player = fct_rev(fct_reorder(player, diff_pp, .desc = FALSE)),
    is_rudy = player == "Rudy Gobert",
    label_flag = rank(diff_pp) <= 3 | is_rudy   # top 5 + Rudy
  )

ggplot(plot_total, aes(x = player, y = diff_pp)) +
  geom_col(aes(fill = is_rudy), width = 0.68, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_text(
    data = subset(plot_total, label_flag),
    aes(label = sprintf("%+.1f", diff_pp)),
    hjust = ifelse(plot_total$diff_pp[plot_total$label_flag] < 0, 1.05, -0.05),
    size = 2.5
  ) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = MIN_NAVY, "FALSE" = "#e3e0da")) +
  scale_y_continuous(labels = function(x) sprintf("%+.1f", x)) +
  labs(
    title = "Defensive FG% Impact (All Shots)",
    subtitle = "Lower = better rim suppression",
    x = NULL,
    y = "Diff (Percentage Points)",
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

On the all-shots leaderboard Gobert is still an elite defender by raw differential. He is defending the second-highest volume of attempts in the league (708), behind only Donovan Clingan. Night after night he absorbs heavy rim traffic and still holds shooters below expectation, even if his impact is not as flashy or versatile as Wembanyama’s.

## Rim-only DFG% Diff Leaderboard
```{r rim-dfg-leaderboard}
plot_rim <- dfg_rim %>%
  mutate(
    player = fct_rev(fct_reorder(player, diff_pp, .desc = FALSE)),
    is_rudy = player == "Rudy Gobert",
    label_flag = rank(diff_pp) <= 3 | is_rudy   # top 5 + Rudy
  )

ggplot(plot_rim, aes(x = player, y = diff_pp)) +
  geom_col(aes(fill = is_rudy), width = 0.68, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_text(
    data = subset(plot_rim, label_flag),
    aes(label = sprintf("%+.1f", diff_pp)),
    hjust = ifelse(plot_rim$diff_pp[plot_rim$label_flag] < 0, 1.05, -0.05),
    size = 2.5
  ) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = MIN_NAVY, "FALSE" = "#e3e0da")) +
  scale_y_continuous(labels = function(x) sprintf("%+.1f", x)) +
  labs(
    title = "Rim Protection Impact (Diff in FG%)",
    subtitle = "Min. 25 GP",
    x = NULL,
    y = "Diff (Percentage Points)",
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

At the rim, Gobert ranks 10th among high-volume protectors, holding shooters to 51.2% versus an expected 63.3% while carrying one of the league’s heaviest defensive loads. That’s why Minnesota treats him as the backbone of its defense.

## Opp FG% rank at rim
```{r opp-fg-rank, fig.cap="Rim defense: opponent FG% at the rim (<6 ft). Min 25 GP and 5+ DFGA/G."}
rim6_raw <- tribble(
~player, ~team, ~age, ~pos, ~gp, ~g, ~freq_pct, ~dfgm_pg, ~dfga_pg, ~dfg_pct, ~exp_fg_pct, ~diff_pp,
"Chet Holmgren",   "OKC", 23, "C-F", 37, 37, 44.2, 3.0, 6.3, 48.3, 63.4, -15.1,
"Rudy Gobert",     "MIN", 33, "C",   41, 41, 36.4, 3.2, 6.3, 51.2, 63.3, -12.2,
"Jaden McDaniels", "MIN", 25, "F",   41, 40, 38.1, 2.9, 5.6, 51.3, 63.6, -12.3,
"Brandon Miller",  "CHA", 23, "F",   25, 24, 43.6, 2.9, 5.7, 51.4, 63.7, -12.3,
"Derrick White",   "BOS", 31, "G",   40, 40, 37.0, 2.8, 5.5, 51.4, 61.1,  -9.7,
"Neemias Queta",   "BOS", 26, "C",   38, 38, 42.6, 2.7, 5.2, 52.3, 63.6, -11.3,
"Nic Claxton",     "BKN", 26, "C",   37, 37, 48.2, 3.5, 6.6, 52.7, 63.6, -10.9,
"Evan Mobley",     "CLE", 24, "C",   37, 37, 44.9, 4.1, 7.8, 53.3, 64.1, -10.8,
"Alex Sarr",       "WAS", 20, "C",   32, 32, 53.5, 5.3, 9.8, 54.0, 63.7,  -9.8,
"Ivica Zubac",     "LAC", 28, "C",   35, 35, 37.1, 3.1, 5.8, 54.0, 64.6, -10.7,
"Jaren Jackson Jr.","MEM",26, "F-C", 38, 36, 36.7, 3.0, 5.6, 54.2, 62.9,  -8.6
)

rim6_plot <- rim6_raw %>%
mutate(
dfg = dfg_pct / 100,
is_rudy = player == "Rudy Gobert",
# order by Opp FG% ascending, then reverse so lowest is at top after flip
player = fct_rev(fct_reorder(player, dfg, .desc = FALSE))
)

rim6_plot <- rim6_plot %>%
  mutate(
    label_flag = rank(dfg) <= 2 | is_rudy   # top 5 + Rudy
  )

ggplot(rim6_plot, aes(x = player, y = dfg)) +
  geom_col(aes(fill = is_rudy), width = 0.68, show.legend = FALSE) +
  geom_text(
    data = subset(rim6_plot, label_flag),
    aes(label = percent(dfg, accuracy = 0.1)),
    hjust = 1.05,
    size = 2.5
  ) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = MIN_LAKE, "FALSE" = "#e3e0da")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Rim Defense: Opponent FG%",
    subtitle = "Min 25 GP and 5+ DFGA/G.",
    x = NULL,
    y = "Opp FG%",
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb() + 
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

Among players defending at least five rim shots per game, Gobert ranks second, holding shooters to 51.2%, behind only Chet Holmgren at 48.3%. Minnesota funnels traffic to him by design, and he still wins more often than not.

# DPOY Case & Conclusion

Gobert’s DPOY case is strong even with voter fatigue working against him. Minnesota is a bottom-tier defense the moment he sits at 123.4 DRTG, worse than last-place Utah, and a top-10 unit when he plays. That is roughly a +15.3 defensive on/off swing, the kind you normally only see from offensive superstars like Jokić or Shai.

By comparison, Chet Holmgren’s defensive on/off is around -0.3. He is obviously incredible, but Gobert is putting up offensive-superstar impact numbers on the defensive end.