---
title: "Jayson Tatum’s Return Opened the Celtics’ Next Chapter"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  dpi = 200
)
```

```{r libraries}
library(tidyverse)
library(scales)
library(knitr)
library(glue)
library(showtext)
library(sysfonts)
library(hoopR)
library(janitor)
library(lubridate)
library(ggrepel)
library(ggimage)
library(gt)
library(purrr)
library(kableExtra)
library(zoo)
```

```{r constants-and-parameters}
# Theme constants
bg_color    <- "#f6f4ef"
text_dark   <- "#141414"
grid_color  <- "#dedad3"

# Team colors
celtics_green <- "#007A33"
celtics_gold  <- "#BA9653"
celtics_dark  <- "#1f2a44"
sixers_blue   <- "#006BB6"
sixers_red    <- "#ED174C"

# NBA identifiers
team_id  <- "1610612738"
tatum_id <- "1628369"

# Report parameters
season_used <- "2025-26"
return_date <- as.Date("2026-03-06")
logo_dir <- "assets/NBA-Team-Logos"

# Captions
caption_line      <- "Data: NBA.com | Viz: @Rambzee_"
caption_databallr <- "Data: Databallr | Viz: @Rambzee_"
caption_manual    <- "Data: Basketball Reference / Databallr / NBA.com | Viz: @Rambzee_"

font_add_google("Inter", "inter")
showtext_auto()
```

```{r helper-functions}
safe_clean_names <- function(df) {
  janitor::clean_names(df)
}

to_num <- function(x) {
  suppressWarnings(as.numeric(x))
}

fmt_num <- function(x, digits = 1) {
  formatC(x, format = "f", digits = digits)
}

fmt_pct <- function(x, digits = 1) {
  paste0(formatC(x, format = "f", digits = digits), "\\%")
}

fmt_signed <- function(x, digits = 1) {
  paste0(if_else(x > 0, "+", ""), formatC(x, format = "f", digits = digits))
}

parse_nba_date <- function(x) {
  suppressWarnings(
    lubridate::parse_date_time(
      x,
      orders = c("Ymd", "ymd", "mdY", "b d, Y", "B d, Y")
    )
  ) %>%
    as.Date()
}

calc_ts <- function(pts, fga, fta) {
  pts / (2 * (fga + 0.44 * fta))
}

safe_divide <- function(num, den) {
  if_else(den == 0 | is.na(den), NA_real_, num / den)
}

make_pct_name <- function(x) {
  stringr::str_replace_all(x, "%", "\\\\%")
}

style_kable_tatum <- function(kbl_obj, font_size = 10.5) {
  kbl_obj %>%
    kableExtra::kable_styling(
      latex_options = c("HOLD_position"),
      full_width = FALSE,
      position = "center",
      font_size = font_size
    ) %>%
    kableExtra::row_spec(0, bold = TRUE) %>%
    kableExtra::column_spec(1, bold = TRUE)
}

theme_caleb <- theme_minimal(base_family = "inter") +
  theme(
    plot.background  = element_rect(fill = bg_color, color = NA),
    panel.background = element_rect(fill = bg_color, color = NA),
    panel.grid.major = element_line(color = grid_color, linewidth = 0.35),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(size = 18, face = "bold", hjust = 0.5, color = text_dark),
    plot.subtitle    = element_text(size = 9.5, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 6)),
    axis.title       = element_text(size = 12, face = "bold", color = text_dark),
    axis.text        = element_text(size = 10, color = "#333333"),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    legend.text      = element_text(size = 9, color = "#333333"),
    plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 1, margin = margin(t = 8)),
    plot.margin      = margin(10, 12, 16, 12)
  )

build_regular_season_kable <- function(raw_df, left_label = "2024-25", right_label = "2025-26") {
  display <- raw_df %>%
    mutate(
      Metric = make_pct_name(Metric),
      left_display = if_else(is_pct, fmt_pct(left), fmt_num(left)),
      right_display = if_else(is_pct, fmt_pct(right), fmt_num(right)),
      change_num = right - left,
      Change = case_when(
        Metric == "GP" ~ "--",
        is_pct ~ paste0(fmt_signed(change_num), "%"),
        TRUE ~ fmt_signed(change_num)
      )
    ) %>%
    transmute(
      Metric,
      !!left_label := left_display,
      !!right_label := right_display,
      Change
    )

  change_colors <- case_when(
    raw_df$is_pct & raw_df$right - raw_df$left <= -2 ~ "#b91c1c",
    raw_df$right - raw_df$left >= 2.5 ~ "green",
    raw_df$right - raw_df$left > 0 ~ "olive",
    raw_df$right - raw_df$left <= -2 ~ "red",
    TRUE ~ "black"
  )

  display$Change <- purrr::map2_chr(
    display$Change,
    change_colors,
    ~ kableExtra::cell_spec(.x, color = .y, bold = TRUE)
  )

  display %>%
    kableExtra::kbl(
      format = "latex",
      booktabs = TRUE,
      escape = FALSE,
      align = c("l", "c", "c", "c"),
      caption = "Tatum's regular-season profile before and after the Achilles return"
    ) %>%
    style_kable_tatum(font_size = 10.5)
}
```

```{r api-functions}
# NBA Stats / hoopR helpers. Downstream chunks use tryCatch fallbacks so the
# report can still render if an endpoint is temporarily unavailable.
get_player_gamelog <- function(player_id, season = season_used, season_type = "Regular Season") {
  hoopR::nba_playergamelog(
    season = season,
    season_type = season_type,
    player_id = player_id
  ) %>%
    purrr::pluck("PlayerGameLog") %>%
    safe_clean_names() %>%
    mutate(
      game_id = as.character(game_id),
      game_date = parse_nba_date(as.character(game_date)),
      player_id = as.character(player_id),
      min = to_num(min),
      pts = to_num(pts),
      oreb = to_num(oreb),
      dreb = to_num(dreb),
      reb = to_num(reb),
      ast = to_num(ast),
      stl = to_num(stl),
      blk = to_num(blk),
      tov = to_num(tov),
      pf = to_num(pf),
      fga = to_num(fga),
      fgm = to_num(fgm),
      fg3m = to_num(fg3m),
      fg3a = to_num(fg3a),
      fta = to_num(fta),
      ftm = to_num(ftm),
      plus_minus = to_num(plus_minus),
      ts = calc_ts(pts, fga, fta)
    ) %>%
    arrange(game_date)
}

get_league_base <- function(season = season_used, season_type = "Regular Season") {
  hoopR::nba_leaguedashplayerstats(
    season = season,
    season_type = season_type,
    measure_type = "Base"
  )$LeagueDashPlayerStats %>%
    clean_names() %>%
    mutate(
      player_name = as.character(player_name),
      team_abbreviation = as.character(team_abbreviation),
      player_id = readr::parse_number(as.character(player_id)),
      gp = readr::parse_number(as.character(gp)),
      min = readr::parse_number(as.character(min)),
      pts = readr::parse_number(as.character(pts)),
      reb = readr::parse_number(as.character(reb)),
      ast = readr::parse_number(as.character(ast)),
      plus_minus = readr::parse_number(as.character(plus_minus))
    )
}

get_league_usage <- function(season = season_used, season_type = "Regular Season") {
  hoopR::nba_leaguedashplayerstats(
    season = season,
    season_type = season_type,
    measure_type = "Usage"
  )$LeagueDashPlayerStats %>%
    clean_names() %>%
    transmute(
      player_id = readr::parse_number(as.character(player_id)),
      usg_pct = readr::parse_number(as.character(usg_pct))
    ) %>%
    mutate(
      usg_pct = if_else(usg_pct <= 1, usg_pct * 100, usg_pct)
    )
}

get_league_advanced <- function(season = season_used, season_type = "Regular Season") {
  hoopR::nba_leaguedashplayerstats(
    season = season,
    season_type = season_type,
    measure_type = "Advanced"
  )$LeagueDashPlayerStats %>%
    clean_names() %>%
    transmute(
      player_id = readr::parse_number(as.character(player_id)),
      ts_pct = readr::parse_number(as.character(ts_pct)),
      ast_pct = readr::parse_number(as.character(ast_pct)),
      dreb_pct = readr::parse_number(as.character(dreb_pct)),
      net_rating = readr::parse_number(as.character(net_rating))
    ) %>%
    mutate(
      ts_pct = if_else(ts_pct <= 1, ts_pct * 100, ts_pct),
      ast_pct = if_else(ast_pct <= 1, ast_pct * 100, ast_pct),
      dreb_pct = if_else(dreb_pct <= 1, dreb_pct * 100, dreb_pct)
    )
}
```

```{r hand-input-data}
# Manually entered source extracts used for the final report.
# These preserve the published values for playoff rankings, on/off splits,
# rebounding leaderboards, and Net Points.
regular_season_split <- tibble(
  Metric = c("GP", "PTS", "REB", "AST", "STL", "TOV", "FGA", "FG%", "3P%", "2P%", "TS%"),
  left = c(72, 26.8, 8.7, 6.0, 1.1, 2.9, 20.3, 45.2, 34.3, 55.9, 58.2),
  right = c(16, 21.8, 10.0, 5.3, 1.4, 2.4, 17.9, 41.1, 32.9, 49.3, 54.1),
  is_pct = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)
)

playoff_profile <- tibble(
  Metric = c("MPG", "PTS", "REB", "AST", "STL", "TOV", "FG%", "3P%", "2P%", "eFG%", "TS%"),
  Value = c(36.3, 23.3, 10.7, 6.8, 1.2, 2.8, 47.5, 36.5, 59.2, 56.9, 60.8),
  is_pct = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE),
  Note = if_else(
    Metric %in% c("AST", "FG%", "2P%", "eFG%", "TS%"),
    "Career playoff high",
    ""
  )
)

playmaking_profile <- tibble(
  Indicator = c("Touches per game", "Last season touches per game", "Potential assists", "Rim assists"),
  Value = c("76.3", "84.9", "15.4", "3.0"),
  Context = c("Slightly smaller role-size", "Pre-injury baseline", "91st percentile", "89th percentile")
)

rebound_leaders <- tibble(
  Rank = c(1, 2, 3, 4, 5, 6, 7, 7, 7),
  Player = c(
    "Nikola Jokic", "Victor Wembanyama", "Karl-Anthony Towns",
    "Donovan Clingan", "Rudy Gobert", "Bam Adebayo", "Evan Mobley",
    "Jalen Duren", "Jayson Tatum"
  ),
  Team = c("DEN", "SAS", "NYK", "POR", "MIN", "MIA", "CLE", "DET", "BOS"),
  RPG = c(13.6, 12.3, 11.9, 11.8, 11.4, 10.7, 10.0, 10.0, 10.0)
)

drb_leaders <- tibble(
  Player = c("Wembanyama", "Sabonis", "KAT", "Nurkic", "Tatum", "Jokic"),
  Team = c("SAS", "UTA", "SAC", "NYK", "BOS", "DEN"),
  DRB_Pct = c(28.5, 28.3, 28.3, 27.9, 27.4, 27.0),
  Team_Color = c("#C4CED4", "#5A2D81", "#006BB6", "#002B5C", celtics_green, "#0E2240")
) %>%
  arrange(DRB_Pct) %>%
  mutate(Player = factor(Player, levels = Player))

impact_split <- tibble(
  Group = c("Tatum ON", "Tatum OFF", "ON - OFF"),
  Minutes = c(522, 3424, NA),
  Off_Rating = c(123.2, 120.7, 2.5),
  Def_Rating = c(111.8, 113.4, -1.6),
  Net_Rating = c(11.4, 7.3, 4.1),
  ThreeP_Off = c(35.3, 36.5, NA),
  ThreeP_Def = c(35.5, 35.3, NA)
)

playoff_impact_split <- tibble(
  Group = c("Tatum ON", "Tatum OFF", "ON - OFF"),
  Minutes = c(201, 92, NA),
  Off_Rating = c(118.0, 104.9, 13.1),
  Def_Rating = c(112.0, 111.5, 0.5),
  Net_Rating = c(6.0, -6.6, 12.7),
  ThreeP_Off = c(35.8, 30.9, NA),
  ThreeP_Def = c(35.4, 35.1, NA)
)

playoff_rank_snapshot <- tibble(
  Category = c(
    "Points", "Points", "Points", "Points", "Points",
    "Rebounds", "Rebounds", "Rebounds", "Rebounds", "Rebounds",
    "Assists", "Assists", "Assists", "Assists", "Assists", "Assists", "Assists"
  ),
  Rank = c(
    12, 13, 14, 15, 16,
    1, 2, 3, 4, 5,
    1, 2, 3, 4, 5, 6, 7
  ),
  Player = c(
    "Victor Wembanyama", "Tyrese Maxey", "Jamal Murray", "Jayson Tatum", "LeBron James",
    "Nikola Jokic", "Victor Wembanyama", "Jayson Tatum", "Karl-Anthony Towns", "Alperen Sengun",
    "Nikola Jokic", "Scottie Barnes", "Shai Gilgeous-Alexander", "Cade Cunningham", "LeBron James", "Jrue Holiday", "Jayson Tatum"
  ),
  Value = c(
    24.0, 23.7, 23.7, 23.3, 23.2,
    13.2, 10.7, 10.7, 10.6, 10.2,
    9.5, 8.6, 7.9, 7.5, 7.3, 7.2, 6.8
  )
)

celtics_net_points <- tibble(
  Player = c(
    "Jayson Tatum", "Neemias Queta", "Jaylen Brown", "Ron Harper Jr.",
    "Jordan Walsh", "Baylor Scheierman", 
    "Luka Garza", "Sam Hauser", "Payton Pritchard", "Hugo González", "Nikola Vucevic", "Derrick White"
  ),
  Team = "BOS",
  Offense = c(16, 5, -2, 1, -5, 0, 5, 0, 5, -4, -8, -16),
  Defense = c(4, 10, 5, 2, 6, 1, -4, 0, -6, 0, -1, 2),
  Overall = c(20, 15, 3, 3, 1, 1, 1, 0, -1, -4, -8, -14)
)
```

# Executive Summary

Less than a year after rupturing his Achilles, Jayson Tatum returned to a Celtics team that had already been better than expected without him.

He was not immediately the same scorer he had been prior to the injury, but the rest of his game still looked familiar. He defended across matchups, rebounded at a high level, kept the offense organized, and gave Boston the steady decision-making it had been missing from its best player.

That is the version of Tatum the Celtics built a champion around in 2024. He was a player who could shift with a series, handle different coverages, and impact games even when the jumper was not carrying him.

Before the injury, Tatum was already a four-time First Team All-NBA player and one of the league’s defining stars. The Achilles rupture put that level in question. His return did not erase every concern, but by the playoffs, the outlook looked a lot better. Even while sharing more of the scoring load with Jaylen Brown, Tatum looked closer to the player Boston had trusted for years.

# The Scoring Fluctuated

```{r regular-season-split-table}
build_regular_season_kable(
  regular_season_split,
  left_label = "2024-25",
  right_label = "2025-26"
)
```

The scoring decline was less about Tatum disappearing and more about the details that usually separate his offense. He was still getting shots up and drawing attention, but the burst, balance, and comfort playing through contact were not fully back yet. The clearest sign was inside the arc, where his two-point percentage dropped from 55.9% to 49.3%.

# The Ramp-Up

```{r api-game-log-data}
# These NBA Stats API calls can fail if hoopR is unavailable, rate-limited, or blocked.
tatum_regular <- tryCatch(
  get_player_gamelog(tatum_id, season = season_used, season_type = "Regular Season"),
  error = function(e) tibble()
)

tatum_playoffs <- tryCatch(
  get_player_gamelog(tatum_id, season = season_used, season_type = "Playoffs"),
  error = function(e) tibble()
)

tatum_since_return <- tatum_regular %>%
  filter(game_date >= return_date) %>%
  arrange(game_date) %>%
  mutate(
    game_number = row_number(),
    pts_roll = purrr::map_dbl(game_number, ~ mean(pts[pmax(1, .x - 4):.x], na.rm = TRUE)),
    reb_roll = purrr::map_dbl(game_number, ~ mean(reb[pmax(1, .x - 4):.x], na.rm = TRUE)),
    ast_roll = purrr::map_dbl(game_number, ~ mean(ast[pmax(1, .x - 4):.x], na.rm = TRUE))
  )
```

```{r game-log-trend, fig.width=7.5, fig.height=4.6}
team_logo_paths <- tibble(
  opponent = c(
    "ATL", "BKN", "CHA", "CHI", "CLE", "DAL", "DEN", "DET", "GSW", "HOU",
    "IND", "LAC", "LAL", "MEM", "MIA", "MIL", "MIN", "NOP", "NYK", "OKC",
    "ORL", "PHI", "PHX", "POR", "SAC", "SAS", "TOR", "UTA", "WAS"
  ),
  logo = file.path(logo_dir, paste0(opponent, ".png"))
)

if (nrow(tatum_since_return) > 0) {

  tatum_plot_games <- tatum_since_return %>%
    mutate(
      Game = row_number(),
      opponent = matchup %>%
        stringr::str_extract("([A-Z]{3})$") %>%
        tidyr::replace_na("OPP")
    ) %>%
    left_join(team_logo_paths, by = "opponent")

  n_games <- max(tatum_plot_games$Game, na.rm = TRUE)

  tatum_ppg <- mean(tatum_plot_games$pts, na.rm = TRUE)
  tatum_rpg <- mean(tatum_plot_games$reb, na.rm = TRUE)
  tatum_apg <- mean(tatum_plot_games$ast, na.rm = TRUE)

  y_max <- max(tatum_plot_games$pts, na.rm = TRUE) + 4

  ggplot(tatum_plot_games, aes(x = Game, y = pts)) +
    geom_hline(
      yintercept = tatum_ppg,
      linetype = "dashed",
      linewidth = 0.45,
      color = "#6e6e6e",
      alpha = 0.55
    ) +
    geom_line(
      linewidth = 1.25,
      color = celtics_green,
      lineend = "round"
    ) +
    geom_point(
      size = 3.05,
      color = celtics_green
    ) +
    geom_text(
      aes(label = pts),
      vjust = -1.05,
      size = 2.55,
      family = "inter",
      fontface = "bold",
      color = text_dark
    ) +
    annotate(
      "text",
      x = n_games - 0.25,
      y = tatum_ppg - 1.2,
      label = paste0(round(tatum_ppg, 1), " PPG after return"),
      family = "inter",
      size = 2.45,
      color = "#6e6e6e",
      hjust = 1
    ) +
    ggimage::geom_image(
      aes(x = Game, y = -1.9, image = logo),
      inherit.aes = FALSE,
      size = 0.048,
      by = "width"
    ) +
    scale_x_continuous(
      breaks = tatum_plot_games$Game,
      labels = rep("", n_games)
    ) +
    scale_y_continuous(
      limits = c(-3.5, y_max),
      breaks = seq(0, ceiling(y_max / 10) * 10, 10),
      expand = expansion(mult = c(0.01, 0.10))
    ) +
    labs(
      title = "The Scoring Rhythm Took Time",
      subtitle = paste0(
        "Tatum post-return scoring by game | ",
        round(tatum_rpg, 1), " REB and ",
        round(tatum_apg, 1), " AST per game"
      ),
      x = NULL,
      y = "Points",
      caption = caption_line
    ) +
    theme_caleb +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "none",
      plot.subtitle = element_text(
        size = 9.3,
        hjust = 0.5,
        color = "#2f2f2f",
        margin = margin(b = 6)
      )
    )

} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "Tatum game log unavailable from hoopR.",
      family = "inter",
      size = 4
    ) +
    xlim(-1, 1) +
    ylim(-1, 1) +
    labs(
      title = "Game log unavailable",
      x = NULL,
      y = NULL,
      caption = caption_line
    ) +
    theme_caleb
}
```

Tatum took some time to find his footing as a scorer, but the late-season direction was promising. After some uneven nights early in the return, he settled into the mid-20s by the end of the regular season.

# The Impact Translated Before the Scoring Fully Did

```{r regular-season-impact-table, results='asis'}
impact_display <- impact_split %>%
  transmute(
    Group,
    MIN = if_else(is.na(Minutes), "--", as.character(Minutes)),
    OFF = fmt_num(Off_Rating),
    DEF = fmt_num(Def_Rating),
    NET = paste0(
      if_else(Net_Rating > 0, "+", ""),
      fmt_num(Net_Rating)
    )
  )

impact_display %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "c", "c", "c"),
    caption = "Boston's regular-season lineup performance with and without Jayson Tatum"
  ) %>%
  style_kable_tatum(font_size = 11) %>%
  kableExtra::column_spec(
    5,
    bold = TRUE,
    color = celtics_green
  )
```

Despite the shaky scoring, Tatum's impact came back quickly. In his 522 regular-season minutes, Boston posted a +11.4 net rating, compared with +7.3 when he sat.

That split is not a perfect measure of individual value, especially across 16 games. Lineups, opponents, and shooting swings all have to be taken into consideration. Still, it matched the larger takeaway that Boston had another level with Tatum back on the floor. His rebounding, decision-making, and control of the game led to immediate impact for the Celtics.

# The Rebounding Remained High-Level

```{r post-return-rebounding-table}
rebound_display <- rebound_leaders %>%
  mutate(
    is_tatum = Player == "Jayson Tatum",
    Player = if_else(
      is_tatum,
      cell_spec(Player, bold = TRUE, color = celtics_green),
      Player
    ),
    RPG = if_else(
      is_tatum,
      cell_spec(fmt_num(RPG), bold = TRUE, color = celtics_green),
      fmt_num(RPG)
    )
  ) %>%
  select(-is_tatum)

rebound_display %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    align = c("c", "l", "c", "c"),
    caption = "Most rebounds per game from Tatum's March 6 return through the end of the regular season | Min. 15 GP"
  ) %>%
  style_kable_tatum(font_size = 10.5)
```

```{r drb-percent-leaders, fig.width=7.2, fig.height=4.7}
drb_plot <- drb_leaders %>%
  mutate(
    Player_Label = case_when(
      Player == "Tatum" ~ "Tatum (F)",
      Player == "KAT" ~ "KAT (C)",
      TRUE ~ paste0(Player, " (C)")
    ),
    is_tatum = Player == "Tatum",
    fill_color = if_else(is_tatum, celtics_green, "#cfcac0")
  ) %>%
  arrange(DRB_Pct) %>%
  mutate(Player_Label = factor(Player_Label, levels = Player_Label))

ggplot(drb_plot, aes(x = Player_Label, y = DRB_Pct, fill = fill_color)) +
  geom_col(width = 0.68, alpha = 0.96) +
  geom_text(
    aes(label = paste0(fmt_num(DRB_Pct), "%")),
    hjust = -0.15,
    size = 3.2,
    family = "inter",
    color = text_dark
  ) +
  coord_flip() +
  scale_fill_identity() +
  scale_y_continuous(
    limits = c(0, 31),
    expand = expansion(mult = c(0, 0.06))
  ) +
  labs(
    title = "Tatum Was in Elite Rebounding Company",
    subtitle = "Defensive rebounding percentage leaders | 2025-26 regular season | Min. 15 GP",
    x = NULL,
    y = "DRB%",
    caption = caption_line
  ) +
  theme_caleb +
  theme(legend.position = "none")
```

From his return through the end of the regular season, Tatum averaged 10.0 boards per game and posted a 27.4% defensive rebounding rate, right in the range of several true centers. That gave Boston lineup flexibility it had missed for most of the season, letting the Celtics play smaller and keep more skill on the floor without giving up the glass.

# A Methodical Offensive Engine

```{r playmaking-table}
playmaking_profile %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "l"),
    caption = "Tatum's playmaking indicators during the return season"
  ) %>%
  style_kable_tatum(font_size = 10.5)
```

The most reassuring development may have been how patiently Tatum controlled the offense. His touches dipped slightly from the year before, but the passing indicators were strong: 15.4 potential assists and 3.0 rim assists, both near the top of the league.

Boston kept him in a familiar role, and that part of his game translated right away. Tatum still looked comfortable reading help, controlling the pace, and finding the right pass without rushing. If anything, his playmaking looked even more polished while the scoring efficiency worked its way back.

# Similar Responsibility, Lower Regular-Season Efficiency

```{r league-usage-ts, fig.width=7.5, fig.height=5.1}
league_usage <- tryCatch({
  league_base <- get_league_base()
  league_usage_stats <- get_league_usage()
  league_advanced <- get_league_advanced()

  league_base %>%
    left_join(league_usage_stats, by = "player_id") %>%
    left_join(league_advanced, by = "player_id") %>%
    filter(!is.na(usg_pct), !is.na(ts_pct), !is.na(min), min >= 500)
}, error = function(e) tibble())

if (nrow(league_usage) > 0) {
  avg_usg <- mean(league_usage$usg_pct, na.rm = TRUE)
  avg_ts  <- mean(league_usage$ts_pct, na.rm = TRUE)

  tatum_compare <- tibble(
    season = c("Tatum 2024-25", "Tatum 2025-26"),
    usg_pct = c(31.2, 30.0),
    ts_pct = c(58.2, 54.1)
  )

  ggplot(league_usage, aes(x = usg_pct, y = ts_pct)) +
    geom_vline(xintercept = avg_usg, linetype = "dashed", linewidth = 0.5, alpha = 0.2) +
    geom_hline(yintercept = avg_ts, linetype = "dashed", linewidth = 0.5, alpha = 0.2) +
    geom_point(alpha = 0.16, size = 1.4, color = "#4d4d4d") +
    geom_segment(
      data = tibble(
        x = 31.2,
        xend = 30.0,
        y = 58.2,
        yend = 54.1
      ),
      aes(x = x, xend = xend, y = y, yend = yend),
      inherit.aes = FALSE,
      color = celtics_green,
      linewidth = 0.8,
      alpha = 0.75,
      arrow = arrow(length = unit(0.10, "inches"), type = "closed")
    ) +
    geom_point(
      data = tatum_compare,
      aes(x = usg_pct, y = ts_pct),
      inherit.aes = FALSE,
      color = celtics_green,
      size = 3.4
    ) +
    geom_text(
      data = tatum_compare,
      aes(x = usg_pct, y = ts_pct, label = season),
      inherit.aes = FALSE,
      nudge_y = c(1.3, -1.1),
      size = 2.7,
      family = "inter",
      color = text_dark
    ) +
    scale_x_continuous(labels = scales::label_number(accuracy = 1, suffix = "%")) +
    scale_y_continuous(labels = scales::label_number(accuracy = 1, suffix = "%")) +
    labs(
      title = "Similar Role, Less Scoring Efficiency",
      subtitle = "2025-26 regular season (min. 500 minutes) | Tatum: 2024-25 vs. 2025-26",
      x = "Usage %",
      y = "True Shooting %",
      caption = caption_line
    ) +
    theme_caleb
} else {
  ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = "League data unavailable from hoopR.",
      family = "inter",
      size = 4
    ) +
    xlim(-1, 1) +
    ylim(-1, 1) +
    labs(
      title = "League data unavailable",
      x = NULL,
      y = NULL,
      caption = caption_line
    ) +
    theme_caleb
}
```

# The Scoring Caught Up in the Playoffs

```{r playoff-profile-table}
playoff_profile %>%
  mutate(
    Metric = make_pct_name(Metric),
    Value = if_else(is_pct, fmt_pct(Value), fmt_num(Value))
  ) %>%
  select(Metric, Value, Note) %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    align = c("l", "c", "l"),
    caption = "Tatum's 2026 playoff production"
  ) %>%
  style_kable_tatum(font_size = 10.5) %>%
  kableExtra::column_spec(3, italic = TRUE, color = celtics_green)
```

The playoffs were the clearest sign that the regular-season efficiency dip was not Tatum’s new baseline. He looked more comfortable creating off the dribble and getting downhill, finishing at 47.5% from the field, 59.2% inside the arc, and 60.8% true shooting.

The rebounding and playmaking were there from the start. Once the scoring started to catch up, his place among the playoff leaders reflected it.

# Playoff Rankings

```{r playoff-rankings-table}
playoff_rank_snapshot %>%
  mutate(
    is_tatum = Player == "Jayson Tatum",
    Player = if_else(
      is_tatum,
      cell_spec(Player, bold = TRUE, color = celtics_green),
      Player
    ),
    Value = if_else(
      is_tatum,
      cell_spec(fmt_num(Value), bold = TRUE, color = celtics_green),
      fmt_num(Value)
    )
  ) %>%
  select(-is_tatum) %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    align = c("l", "c", "l", "c"),
    caption = "Where Tatum ranked among 2026 playoff leaders"
  ) %>%
  style_kable_tatum(font_size = 9.5) %>%
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

# Boston Still Ran Through Him

**Net Points Definition:** Net Points is a play-by-play based metric that assigns credit and blame for every possession while a player is on the floor, capturing scoring, turnovers, defense, and context beyond the box score. It reflects how much a player moved the score rather than only what appeared in his individual line.

```{r playoff-net-points-chart, fig.width=7.5, fig.height=5.0}
celtics_net_points %>%
  filter(
    Player %in% c(
      "Jayson Tatum", "Neemias Queta", "Ron Harper Jr.", "Jaylen Brown",
      "Luka Garza", "Baylor Scheierman", "Jordan Walsh", "Sam Hauser",
      "Payton Pritchard", "Hugo González", "Nikola Vucevic", "Derrick White"
    )
  ) %>%
  arrange(Overall) %>%
  mutate(
    Player = factor(Player, levels = Player),
    label = fmt_signed(Overall, 0),
    fill_color = case_when(
      Player == "Jayson Tatum" ~ celtics_green,
      Overall > 0 ~ "#b8c5ad",
      Overall == 0 ~ "#d6d3c8",
      TRUE ~ "#cf7d6d"
    ),
    label_hjust = if_else(Overall >= 0, -0.15, 1.12)
  ) %>%
  ggplot(aes(x = Player, y = Overall, fill = fill_color)) +
  geom_col(width = 0.68, alpha = 0.96) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.45,
    color = "#555555",
    alpha = 0.55
  ) +
  geom_text(
    aes(label = label, hjust = label_hjust),
    size = 3.15,
    family = "inter",
    color = text_dark
  ) +
  coord_flip() +
  scale_fill_identity() +
  scale_y_continuous(
    expand = expansion(mult = c(0.14, 0.16))
  ) +
  labs(
    title = "Tatum Gave Boston Its Biggest Playoff Lift",
    subtitle = "Celtics playoff net points",
    x = NULL,
    y = "Net points added",
    caption = "Data: ESPN Analytics | Viz: @Rambzee_"
  ) +
  theme_caleb +
  theme(legend.position = "none")
```

```{r playoff-impact-table, results='asis'}
playoff_impact_display <- playoff_impact_split %>%
  transmute(
    Group,
    MIN = if_else(is.na(Minutes), "--", as.character(Minutes)),
    OFF = fmt_num(Off_Rating),
    DEF = fmt_num(Def_Rating),
    NET = paste0(
      if_else(Net_Rating > 0, "+", ""),
      fmt_num(Net_Rating)
    ),
    net_color = case_when(
      Net_Rating > 0 ~ celtics_green,
      Net_Rating < 0 ~ "#B23A3A",
      TRUE ~ text_dark
    )
  )

playoff_impact_display %>%
  select(-net_color) %>%
  kableExtra::kbl(
    format = "latex",
    booktabs = TRUE,
    align = c("l", "c", "c", "c", "c"),
    caption = paste0(
      "Boston's playoff performance with and without Jayson Tatum ",
      "| 50\\% offensive and defensive 3-point luck adjustments"
    )
  ) %>%
  style_kable_tatum(font_size = 11) %>%
  kableExtra::column_spec(
    5,
    bold = TRUE,
    color = playoff_impact_display$net_color
  )
```

By the end of the playoffs, Tatum had answered much of the doubt that came with the injury. Boston’s future was still going to run through him.

The Celtics struggled immensely in the minutes without him, especially on offense. He finished with Boston’s best net-points mark at +20, and the adjusted on/off split showed a 13.1-point offensive gap with him on the floor.

Note: Playoff on/off ratings use Databallr’s 50% offensive and defensive three-point luck adjustment, which helps reduce the effect of hot or cold shooting in a small sample. It does not remove the other noise that comes with 201 Tatum minutes and 92 non-Tatum minutes, including lineup context, matchup patterns, and rotation timing. The split should be read as directional, not definitive.

# What This Means Going Forward

After the first major physical setback of his career, Tatum gave Boston real reason to believe. The scoring needed time, but the rebounding, playmaking, defense, and versatility were there almost immediately. By the playoffs, the scoring started to follow. The roster around him may change, the money may get tighter, and Boston may have to build differently, but the postseason brought the same truth back into focus: whatever comes next for the Celtics, it starts with Tatum.
