---
title: "What Are We Watching With Victor Wembanyama?"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    fig_crop: false
geometry: margin=0.78in
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  dpi = 300,
  out.width = "95%"
)

library(tidyverse)
library(scales)
library(kableExtra)
library(showtext)
library(sysfonts)
library(forcats)

try(font_add_google("Inter", "inter"), silent = TRUE)
showtext_auto()
options(scipen = 999)

# ------------------------------------------------------------
# Theme + palette
# ------------------------------------------------------------

BG <- "#f6f4ef"
TEXT_DARK <- "#141414"
TEXT_LIGHT <- "#5f5f5f"
GRID <- "#dedad3"

SPURS_BLACK <- "#141414"
SPURS_SILVER <- "#C4CED4"

CAPTION_LINE <- "Data: Stathead Basketball | Viz: @Rambzee_"

theme_wemby <- function(base_size = 11) {
  theme_minimal(base_family = "inter", base_size = base_size) +
    theme(
      plot.background = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = TEXT_DARK),
      plot.title = element_text(size = 18.5, face = "bold", hjust = 0.5, margin = margin(b = 4)),
      plot.subtitle = element_text(size = 10.2, color = TEXT_LIGHT, hjust = 0.5, margin = margin(b = 8)),
      axis.title = element_text(size = 10.3, face = "bold"),
      axis.text = element_text(size = 9.2, color = "#333333"),
      axis.ticks = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 9.0),
      plot.caption = element_text(size = 7.7, color = "#6e6e6e", hjust = 1, margin = margin(t = 7)),
      plot.margin = margin(10, 12, 12, 12)
    )
}

# ------------------------------------------------------------
# Formatting helpers
# ------------------------------------------------------------

style_kable <- function(
  df,
  caption = NULL,
  align = NULL,
  font_size = 8.8,
  full_width = FALSE,
  escape = TRUE
) {
  if (is.null(align)) align <- strrep("c", ncol(df))

  kbl(
    df,
    booktabs = TRUE,
    format = "latex",
    caption = caption,
    align = align,
    linesep = "",
    escape = escape
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = full_width,
      font_size = font_size,
      stripe_color = "#fbfaf7"
    )
}

fmt_num <- function(x, digits = 1) sprintf(paste0("%.", digits, "f"), x)
fmt_statline <- function(ppg, rpg, apg) paste0(fmt_num(ppg), " / ", fmt_num(rpg), " / ", fmt_num(apg))
fmt_stockline <- function(spg, bpg) {
  if_else(is.na(spg) | is.na(bpg), "—", fmt_num(spg + bpg, 1))
}

highlight_wemby_row <- function(tbl, df, name_col = "Player") {
  wemby_row <- which(df[[name_col]] == "Victor Wembanyama")
  if (length(wemby_row) == 1) {
    tbl %>% row_spec(wemby_row, bold = TRUE, color = "white", background = SPURS_BLACK)
  } else {
    tbl
  }
}
```

```{r data-inputs, include=FALSE}
# ------------------------------------------------------------
# Hand-entered leaderboards from Stathead screenshots
# ------------------------------------------------------------

# ------------------------------------------------------------
# Wembanyama playoff game log
# ------------------------------------------------------------
# Update this one table as the run continues.
# Full-workload averages use games where Wembanyama played at least 20 minutes.
# Two early-exit games remain in the full game log, but are excluded from the main run averages.

wemby_game_log <- tribble(
  ~Date, ~Round_Opp, ~Result, ~MIN, ~PTS, ~REB, ~AST, ~STL, ~BLK, ~TOV, ~GmSc, ~Note,
  "2026-04-19", "POR Game 1", "W", "32:50", 35, 5, 1, 0, 2, 4, 23.9, "Playoff debut",
  "2026-04-21", "POR Game 2", "L", "11:41", 5, 4, 1, 0, 1, 2, 3.9, "Early exit: injury",
  "2026-04-26", "POR Game 4", "W", "34:19", 27, 12, 3, 4, 7, 4, 28.5, "First defensive takeover",
  "2026-04-28", "POR Game 5", "W", "33:36", 17, 14, 3, 0, 6, 1, 22.4, "Closeout control",
  "2026-05-04", "MIN Game 1", "L", "39:52", 11, 15, 5, 0, 12, 3, 13.7, "Most blocks ever in a playoff game",
  "2026-05-06", "MIN Game 2", "W", "25:54", 19, 15, 2, 1, 2, 2, 18.8, "Bounce-back win",
  "2026-05-08", "MIN Game 3", "W", "36:58", 39, 15, 1, 1, 5, 1, 37.9, "First true superstar playoff game",
  "2026-05-10", "MIN Game 4", "L", "12:29", 4, 4, 1, 0, 0, 2, 0.4, "Early exit: ejection",
  "2026-05-12", "MIN Game 5", "W", "32:38", 27, 17, 5, 0, 3, 4, 25.7, "Series control",
  "2026-05-15", "MIN Game 6", "W", "27:22", 19, 6, 2, 0, 3, 4, 13.4, "Series closeout",
  "2026-05-18", "OKC Game 1", "W", "48:42", 41, 24, 3, 1, 3, 3, 40.1, "Conference Finals eruption",
  "2026-05-20", "OKC Game 2", "L", "37:17", 21, 17, 6, 1, 4, 4, 22.9, "Road loss",
  "2026-05-22", "OKC Game 3", "L", "39:09", 26, 4, 3, 1, 2, 2, 20.4, "Home loss",
  "2026-05-24", "OKC Game 4", "W", "31:02", 33, 8, 5, 2, 3, 4, 28.0, "Response game to tie series 2-2",
  "2026-05-26", "OKC Game 5", "L", "37:38", 20, 6, 1, 2, 3, 2, 15.3, "Game 5 road loss",
  "2026-05-28", "OKC Game 6", "W", "28:00", 28, 10, 2, 2, 3, 3, 24.4, "Game 6 response to force Game 7"
) %>%
  mutate(
    Date = as.Date(Date),
    min_split = str_split(MIN, ":"),
    MIN_num = map_dbl(min_split, ~ as.numeric(.x[1]) + as.numeric(.x[2]) / 60),
    full_workload = MIN_num >= 20,
    STOCKS = STL + BLK
  ) %>%
  select(-min_split)

wemby_full_workload <- wemby_game_log %>%
  filter(full_workload, !is.na(PTS))

wemby_summary <- wemby_full_workload %>%
  summarise(
    GP = n(),
    PPG = mean(PTS, na.rm = TRUE),
    RPG = mean(REB, na.rm = TRUE),
    APG = mean(AST, na.rm = TRUE),
    SPG = mean(STL, na.rm = TRUE),
    BPG = mean(BLK, na.rm = TRUE),
    STOCKS = mean(STOCKS, na.rm = TRUE),
    GmSc = mean(GmSc, na.rm = TRUE),
    .groups = "drop"
  )

wemby_gp <- wemby_summary$GP
wemby_ppg <- wemby_summary$PPG
wemby_rpg <- wemby_summary$RPG
wemby_apg <- wemby_summary$APG
wemby_spg <- wemby_summary$SPG
wemby_bpg <- wemby_summary$BPG
wemby_stocks <- wemby_summary$STOCKS
wemby_games_label <- paste0(wemby_gp, " full-workload games")

wemby_okc_games <- wemby_game_log %>%
  filter(Round_Opp %in% c("OKC Game 1", "OKC Game 4")) %>%
  transmute(
    Date = format(Date, "%m/%d/%Y"),
    Series_Game = str_remove(Round_Opp, "OKC "),
    Result,
    Score = case_when(
      Round_Opp == "OKC Game 1" ~ "SAS 122, OKC 115 (2OT)",
      Round_Opp == "OKC Game 4" ~ "SAS 103, OKC 82",
      TRUE ~ ""
    ),
    PTS,
    REB,
    AST,
    STL,
    BLK,
    Note
  )

signature_games_tbl <- wemby_game_log %>%
  filter(Round_Opp %in% c("POR Game 1", "POR Game 4", "MIN Game 1", "MIN Game 3", "OKC Game 1", "OKC Game 4", "OKC Game 6")) %>%
  transmute(
    Opponent = Round_Opp,
    Line = case_when(
      Round_Opp == "MIN Game 1" ~ paste0(PTS, " / ", REB, " / ", AST, " / ", BLK, " BLK"),
      Round_Opp == "POR Game 4" ~ paste0(PTS, " / ", REB, " / ", BLK, " BLK"),
      TRUE ~ paste0(PTS, " / ", REB, " / ", AST)
    ),
    `Why It Mattered` = Note
  )

wemby_series_tbl <- tribble(
  ~Series, ~Games, ~Record, ~`PTS / REB / AST`, ~`STL + BLK`, ~Signature,
  "Round 1 vs Portland", "3 full games", "3-1", "26.3 / 10.3 / 2.3", "6.7", "35-point debut; defensive closeout after injury scare",
  "Round 2 vs Minnesota", "5 full games", "4-2", "23.2 / 13.6 / 3.0", "5.2", "12-block playoff record; 39/15/5 takeover",
  "WCF vs Oklahoma City", "6 full games", "3-3", "28.2 / 11.5 / 3.3", "5.0", "41/24, 33 to even the series, 28/10 to force Game 7"
)

first_playoff_runs <- tibble::tibble(
  Player = c(
    "Michael Jordan",
    "LeBron James",
    "Kareem Abdul-Jabbar",
    "Magic Johnson",
    "Tim Duncan",
    "Shaquille O'Neal",
    "Hakeem Olajuwon",
    "Luka Doncic",
    "Victor Wembanyama"
  ),
  Season = c(
    "1984-85",
    "2005-06",
    "1969-70",
    "1979-80",
    "1997-98",
    "1993-94",
    "1984-85",
    "2019-20",
    "2025-26"
  ),
  Age = c(21, 21, 22, 20, 21, 21, 22, 21, 22),
  Team = c("CHI", "CLE", "MIL", "LAL", "SAS", "ORL", "HOU", "DAL", "SAS"),
  Team_Record = c(
    "38-44",
    "50-32",
    "56-26",
    "60-22",
    "56-26",
    "50-32",
    "48-34",
    "43-32",
    "62-20"
  ),
  Seed = c(
    "7th East",
    "4th East",
    "2nd East",
    "1st West",
    "5th West",
    "4th East",
    "6th West",
    "7th West",
    "2nd West"
  ),
  Result = c(
    "Lost R1",
    "Lost ECSF",
    "Lost EDF",
    "Won Finals",
    "Lost WCSF",
    "Lost R1",
    "Lost R1",
    "Lost R1",
    "Series tied 3-3 in WCF"
  ),
  Context = c(
  "Below-.500 roster around a rookie star",
  "50-win team carried by a young LeBron",
  "Second-year expansion team turned contender",
  "60-win contender with Kareem already in place",
  "Established Spurs core rebounding from injury collapse",
  "Young 50-win Magic team still ahead of schedule",
  "Young Rockets team before the title-era build",
  "Bubble 7-seed built around Luka's creation",
  "62-win Spurs team with a young core still arriving"
  ),
  GP = c(4, 13, 10, 16, 9, 3, 5, 6, wemby_gp),
  PPG = c(29.3, 30.8, 35.2, 18.3, 20.7, 20.7, 21.2, 31.0, wemby_ppg),
  RPG = c(5.8, 8.1, 16.8, 10.5, 9.0, 13.3, 13.0, 9.8, wemby_rpg),
  APG = c(8.5, 5.8, 4.1, 9.4, 1.9, 2.3, 1.4, 8.7, wemby_apg),
  SPG = c(2.8, 1.4, NA, 3.1, 0.6, 0.7, 1.4, 1.2, wemby_spg),
  BPG = c(1.0, 0.7, NA, 0.4, 2.6, 3.0, 2.6, 0.5, wemby_bpg)
) %>%
  mutate(
    STOCKS = SPG + BPG,
    `PTS / REB / AST` = fmt_statline(PPG, RPG, APG),
    `STL + BLK` = fmt_stockline(SPG, BPG)
  )

cf_debut_pts <- tribble(
  ~Rank, ~Player, ~PTS, ~Date, ~Season, ~Team, ~Opponent,
  1, "Trae Young", 48, "6/23/2021", "2020-21", "ATL", "MIL",
  2, "Victor Wembanyama", 41, "5/18/2026", "2025-26", "SAS", "OKC",
  3, "Amar'e Stoudemire", 41, "5/22/2005", "2004-05", "PHX", "SAS",
  4, "Devin Booker", 40, "6/20/2021", "2020-21", "PHX", "LAC",
  5, "Kevin Durant", 40, "5/17/2011", "2010-11", "OKC", "DAL",
  6, "Carmelo Anthony", 39, "5/19/2009", "2008-09", "DEN", "LAL",
  7, "Dirk Nowitzki", 38, "5/19/2003", "2002-03", "DAL", "SAS",
  8, "Anthony Davis", 37, "9/18/2020", "2019-20", "LAL", "DEN",
  9, "Stephen Curry", 34, "5/19/2015", "2014-15", "GSW", "HOU",
  10, "Deron Williams", 34, "5/20/2007", "2006-07", "UTA", "SAS",
  11, "Boris Diaw", 34, "5/24/2006", "2005-06", "PHX", "DAL",
  12, "Allen Iverson", 34, "5/22/2001", "2000-01", "PHI", "MIL",
  13, "George Gervin", 34, "5/4/1979", "1978-79", "SAS", "WAS"
)

playoff_40_20 <- tribble(
  ~Rank, ~Player, ~Date, ~PTS, ~REB, ~Team, ~Opponent,
  1, "Victor Wembanyama", "5/18/2026", 41, 24, "SAS", "OKC",
  2, "Nikola Jokic", "5/5/2025", 42, 22, "DEN", "OKC",
  3, "Giannis Antetokounmpo", "5/13/2022", 44, 20, "MIL", "BOS",
  4, "Shaquille O'Neal", "6/6/2001", 44, 20, "LAL", "PHI",
  5, "Shaquille O'Neal", "5/8/2001", 43, 20, "LAL", "SAC",
  6, "Shaquille O'Neal", "5/6/2001", 44, 21, "LAL", "SAC",
  7, "Shaquille O'Neal", "6/9/2000", 40, 24, "LAL", "IND",
  8, "David Robinson", "4/28/1996", 40, 21, "SAS", "PHX",
  9, "Charles Barkley", "6/5/1993", 44, 24, "PHX", "SEA",
  10, "Hakeem Olajuwon", "4/30/1988", 41, 26, "HOU", "DAL",
  11, "Hakeem Olajuwon", "5/14/1987", 49, 25, "HOU", "SEA",
  12, "Moses Malone", "4/26/1981", 42, 23, "HOU", "KCK",
  13, "Kareem Abdul-Jabbar", "5/1/1977", 43, 20, "LAL", "GSW",
  14, "Bob McAdoo", "4/18/1975", 50, 21, "BUF", "WAS",
  15, "Kareem Abdul-Jabbar", "4/18/1974", 44, 21, "MIL", "CHI",
  16, "Wilt Chamberlain", "5/6/1970", 45, 27, "LAL", "NYK",
  17, "Kareem Abdul-Jabbar", "4/3/1970", 46, 25, "MIL", "PHI"
)

youngest_40_playoff <- tribble(
  ~Rank, ~Player, ~Age, ~PTS, ~Date, ~Team, ~Opponent, ~MIN, ~REB, ~AST, ~STL, ~BLK,
  1, "Magic Johnson", "20-276", 42, "5/16/1980", "LAL", "PHI", 47, 15, 7, 3, 1,
  2, "LeBron James", "21-119", 41, "4/28/2006", "CLE", "WAS", 48, 5, 3, 2, 0,
  3, "LeBron James", "21-124", 45, "5/3/2006", "CLE", "WAS", 46, 7, 6, 2, 0,
  4, "Luka Doncic", "21-171", 42, "8/17/2020", "DAL", "LAC", 38, 7, 9, 3, 0,
  5, "Luka Doncic", "21-177", 43, "8/23/2020", "DAL", "LAC", 46, 17, 13, 2, 1,
  6, "Anthony Edwards", "21-257", 41, "4/19/2023", "MIN", "DEN", 42, 2, 4, 2, 3,
  7, "Ja Morant", "21-289", 47, "5/26/2021", "MEM", "UTA", 43, 4, 7, 0, 0,
  8, "Tracy McGrady", "21-339", 42, "4/28/2001", "ORL", "MIL", 50, 8, 10, 3, 2,
  9, "Luka Doncic", "22-089", 44, "5/28/2021", "DAL", "LAC", 39, 9, 9, 1, 0,
  10, "Luka Doncic", "22-094", 42, "6/2/2021", "DAL", "LAC", 43, 8, 14, 1, 0,
  11, "Luka Doncic", "22-098", 46, "6/6/2021", "DAL", "LAC", 45, 7, 14, 0, 0,
  12, "Victor Wembanyama", "22-134", 41, "5/18/2026", "SAS", "OKC", 49, 24, 3, 1, 3
)
```

# Introduction

Victor Wembanyama’s playoff run is forcing a hard-to-address question earlier than expected.

The stat lines are already historic: 35 in his playoff debut, 39 against the defense that had just slowed the 3x MVP, 41/24 in a Conference Finals debut, and multiple signature games against the defending champions.

But the numbers only tell part of the story. The bigger question is why this run is different from the usual young-superstar breakout.

# The Run So Far

```{r run-snapshot, results='asis'}
run_snapshot_tbl <- tibble(
  Games = wemby_games_label,
  PPG = fmt_num(wemby_ppg, 1),
  RPG = fmt_num(wemby_rpg, 1),
  APG = fmt_num(wemby_apg, 1),
  BPG = fmt_num(wemby_bpg, 1),
  `STL + BLK` = fmt_num(wemby_stocks, 1),
  `TS%` = fmt_num(64.4)
)

style_kable(
  run_snapshot_tbl,
  "Wembanyama first playoff run snapshot; full-workload games only",
  align = "ccccccc",
  font_size = 8.6
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:7, bold = TRUE)
```

The two early-exit games remain part of the full playoff record, but the primary run averages use games where Wembanyama played at least 20 minutes.

# First Playoff Runs Matter 

Before getting to the single-game leaderboards, the better starting point is the run itself. The all-time greats usually announce themselves early, but not under identical conditions. Some entered rebuilding situations, others joined rosters already positioned to win. That does not erase the production, but it changes how quickly the playoff stage arrived and how ready they were for the bright lights.

```{r first-run-production-table, results='asis'}
first_run_production_tbl <- first_playoff_runs %>%
  transmute(
    Player,
    Age,
    Season,
    GP,
    `PTS / REB / AST`,
    `STL + BLK`
  )

first_prod_kable <- style_kable(
  first_run_production_tbl,
  "First playoff run production by comparable all-time greats",
  align = "lccccc",
  font_size = 8.0
) %>%
  column_spec(1, bold = TRUE, width = "3.0cm") %>%
  column_spec(5, bold = TRUE) %>%
  column_spec(6, bold = TRUE)

highlight_wemby_row(first_prod_kable, first_run_production_tbl)
```

```{r first-run-context-table, results='asis'}
first_run_context_tbl <- first_playoff_runs %>%
  transmute(
    Player,
    Team,
    Record = Team_Record,
    Seed,
    Context,
    Result
  )

first_context_kable <- style_kable(
  first_run_context_tbl,
  "Team context for first playoff runs (Source: BasketballReference)",
  align = "lcccc",
  font_size = 8.0
) %>%
  column_spec(1, bold = TRUE, width = "3.0cm") %>%
  column_spec(5, width = "4.0cm")

highlight_wemby_row(first_context_kable, first_run_context_tbl)
```

The Spurs are not simply benefiting from Wembanyama’s production. Their entire gameplan depends on him, and their playoff ceiling is tied directly to his presence on the floor. Across the `r wemby_gp` games where he has played a full workload, this already belongs among the best first playoff runs we have seen.

If it ends in a championship, the conversation changes completely. At that point, the question may not be whether it belongs near the top of the list, but whether any player this young has ever carried a more demanding two-way postseason burden in the modern era.

# Impact On Winning

```{r defensive-on-off-table, results = 'asis'}
wemby_defense_onoff_tbl <- tribble(
  ~Metric, ~`Wemby On`, ~`Wemby Off`, ~Diff,
  "Defensive Rating", "105.7", "114.9", "-9.2",
  "Opponent eFG%", "49.3%", "54.1%", "-4.8 pp",
  "Opponent 2PT%", "47.4%", "55.2%", "-7.8 pp",
  "Rim Frequency Allowed", "23.4%", "28.2%", "-4.8 pp",
  "Opponent Rim FG%", "61.0%", "65.5%", "-4.5 pp",
  "Opponent Short Mid FG%", "38.1%", "46.5%", "-8.4 pp",
  "Opponent Long Mid FG%", "36.2%", "46.9%", "-10.7 pp"
)

style_kable(
  wemby_defense_onoff_tbl,
  "How Wembanyama changes San Antonio's playoff defensive environment (Source: PBPStats.com)",
  align = "lccc",
  font_size = 8.4
) %>%
  column_spec(1, bold = TRUE, width = "5.1cm") %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(4, bold = TRUE)
```

Wembanyama's defensive impact goes far beyond blocks.

With him on the floor, opponents' entire gameplan shifts. Rim attempts fall, finishing efficiency drops, and San Antonio allows 9.2 fewer points per 100 possessions. Opponents are shooting nearly eight percentage points worse on two-point attempts and over 10 percentage points worse on long midrange shots.

# Offense: Still Developing

```{r shot-diet, results = 'asis'}
wemby_shot_profile <- tribble(
  ~Zone, ~FGM, ~FGA, ~`FG%`, ~`eFG%`, ~`AST%`, ~`UAST%`,
  "Restricted Area", 72, 96, "75.0%", "75.0%", "59.7%", "40.3%",
  "Paint (Non-RA)", 15, 35, "42.9%", "42.9%", "66.7%", "33.3%",
  "Midrange", 5, 25, "20.0%", "20.0%", "60.0%", "40.0%",
  "3PT", 22, 65, "33.8%", "50.8%", "77.3%", "22.7%"
)

style_kable(
  wemby_shot_profile,
  "Wembanyama playoff shooting profile by zone (Source: NBA.com)",
  align = "lcccccc",
  font_size = 8.2
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(4, bold = TRUE)
```

Wembanyama’s offense is still a little clunky. That might be the scariest part.

At the rim, he is already overwhelming: 75.0% in the restricted area on 96 attempts. And it is not just spoon-fed finishing. Over 40% of his makes at the rim have been unassisted, which is absurd self-creation for a 7-foot-5 player.

The jumper is still inconsistent. The midrange has mostly clanked, even when he gets to his fadeaway cleanly. The three-point shot has cooled as the run has gone on, now sitting at 33.8% on 65 attempts, but the volume and range still force defenses to guard him well beyond the paint.

Some possessions are as simple as throwing it up and trusting Wembanyama is somewhere near the rim. Others come from teammates shooting with the confidence that he is waiting behind the play to clean misses up.

# How the Run Built

```{r series-by-series-table, results='asis'}
# ---- Series-level run summary ----
style_kable(
  wemby_series_tbl,
  "Wembanyama's playoff run by series",
  align = "lccccl",
  font_size = 7.9
) %>%
  column_spec(1, bold = TRUE, width = "3.1cm") %>%
  column_spec(4, bold = TRUE) %>%
  column_spec(6, width = "5.8cm")
```

The 41/24 against Oklahoma City will be the game people remember, but it was not the first time Wembanyama put his stamp on this run.

He scored 35 in his playoff debut against Portland, overcame an early injury scare, then closed the series with a defensive performance that erased whatever hope remained. Minnesota entered the second round with momentum after its first-round upset, and Wembanyama answered with the all-time playoff blocks record, 14 rebounds per game for the series, and a 39/15/5 takeover two games later.

Oklahoma City has been the ultimate test. Against the defending champs, fresh off two straight sweeps, 64 wins, and led by the 2x MVP, Wemby opened the series with 41 and 24, answered with 33 to pull San Antonio even after falling behind 2-1, then showed out when they needed him most with 28 and 10 to force a Game 7.

Every series has demanded MVP-level production from Wembanyama. Each round has made that job harder, and he continues to deliver.

# Why The Run Already Matters Historically

Each table is measuring something different: scoring translation, physical dominance, defensive ceiling, and age. None of them alone explain the run. Together, they show how unusual this level of postseason responsibility and production is for a player this young.

## Conference Finals Debut Scoring

```{r cf-debut-table, results='asis'}
cf_tbl <- cf_debut_pts %>%
  transmute(
    Rank,
    Player,
    PTS,
    Date,
    Season,
    Team,
    Opp = Opponent
  )

cf_kable <- style_kable(
  cf_tbl,
  "Most points in a Conference Finals debut (Source: StatMuse)",
  align = "clccccc",
  font_size = 8.0
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, bold = TRUE)

highlight_wemby_row(cf_kable, cf_tbl)
```

## The 40/20 Club

```{r forty-twenty-players-table, results='asis'}
forty_twenty_players <- playoff_40_20 %>%
  count(Player, name = "Times") %>%
  arrange(desc(Times), Player)

ft_kable <- style_kable(
  forty_twenty_players,
  "Players with 40-point, 20-rebound playoff games since 1970 (Source: StatMuse)",
  align = "lc",
  font_size = 8.2
) %>%
  column_spec(1, bold = TRUE)

highlight_wemby_row(ft_kable, forty_twenty_players)
```

## Playoff Shot Blocking Peaks

```{r playoff-blocks-table, results='asis'}
blocks_tbl <- tribble(
  ~Rank, ~Player, ~BLK, ~Date, ~Team,
  1, "Victor Wembanyama", 12, "2026-05-04", "SAS",
  2, "Andrew Bynum", 10, "2012-04-29", "LAL",
  2, "Hakeem Olajuwon", 10, "1990-04-29", "HOU",
  2, "Mark Eaton", 10, "1985-04-26", "UTA"
)

blocks_kable <- style_kable(
  blocks_tbl,
  "Most blocks in a playoff game (Source: NBA.com)",
  align = "clccc",
  font_size = 8.0
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, bold = TRUE)

highlight_wemby_row(blocks_kable, blocks_tbl)
```

## Youngest 40-Point Playoff Games

```{r young-forty-table, results='asis'}
young_tbl <- youngest_40_playoff %>%
  group_by(Player) %>%
  slice_min(Age, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(Age) %>%
  transmute(
    Player,
    Age,
    PTS,
    Date,
    Team,
    Opp = Opponent,
    REB,
    AST,
    BLK
  )

young_kable <- style_kable(
  young_tbl,
  "Youngest players to score 40+ points in a playoff game, one entry per player (Source: StatMuse)",
  align = "lcccccccc",
  font_size = 7.8
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, bold = TRUE)

highlight_wemby_row(young_kable, young_tbl)
```

The 41/24 game put Wembanyama in old-school big-man company, the 12-block game put him alone atop a defensive leaderboard, and the age table puts the timing in perspective. The same run is touching scoring, rebounding, rim protection, and developmental history at once.

# Context Matters

```{r spurs-support-table, results='asis'}
spurs_support_tbl <- tribble(
  ~Player, ~Age, ~Role, ~MPG, ~PPG, ~RPG, ~APG, ~`TOV/G`, ~`TS%`,
  "Stephon Castle", 21, "Secondary creator", 33.3, 19.4, 4.9, 6.7, 3.5, "61.6%",
  "De'Aaron Fox", 28, "Lead guard", 32.4, 16.5, 4.2, 6.0, 1.8, "51.0%",
  "Devin Vassell", 25, "Spacing / wing scoring", 33.9, 13.1, 4.8, 2.6, 1.1, "55.4%",
  "Dylan Harper", 19, "Young scoring guard", 25.4, 13.1, 5.2, 2.6, 1.6, "62.1%",
  "Julian Champagnie", 24, "Connector / spacing wing", 30.0, 10.8, 5.8, 1.5, 0.4, "60.9%"
)

style_kable(
  spurs_support_tbl,
  "San Antonio's main playoff support around Wembanyama through Game 6 through Game 6",
  align = "lclcccccc",
  font_size = 8.0
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, width = "3.8cm") %>%
  column_spec(9, bold = TRUE)
```

There is an obvious caveat: Wembanyama has help.

NBA history is not built by isolated players. Magic had Kareem. Duncan had David Robinson and a complete Spurs infrastructure. Kobe entered alongside Shaq. Even Jordan and LeBron needed the right roster construction before championships followed.

The Spurs have real help around Wembanyama, too. Castle has played like a legitimate two-way cornerstone, Harper has supplied downhill pressure and timely scoring, and Vassell and Champagnie have consistently stretched defenses with their shooting. Fox has been less consistent offensively than expected, but his ability to handle creation duties still gives San Antonio another source of ball pressure.

But situation does not explain everything. The Spurs have enough talent around Wembanyama to make this run possible, but he is the reason it has become historic.

# Closing

Maybe this remains a historic run inside a merely great career. Maybe the Spurs’ roster context makes the timeline look clearer than it actually is. Maybe injuries, regression, or matchup variance interrupt what feels inevitable right now. Basketball history is full of trajectories that looked obvious until they weren’t.

With that said, Wembanyama is not just putting up numbers that are absurd for his age. He is the superstar leader behind one of the best teams on the planet at 22 years old, with a competitive edge that seems to run up and down the roster. San Antonio is no longer winning on the idea of what he could become. They are winning because of what he already is, arguably the best player in the world right now and already one of the most overwhelming defensive forces the game has ever seen.

For the past two decades, the GOAT debate has mostly revolved around two names. Wembanyama has not entered that arena yet, but this run has shown why the path is visible. The offense is already superstar-level, and the defensive gap he is creating looks unlike anything in the modern game. If that combination holds, his ceiling becomes something basketball has rarely, if ever, seen.
