---
title: "Darius Garland for James Harden: A Timeline Trade"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
params:
  season: "2025-26"
  as_of_date: "2026-02-01"
  use_cache_files: TRUE
  cache_dir: "data"
  garland_zones_xlsx: "Darius Garland Shot Zones (Last 3 Seasons).xlsx"
  harden_zones_xlsx: "James Harden Shot Zones (Last 3 Seasons).xlsx"
  ast_leaderboard_xlsx: "Ast Leaderboard 2025-26.xlsx"
  garland_playtypes_xlsx: "Darius Garland Playtypes (Last 3 Seasons).xlsx"
  harden_playtypes_xlsx: "James Harden Playtypes (Last 3 Seasons).xlsx"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.pos = "H",
  dpi = 300,
  out.width = "100%"
)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(scales)
library(knitr)
library(kableExtra)
library(hoopR)
library(lubridate)
library(janitor)
library(showtext)
library(sysfonts)
```

```{r constants-and-parameters}
# Report parameters
season_used <- params$season
as_of_date <- params$as_of_date
cache_dir <- params$cache_dir
use_cache_files <- isTRUE(params$use_cache_files)

# Theme constants
bg_color   <- "#f6f4ef"
text_dark  <- "#141414"
grid_color <- "#dedad3"

# Trade framing colors
cavs_wine <- "#6F263D"
clippers_blue <- "#1D428A"

player_colors <- c(
  "James Harden" = cavs_wine,
  "Darius Garland" = clippers_blue
)

plot_credits <- "Source: NBA.com | Viz: @Rambzee_"

sysfonts::font_add_google("Inter", "inter")
showtext::showtext_auto()
```

```{r helper-functions}
theme_caleb <- function() {
  theme_minimal(base_family = "inter") +
    theme(
      plot.background  = element_rect(fill = bg_color, color = NA),
      panel.background = element_rect(fill = bg_color, color = NA),
      panel.grid.major = element_line(color = grid_color, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 16, face = "bold", hjust = 0.5, color = text_dark),
      plot.subtitle    = element_text(size = 10, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 8)),
      axis.title       = element_text(size = 12, face = "bold", color = text_dark),
      axis.text        = element_text(size = 10, color = "#333333"),
      plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 0, margin = margin(t = 8)),
      legend.position  = "top",
      plot.margin      = margin(10, 12, 14, 12)
    )
}

style_table <- function(table_obj, font_size = 9, scale_down = FALSE) {
  latex_options <- c("striped", "hold_position")

  if (scale_down) {
    latex_options <- c(latex_options, "scale_down")
  }

  table_obj %>%
    kableExtra::kable_styling(
      full_width = FALSE,
      latex_options = latex_options,
      font_size = font_size
    )
}

file_required <- function(path) {
  if (!file.exists(path)) {
    stop(paste0("Missing file: ", path, ". Put this Excel file in the same folder as the RMD or update the YAML params."))
  }

  path
}
```

```{r hand-entered-data}
intro_tbl <- tibble::tribble(
  ~Player, ~GP, ~MPG, ~PPG, ~APG, ~FGA, ~FTA, ~`3P%`, ~`TS%`, ~`+/-`,
  "James Harden", 44, 35.4, 25.4, 8.1, 17.5, 8.5, 34.7, 59.8, 20,
  "Darius Garland", 26, 30.5, 18.0, 6.9, 14.5, 3.0, 36.0, 56.7, -40
)

pullup_context <- tibble(
  season_group = c("2023–24 and 2024–25", "2025–26"),
  pullup_3_pct = c(38.7, 33.0),
  attempts = c(506, 79)
)

wowy_tbl <- tibble::tribble(
  ~player, ~minutes_on, ~off_on, ~def_on, ~minutes_off, ~off_off, ~def_off, ~delta_off, ~delta_def, ~delta_net,
  "James Harden", 6818, 119.7, 115.0, 3451, 112.2, 111.9, 7.5, -3.1, 4.3,
  "Darius Garland", 4993, 120.3, 115.5, 5377, 117.1, 111.2, 3.2, 4.3, -1.1
)

garland_split <- tibble::tribble(
  ~Context, ~Minutes, ~USG, ~PTS, ~AST, ~REB, ~TOV, ~TS, ~FTA, ~ThreeP, ~Net,
  "With Mitchell", 7394, 23.4, 19.6, 7.5, 2.7, 2.9, 57.9, 3.1, 38.3, 6.8,
  "Without Mitchell", 3646, 29.5, 25.4, 8.3, 3.4, 3.4, 58.6, 5.2, 40.0, 3.0
)

garland_delta <- garland_split %>%
  summarise(
    Context = "Δ (Without − With)",
    Minutes = NA,
    USG = diff(USG),
    PTS = diff(PTS),
    AST = diff(AST),
    REB = diff(REB),
    TOV = diff(TOV),
    TS = diff(TS),
    FTA = diff(FTA),
    ThreeP = diff(ThreeP),
    Net = diff(Net)
  )

garland_tbl <- bind_rows(garland_split, garland_delta)
```

```{r local-file-inputs}
# Local source extracts. Do not commit restricted/proprietary files publicly unless you have permission.
garland_zones <- read_excel(file_required(params$garland_zones_xlsx)) %>%
  janitor::clean_names()

harden_zones <- read_excel(file_required(params$harden_zones_xlsx)) %>%
  janitor::clean_names()

ast_tbl <- read_excel(file_required(params$ast_leaderboard_xlsx)) %>%
  janitor::clean_names()

garland_play <- read_excel(file_required(params$garland_playtypes_xlsx)) %>%
  janitor::clean_names()

harden_play <- read_excel(file_required(params$harden_playtypes_xlsx)) %>%
  janitor::clean_names()
```

```{r data-cleaning}
zones_long <- bind_rows(
  garland_zones %>% mutate(player = "Darius Garland"),
  harden_zones %>% mutate(player = "James Harden")
) %>%
  mutate(
    season = as.character(season),
    zone = str_to_title(zone),
    fg_percent = as.numeric(fg_percent),
    fga = as.numeric(fga),
    season = factor(season, levels = sort(unique(season)))
  )

play_3yr <- bind_rows(
  garland_play %>% mutate(player = "Darius Garland"),
  harden_play %>% mutate(player = "James Harden")
) %>%
  mutate(
    possessions = as.numeric(possessions),
    percent_time = as.numeric(percent_time),
    playtype = case_when(
      stringr::str_detect(playtype, regex("pnr", ignore_case = TRUE)) ~ "P&R Ballhandler",
      TRUE ~ playtype
    )
  ) %>%
  group_by(player, playtype) %>%
  summarise(
    total_possessions = sum(possessions, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(player) %>%
  mutate(share = total_possessions / sum(total_possessions)) %>%
  ungroup()

playtype_levels <- c(
  "P&R Ballhandler",
  "Isolation",
  "Transition",
  "Spot-Up",
  "Handoffs"
)

play_3yr <- play_3yr %>%
  mutate(playtype = factor(playtype, levels = playtype_levels))

play_3yr_tbl <- bind_rows(
  garland_play %>% mutate(player = "Darius Garland"),
  harden_play %>% mutate(player = "James Harden")
) %>%
  transmute(
    player,
    playtype,
    poss_w = readr::parse_number(as.character(possessions)),
    pts_w = readr::parse_number(as.character(points)),
    ppp_pct = suppressWarnings(as.numeric(ppp_rank))
  ) %>%
  filter(!is.na(poss_w), poss_w > 0, !is.na(pts_w)) %>%
  group_by(player, playtype) %>%
  summarise(
    possessions = sum(poss_w, na.rm = TRUE),
    points = sum(pts_w, na.rm = TRUE),
    ppp = points / possessions,
    ppp_pctile_w = {
      ok <- !is.na(ppp_pct) & !is.na(poss_w) & poss_w > 0
      if (sum(ok) == 0) NA_real_ else sum(ppp_pct[ok] * poss_w[ok]) / sum(poss_w[ok])
    },
    .groups = "drop"
  ) %>%
  group_by(player) %>%
  mutate(share = possessions / sum(possessions)) %>%
  slice_max(order_by = possessions, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(player, desc(share)) %>%
  mutate(
    ppp_tier = case_when(
      is.na(ppp_pctile_w) ~ NA_character_,
      ppp_pctile_w < 0.45 ~ "Below average",
      ppp_pctile_w < 0.55 ~ "Average",
      ppp_pctile_w < 0.65 ~ "Good",
      ppp_pctile_w < 0.80 ~ "Very good",
      TRUE ~ "Excellent"
    )
  )
```

# Executive Summary

A straight-up star swap is rare in the modern NBA, especially one with a ten-year age gap, but this trade reflects two teams aligning with different competitive timelines.

Cleveland prioritizes availability and postseason reliability. Garland’s recent seasons have been disrupted by injuries at inopportune moments, while Harden has played fewer than 62 games in just two of his 17 NBA seasons. The move signals a push to maximize Cleveland’s title window around Donovan Mitchell and Evan Mobley, while preserving long-term flexibility to pivot around Mitchell, Mobley, and a stable core of wing 3-and-D defenders if needed.

For the Clippers, the bet is on ceiling. Garland offers a younger offensive engine entering his prime, elevating Los Angeles’ upside through elite pull-up shooting gravity, touch, and advantage creation. With no control of their own first-round picks until 2030, the Clippers are betting that a high-level young guard can help sustain winning seasons as Kawhi Leonard enters the latter stages of his career.

# 1) Overview

```{r overview-table}
intro_tbl %>%
  knitr::kable(
    digits = 1,
    caption = "2025–26 Season Overview (Through Midseason)"
  ) %>%
  style_table()
```

The midseason numbers reflect differences in role and responsibility more than differences in talent. Harden carries a heavier offensive load, producing higher scoring and playmaking volume while relying on foul drawing to sustain efficiency. Garland’s output has been more limited by injury and shared creation, resulting in lower raw totals despite strong shooting efficiency.

# 2) Scoring Profile and Efficiency: Last 3 Seasons

## 2A) Shot Zone Efficiency

```{r shot-zone-efficiency-plot}
ggplot(
  zones_long,
  aes(
    x = season,
    y = fg_percent,
    group = player,
    color = player,
    linetype = player
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ zone) +
  scale_y_continuous(limits = c(30, 70)) +
  scale_color_manual(values = player_colors) +
  labs(
    title = "Shot Zone FG% Trends (Last 3 Seasons)",
    x = NULL,
    y = "FG%",
    color = NULL,
    linetype = NULL,
    caption = plot_credits
  ) +
  theme_caleb() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 2B) Shot Diet Visualization

```{r shot-zone-barplot, fig.height=4.5}
zones_long %>%
  group_by(player) %>%
  mutate(share = fga / sum(fga, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(zone, share, fill = player)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Shot Distribution by Zone (Last 3 Seasons)",
    x = NULL,
    y = "Attempt Share",
    fill = NULL,
    caption = plot_credits
  ) +
  theme_caleb()
```

Taken together, the shot distribution and efficiency trends highlight a stylistic contrast rather than a gap in quality. Both guards take a similar share of their shots in the paint and from three, but Garland converts those attempts at higher rates, supported by elite floater touch and increased rim pressure. Harden’s efficiency is more volume-driven, sustained by three-point attempts and free-throw generation, while Garland’s is rooted in touch and shot-making across the paint and intermediate areas.

# 3) Pull-Up Shooting Context

```{r pullup-summary-table}
pullup_context %>%
  kbl(
    digits = 1,
    caption = "Darius Garland Pull-Up 3PT Shooting Context"
  ) %>%
  style_table()
```

Garland’s pull-up three-point percentage has dipped in 2025–26 at 33% on 79 attempts, but that decline is far more about sample size and battling injuries than any erosion of skill. Over the prior two seasons, he shot 38.7% on 506 pull-up attempts, a volume that firmly establishes him as one of the league’s best pull-up shooters.

# 4) Assist Creation Context

```{r ast-leaderboard-table}
ast_tbl %>%
  arrange(desc(potential_ast)) %>%
  transmute(
    Player = player,
    Team = team,
    `Potential AST / G` = potential_ast
  ) %>%
  slice(1:14) %>%
  kbl(
    digits = 1,
    caption = "Potential Assists per Game (2025–26 Season)"
  ) %>%
  style_table()
```

Both guards rank near the top of the league in creation burden and teammate shot generation. James Harden ranks 9th in potential assists per game at 13.5, underscoring his continued role as a high-level primary organizer even as his scoring has tapered in the latter stages of his career. Darius Garland ranks 14th at 12.4, placing him firmly within the same top-tier creation cohort despite operating in a more usage-compressed environment alongside Donovan Mitchell.

# 5) Playtype Portfolio: Last 3 Seasons

## 5A) Playtype Share Comparison

```{r playtype-share-plot, fig.height=5}
ggplot(
  play_3yr,
  aes(
    x = reorder(playtype, -share),
    y = share,
    fill = player
  )
) +
  geom_col(position = "dodge") +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 0.5)
  ) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Top Offensive Playtypes (Last 3 Seasons)",
    x = NULL,
    y = "Share of Possessions",
    fill = NULL,
    caption = plot_credits
  ) +
  theme_caleb()
```

Both Garland and Harden share the same top five playtypes by possession share, as expected for high-level creation guards. The separation comes in how those possessions are allocated. Harden remains one of the league’s highest-volume isolation scorers, using strength, foul drawing, and off-the-dribble three-point shooting to generate offense late in the clock. Garland, meanwhile, is a pick-and-roll maestro and one of the league’s best ball handlers, capable of scoring efficiently in a variety of ways off his handle, pace, and touch.

## 5B) Playtype Efficiency Table

```{r playtype-efficiency-table}
play_3yr_tbl %>%
  mutate(ppp_pctile_w = scales::percent(ppp_pctile_w, accuracy = 1)) %>%
  select(player, playtype, possessions, points, ppp, share, ppp_pctile_w, ppp_tier) %>%
  knitr::kable(
    digits = 2,
    col.names = c("Player", "Playtype", "Poss", "Points", "PPP", "Share", "PPP Rating", "Tier"),
    caption = "Top Playtypes (3-Year Aggregate): Volume, PPP, and PPP Rating Tier."
  ) %>%
  style_table()
```

The efficiency breakdown adds another layer to the stylistic divide. Garland’s impact is spread across actions, where he grades as solid to good across the board, using speed, touch, and shooting versatility to compound advantages, particularly as a lead ball handler in Los Angeles. Harden’s efficiency is more concentrated, anchored by elite isolation production at heavy volume, while his continued effectiveness in pick-and-roll and spot-up roles gives Cleveland dependable offense that should translate with another elite guard.

# 6) On/Off

```{r databallr-wowy-table}
wowy_tbl %>%
  mutate(
    `On Min` = minutes_on,
    `Off Min` = minutes_off,
    `ON: OFF` = off_on,
    `ON: DEF` = def_on,
    `OFF: OFF` = off_off,
    `OFF: DEF` = def_off,
    `ON/OFF OFF` = delta_off,
    `ON/OFF DEF` = delta_def,
    `ON/OFF NET` = delta_net
  ) %>%
  select(
    player,
    `On Min`,
    `Off Min`,
    `ON: OFF`,
    `ON: DEF`,
    `OFF: OFF`,
    `OFF: DEF`,
    `ON/OFF OFF`,
    `ON/OFF DEF`,
    `ON/OFF NET`
  ) %>%
  knitr::kable(
    digits = 1,
    caption = "On/Off (Last 3 Seasons). Source: Databallr"
  ) %>%
  style_table(font_size = 5)
```

The on/off splits reinforce familiar themes. Harden remains a consistent offensive floor raiser, reliably lifting team offense when on the court. Garland also boosts offensive output, though his impact is more lineup-dependent due to size and defensive limitations.

# 7) Garland Without Donovan

```{r garland-mitchell-split-table}
garland_tbl %>%
  kbl(
    digits = 1,
    col.names = c(
      "Context",
      "Minutes",
      "USG%",
      "PTS",
      "AST",
      "REB",
      "TOV",
      "TS%",
      "FTA",
      "3P%",
      "Net +/-"
    ),
    caption = "Darius Garland With vs. Without Donovan Mitchell (Last 4 Seasons, per 75 possessions). Source: Databallr"
  ) %>%
  style_table() %>%
  row_spec(3, bold = TRUE)
```

This helps explain why Garland’s raw on/off splits should not be overweighted. When Mitchell was off the floor, Garland took over. Usage climbed, scoring and assists jumped, and efficiency improved over a very large sample. The net rating dip reflects lineup context and real defensive limitations, not offensive drop-off. When Garland was allowed to run the offense, the upside was clear.

# 8) What Each Team Is Buying

**Cleveland**

- Reliability and durability: Harden brings a long track record of availability and nightly offensive presence, addressing a key postseason health concern.
- Half-court shot creation: An elite isolation scorer who can consistently generate offense alongside Donovan Mitchell when possessions slow.
- Free-throw pressure: One of the league’s best foul drawers, providing efficient scoring and a stabilizing force.
- Defensive survivability: Harden is not a plus defender, but his size and strength allow him to hold up better in playoff settings and avoid being relentlessly targeted possession after possession.

**Los Angeles Clippers**

- Offensive ceiling: Garland adds a prime-age on-ball creator whose scoring and playmaking could scale into one of the league’s best offensive engines.
- Pull-up shooting gravity: Elite off-the-dribble shooting that stretches defenses and simplifies offense for everyone else.
- Advantage creation: Speed and handle allow Garland to create downhill pressure and quick scoring opportunities.
- Long-term relevance: A younger offensive engine capable of sustaining competitiveness as Kawhi Leonard enters the latter stages of his career.

# 9) Verdict: Timeline Lens

**2025–26 Playoffs / 2027 Season**

Advantage: Cleveland. Harden’s durability, foul pressure, and half-court shot creation raise Cleveland’s playoff floor, likely giving them a great possession-to-possession offense. In this context, Harden functions closer to his Brooklyn role: orchestrating at a high level and complementing another elite scorer rather than carrying the entire load.

**2027 Season and Onward**

Advantage: LA Clippers. Garland offers a higher offensive ceiling and better age-curve alignment, with upside tied to health and a clearly defined on-ball role. The Clippers are betting on a prime-age offensive engine capable of driving efficient offense and sustaining winning seasons, a necessary gamble given their lack of incentive to bottom out.
