---
title: "Ayo Dosunmu and the Value of Playoff Stability"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    fig_crop: false
geometry: margin=0.78in
params:
  playtypes_po_xlsx: "Ayo Dosunmu Offensive Playtypes (2025-26 Playoffs).xlsx"
  playtypes_rs_xlsx: "Ayo Dosunmu Offensive Playtypes (2025-26).xlsx"
  shots_xlsx: "Ayo Dosunmu Shot Zones (2025-26).xlsx"
  pnr_leaders_xlsx: "Playoff Leaders (PnR PPP).xlsx"
  three_leaders_xlsx: "Playoff Leaders (3FG%).xlsx"
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  dpi = 300,
  out.width = "95%"
)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(readr)
library(janitor)
library(scales)
library(kableExtra)
library(showtext)
library(sysfonts)
library(forcats)
```

```{r constants-and-parameters}
# Theme constants
bg_color    <- "#f6f4ef"
text_dark   <- "#141414"
text_light  <- "#5f5f5f"
grid_color  <- "#dedad3"

# Timberwolves palette
wolves_navy  <- "#0C2340"
wolves_green <- "#78BE20"
wolves_blue  <- "#236192"
wolves_gray  <- "#9EA2A2"

# Change colors
rise_color <- "#15803d"
drop_color <- "#b91c1c"

caption_line <- "Data: NBA.com & Databallr | Viz: @Rambzee_"

try(font_add_google("Inter", "inter"), silent = TRUE)
showtext_auto()
options(scipen = 999)
```

```{r helper-functions}
theme_ayo <- function(base_size = 11) {
  theme_minimal(base_family = "inter", base_size = base_size) +
    theme(
      plot.background  = element_rect(fill = bg_color, color = NA),
      panel.background = element_rect(fill = bg_color, color = NA),
      panel.grid.major = element_line(color = grid_color, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = text_dark),
      plot.title = element_text(
        size = 19,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 4)
      ),
      plot.subtitle = element_text(
        size = 10.5,
        color = text_light,
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      axis.title = element_text(size = 10.5, face = "bold"),
      axis.text = element_text(size = 9.4, color = "#333333"),
      axis.ticks = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 9.2),
      plot.caption = element_text(size = 7.8, color = "#6e6e6e", hjust = 1, margin = margin(t = 7)),
      plot.margin = margin(10, 12, 12, 12)
    )
}

style_kable <- function(
  df,
  caption = NULL,
  align = NULL,
  font_size = 8.8,
  full_width = FALSE,
  escape = TRUE
) {
  if (is.null(align)) {
    align <- strrep("c", ncol(df))
  }

  kbl(
    df,
    booktabs = TRUE,
    format = "latex",
    caption = caption,
    align = align,
    linesep = "",
    escape = escape
  ) %>%
    kable_styling(
      latex_options = c("hold_position", "striped"),
      full_width = full_width,
      font_size = font_size,
      stripe_color = "#fbfaf7"
    )
}

fmt_num <- function(x, digits = 1) {
  sprintf(paste0("%.", digits, "f"), x)
}

fmt_plus <- function(x, digits = 1) {
  sprintf(paste0("%+.", digits, "f"), x)
}

fmt_pct <- function(x, accuracy = 0.1) {
  scales::percent(x, accuracy = accuracy)
}

scale_pct <- function(x) {
  if_else(abs(x) > 1, x / 100, x)
}

file_required <- function(path) {
  if (!file.exists(path)) {
    stop(paste0("Missing file: ", path, ". Put this Excel file in the same folder as the RMD or update the YAML params."))
  }

  path
}

clean_playtypes <- function(df) {
  df %>%
    clean_names() %>%
    mutate(
      across(
        any_of(c(
          "possessions", "percent_time", "points", "ppp",
          "fgm", "fga", "fg_percent", "e_fg_percent",
          "to_percent", "x2fgm", "x2fga", "x2fg_percent",
          "x3fgm", "x3fga", "x3fga_percent"
        )),
        ~ suppressWarnings(as.numeric(.))
      ),
      ppp_rank_num = suppressWarnings(parse_number(as.character(ppp_rank))),
      percent_time_rank_num = suppressWarnings(parse_number(as.character(percent_time_rank))),
      playtype_clean = case_when(
        str_detect(str_to_lower(playtype), "pnr|pick") ~ "PnR Ballhandler",
        str_detect(str_to_lower(playtype), "transition") ~ "Transition",
        str_detect(str_to_lower(playtype), "spot") ~ "Spot Up",
        str_detect(str_to_lower(playtype), "isolation") ~ "Isolation",
        str_detect(str_to_lower(playtype), "handoff") ~ "Handoffs",
        str_detect(str_to_lower(playtype), "off") ~ "Off Screen",
        str_detect(str_to_lower(playtype), "cut") ~ "Cut",
        TRUE ~ playtype
      )
    )
}
```

```{r hand-input-data}
ayo_team_splits <- tribble(
  ~Split, ~Team, ~GP, ~GS, ~MIN, ~FGM, ~FGA, ~FG_PCT, ~FG3M, ~FG3A, ~FG3_PCT, ~FTM, ~FTA, ~FT_PCT,
  ~OREB, ~DREB, ~REB, ~AST, ~BLK, ~STL, ~PF, ~TOV, ~PTS,
  "Pre-Trade",  "Chicago",   24,  9, 28.9, 5.7, 10.9, 0.521, 1.5, 3.6, 0.414, 1.5, 1.7, 0.925,
  0.6, 3.6, 4.2, 3.5, 0.3, 1.0, 2.2, 1.5, 14.4,
  "Post-Trade", "Minnesota", 45, 10, 26.4, 5.6, 10.8, 0.514, 1.9, 4.3, 0.451, 2.0, 2.3, 0.857,
  0.7, 2.3, 3.0, 3.6, 0.3, 0.8, 1.7, 1.4, 15.0
)

ayo_recent_form <- tribble(
  ~Split, ~GP, ~PTS, ~AST, ~REB, ~TS_PCT, ~FG2_PCT, ~FG3_PCT, ~FT_PCT, ~FTA, ~TOV, ~STL, ~BLK, ~MPG, ~PLUS_MINUS, ~FG2A, ~FG3A, ~FTR,
  "Regular Season", 69, 14.8, 3.6, 3.4, 0.630, 0.563, 0.439, 0.876, 2.1, 1.4, 0.8, 0.3, 27.1, -0.9, 6.8, 4.1, 0.194,
  "Postseason",     5, 21.8, 4.0, 2.8, 0.749, 0.643, 0.545, 0.950, 4.0, 1.4, 0.4, 0.2, 32.4, -1.4, 8.4, 4.4, 0.313
)

bench_40_tbl <- tribble(
  ~Rank, ~Player, ~Points, ~`TS%`,
  1, "Fred Brown", 45, "59.1%",
  2, "Ayo Dosunmu", 43, "96.5%",
  3, "Steph Curry", 40, "60.0%",
  4, "Nick Van Exel", 40, "68.8%"
)

efficient_40_tbl <- tribble(
  ~Rank, ~Player, ~Points, ~`TS%`, ~Playoffs,
  1, "Terry Porter", 41, "101.7%", "1992",
  2, "Damian Lillard", 55, "96.8%", "2021",
  3, "Ayo Dosunmu", 43, "96.5%", "2026",
  4, "Dirk Nowitzki", 48, "93.9%", "2011",
  5, "Devin Booker", 47, "90.8%", "2023",
  6, "Anthony Davis", 43, "90.6%", "2020",
  7, "Devin Booker", 47, "90.5%", "2021",
  8, "Shai Gilgeous-Alexander", 42, "90.2%", "2026",
  9, "Jason Richardson", 42, "89.7%", "2010",
  10, "Jamal Murray", 50, "89.4%", "2020"
)
```

```{r proprietary-file-inputs}
# These Excel inputs are local/proprietary project files.
# Do not commit the raw files to a public GitHub repository unless you have permission.
ayo_playtypes_po <- read_excel(file_required(params$playtypes_po_xlsx)) %>%
  clean_names()

ayo_playtypes_rs <- read_excel(file_required(params$playtypes_rs_xlsx)) %>%
  clean_names()

ayo_shots <- read_excel(file_required(params$shots_xlsx)) %>%
  clean_names()

pnr_leaders <- read_excel(file_required(params$pnr_leaders_xlsx)) %>%
  clean_names()

three_leaders <- read_excel(file_required(params$three_leaders_xlsx)) %>%
  clean_names()
```

```{r data-cleaning}
play_rs <- clean_playtypes(ayo_playtypes_rs)
play_po <- clean_playtypes(ayo_playtypes_po)

play_compare <- play_rs %>%
  select(
    playtype_clean,
    rs_poss = possessions,
    rs_time = percent_time,
    rs_ppp = ppp,
    rs_efg = e_fg_percent,
    rs_tov = to_percent
  ) %>%
  inner_join(
    play_po %>%
      select(
        playtype_clean,
        po_poss = possessions,
        po_time = percent_time,
        po_ppp = ppp,
        po_efg = e_fg_percent,
        po_tov = to_percent
      ),
    by = "playtype_clean"
  ) %>%
  mutate(
    ppp_change = po_ppp - rs_ppp,
    time_change = po_time - rs_time
  ) %>%
  filter(rs_poss >= 8 | po_poss >= 3)

play_compare_long <- play_compare %>%
  select(playtype_clean, rs_ppp, po_ppp) %>%
  pivot_longer(c(rs_ppp, po_ppp), names_to = "Split", values_to = "PPP") %>%
  mutate(
    Split = recode(Split, rs_ppp = "Regular Season", po_ppp = "Postseason"),
    playtype_clean = fct_reorder(playtype_clean, PPP, .fun = max)
  )

play_volume_long <- play_compare %>%
  select(playtype_clean, rs_time, po_time) %>%
  pivot_longer(c(rs_time, po_time), names_to = "Split", values_to = "Time") %>%
  mutate(
    Split = recode(Split, rs_time = "Regular Season", po_time = "Postseason"),
    playtype_clean = fct_reorder(playtype_clean, Time, .fun = max)
  )

shot_summary <- ayo_shots %>%
  transmute(
    zone,
    FGM = as.numeric(fgm),
    FGA = as.numeric(fga),
    FG_PCT = scale_pct(as.numeric(fg_percent))
  ) %>%
  group_by(zone) %>%
  summarise(
    FGM = sum(FGM, na.rm = TRUE),
    FGA = sum(FGA, na.rm = TRUE),
    FG_PCT = sum(FGM, na.rm = TRUE) / sum(FGA, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Share = FGA / sum(FGA),
    zone = factor(
      zone,
      levels = c(
        "Restricted Area",
        "In The Paint (Non-RA)",
        "Midrange",
        "Corner Threes",
        "ATB Threes"
      )
    )
  )

pnr_tbl <- pnr_leaders %>%
  mutate(
    poss_g = as.numeric(poss_g),
    ppp = as.numeric(ppp),
    Rank = row_number()
  ) %>%
  transmute(
    Rank,
    Player = player,
    `Poss/G` = fmt_num(poss_g, 1),
    PPP = fmt_num(ppp, 2)
  )

three_tbl <- three_leaders %>%
  mutate(
    attempts = as.numeric(attempts),
    three_pct = as.numeric(x3fg_percent),
    Rank = row_number()
  ) %>%
  transmute(
    Rank,
    Player = player,
    `3PA/G` = fmt_num(attempts, 1),
    `3P%` = paste0(fmt_num(three_pct, 1), "%")
  )
```

```{r analysis-summary-objects}
regular <- ayo_recent_form %>%
  filter(Split == "Regular Season")

post <- ayo_recent_form %>%
  filter(Split == "Postseason")

ppg_change <- post$PTS - regular$PTS
ts_change <- post$TS_PCT - regular$TS_PCT
ftr_change <- post$FTR - regular$FTR

pnp_ppp <- play_po %>%
  filter(playtype_clean == "PnR Ballhandler") %>%
  pull(ppp) %>%
  .[1]

post_3p <- post$FG3_PCT
```

# Introduction

Ayo Dosunmu has been one of the more interesting bench pieces in the NBA over the last half decade. Even on a middling Bulls roster where young talent often seemed to stall out, his efficiency stood out.

At the deadline, Chicago moved on from its hometown guard. For Dosunmu, the move to Minnesota was an opening. For the Wolves, it was a bet on a guard who could fit around stars while still handling real on-ball responsibility in smaller doses.

That bet has paid off.

Dosunmu took an already efficient offensive profile and added another layer Minnesota’s offense needed: rim pressure, shooting, and enough on-ball creation to take some of the burden off Julius Randle and Anthony Edwards.

Now, with the Wolves up 3-2 despite losing both Edwards and Donte DiVincenzo, that skill set has become central to the series. Denver has made a habit of surviving deficits, and Minnesota cannot afford to let this turn into another one of those comebacks. If Minnesota is going to close this out, Dosunmu has to keep giving them enough creation to avoid offensive droughts.

# Deadline Context

```{r pre-post-table, results='asis'}
pre_post_tbl <- ayo_team_splits %>%
  mutate(
    `FG%` = fmt_pct(FG_PCT),
    `3P%` = fmt_pct(FG3_PCT),
    `FT%` = fmt_pct(FT_PCT)
  ) %>%
  transmute(
    Split,
    Team,
    GP,
    MPG = MIN,
    PTS,
    REB,
    AST,
    STL,
    TOV,
    `FG%`,
    `3P%`,
    `FT%`,
    FTA
  )

style_kable(
  pre_post_tbl,
  "Ayo Dosunmu pre-trade vs post-trade production",
  align = "llccccccccccc"
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE)
```

The post-deadline box score shows why the fit worked. Dosunmu has been able to amplify his best traits with more free throw pressure, better shooting, and the same level of control.

That has carried into the playoffs. The role has expanded in spots, with more on-ball responsibility, but he remains incredibly efficient.

# Playoff Role Shift

```{r recent-form-table, results='asis'}
recent_form_tbl <- ayo_recent_form %>%
  mutate(
    `TS%` = fmt_pct(TS_PCT),
    `2P%` = fmt_pct(FG2_PCT),
    `3P%` = fmt_pct(FG3_PCT),
    `FT%` = fmt_pct(FT_PCT),
    FTr = fmt_pct(FTR),
    `+/-` = fmt_plus(PLUS_MINUS)
  ) %>%
  transmute(
    Split,
    GP,
    PTS,
    AST,
    REB,
    `TS%`,
    `2P%`,
    `3P%`,
    `FT%`,
    FTA,
    TOV,
    MPG,
    `+/-`,
    FG2A,
    FG3A,
    FTr
  )

style_kable(
  recent_form_tbl,
  "Season baseline vs current playoff series",
  align = "lccccccccccccccc"
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(6, bold = TRUE)
```

```{r role-expansion-plot, fig.height=4.4, fig.width=7.4}
role_plot <- ayo_recent_form %>%
  select(Split, PTS, AST, REB, FTA, MPG) %>%
  pivot_longer(-Split, names_to = "Metric", values_to = "Value") %>%
  mutate(
    Split = factor(Split, levels = c("Regular Season", "Postseason")),
    Metric = factor(Metric, levels = c("PTS", "AST", "REB", "FTA", "MPG"))
  )

ggplot(role_plot, aes(x = Metric, y = Value, fill = Split)) +
  geom_col(position = position_dodge(width = 0.74), width = 0.64) +
  geom_text(
    aes(label = fmt_num(Value, 1)),
    position = position_dodge(width = 0.74),
    vjust = -0.35,
    size = 3.0
  ) +
  scale_fill_manual(
    values = c(
      "Regular Season" = wolves_gray,
      "Postseason" = wolves_navy
    )
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Ayo's Role Has Expanded Without More Mistakes",
    subtitle = "Regular season baseline vs first five playoff games",
    x = NULL,
    y = "Per game",
    caption = caption_line
  ) +
  theme_ayo()
```

The jump is not just points. His minutes are up, the shot volume is up, and his free throws have nearly doubled. Dosunmu has taken advantage of a soft Denver defense and consistently punished the gaps it gives him.

More importantly, the added responsibility has not come with more mistakes. More on-ball work usually brings that, but Dosunmu has stayed at **`r post$TOV` turnovers per game**, right in line with his regular-season baseline, even with the added creation.

```{r efficiency-plot, fig.height=4.35, fig.width=7.4}
efficiency_plot <- ayo_recent_form %>%
  select(Split, TS_PCT, FG2_PCT, FG3_PCT, FT_PCT, FTR) %>%
  pivot_longer(-Split, names_to = "Metric", values_to = "Value") %>%
  mutate(
    Split = factor(Split, levels = c("Regular Season", "Postseason")),
    Metric = recode(
      Metric,
      TS_PCT = "TS%",
      FG2_PCT = "2P%",
      FG3_PCT = "3P%",
      FT_PCT = "FT%",
      FTR = "FTr"
    ),
    Metric = factor(Metric, levels = c("TS%", "2P%", "3P%", "FT%", "FTr"))
  )

ggplot(efficiency_plot, aes(x = Metric, y = Value, fill = Split)) +
  geom_col(position = position_dodge(width = 0.74), width = 0.64) +
  geom_text(
    aes(label = fmt_pct(Value)),
    position = position_dodge(width = 0.74),
    vjust = -0.35,
    size = 2.9
  ) +
  scale_fill_manual(
    values = c(
      "Regular Season" = wolves_gray,
      "Postseason" = wolves_navy
    )
  ) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "The Scoring Jump Has Held Efficiency",
    subtitle = "More usage, no clear drop in shot quality",
    x = NULL,
    y = NULL,
    caption = caption_line
  ) +
  theme_ayo()
```

Ayo was already one of the more efficient guards in basketball during the regular season, sitting around +5 rTS. The playoffs have pushed that even further. He is at a staggering 75% TS, which is not sustainable, but it is not built only on three-point variance either.

The more important shift is how often he is getting downhill. Dosunmu has used his quickness to create advantages earlier in possessions, which has opened cleaner looks for both himself and his teammates. That added pressure has shown up in the numbers too, with his free throw rate and true shooting both jumping by nearly 10 percentage points.

# Shot Profile

```{r ayo-shot-profile, results='asis'}
league_avg <- tribble(
  ~Zone, ~LG_PCT,
  "Restricted Area", 0.669,
  "Paint (Non-RA)", 0.445,
  "Midrange", 0.433,
  "3PT", 0.362
)

ayo_shot_profile <- shot_summary %>%
  mutate(
    Zone = case_when(
      zone == "Restricted Area" ~ "Restricted Area",
      zone == "In The Paint (Non-RA)" ~ "Paint (Non-RA)",
      zone == "Midrange" ~ "Midrange",
      zone %in% c("Corner Threes", "ATB Threes") ~ "3PT",
      TRUE ~ as.character(zone)
    )
  ) %>%
  group_by(Zone) %>%
  summarise(
    FGM = sum(FGM, na.rm = TRUE),
    FGA = sum(FGA, na.rm = TRUE),
    FG_PCT = FGM / FGA,
    .groups = "drop"
  ) %>%
  left_join(league_avg, by = "Zone") %>%
  mutate(
    Zone = factor(Zone, levels = c("Restricted Area", "Paint (Non-RA)", "Midrange", "3PT")),
    Volume = paste0(FGA, " att"),
    `FG% (League Avg)` = paste0(fmt_pct(FG_PCT), " (", fmt_pct(LG_PCT), ")"),
    `rFG%` = paste0(fmt_plus(100 * (FG_PCT - LG_PCT)), "%")
  ) %>%
  arrange(Zone) %>%
  select(Zone, Volume, `FG% (League Avg)`, `rFG%`)

style_kable(
  ayo_shot_profile,
  "Ayo Dosunmu shot profile by zone",
  align = "lccc",
  font_size = 9
) %>%
  row_spec(1, bold = TRUE) %>%
  row_spec(4, bold = TRUE)
```

The shot diet fits what Minnesota needed around Edwards, Gobert, and Randle. Dosunmu can run, attack space, and stretch the floor without drifting into a midrange-heavy diet.

That is why the playoff burst does not feel completely random. Some shooting regression is likely, especially from three, but the foundation is strong with consistent rim pressure, enough touch in the paint, strong spot-up value, and added free throw pressure in the postseason. When defenders go under, he can punish them. When they chase, he has enough burst to get downhill.

# Playtype Translation

```{r playtype-efficiency, fig.height=4.7, fig.width=7.4}
play_compare_long_plot <- play_compare_long %>%
  filter(!playtype_clean %in% c("Isolation", "Handoffs"))

ggplot(play_compare_long_plot, aes(x = PPP, y = playtype_clean, fill = Split)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#bcbcbc") +
  geom_text(
    aes(label = fmt_num(PPP, 2)),
    position = position_dodge(width = 0.72),
    hjust = -0.08,
    size = 2.5
  ) +
  scale_fill_manual(
    values = c(
      "Regular Season" = wolves_gray,
      "Postseason" = wolves_navy
    )
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(
    title = "Efficiency Has Improved Across Playtypes",
    subtitle = "Points per possession by playtype, regular season vs postseason",
    x = "Points per possession",
    y = NULL,
    caption = "Data: Synergy / SportsRadar | Viz: @Rambzee_"
  ) +
  theme_ayo() +
  theme(panel.grid.major.y = element_blank())
```

Ayo has been even more effective within the same playtypes that already defined his scoring profile. The most important one is pick-and-roll.

He’s at **`r fmt_num(pnp_ppp, 2)` PPP** as a postseason pick-and-roll ballhandler, even if it’s only about 20 possessions. That sample isn’t enough to treat him like a primary initiator, but it does show that Minnesota can put the ball in his hands, run a simple action, and get a good outcome.

That matters more with Edwards out. The Wolves don’t need Dosunmu to become a high-end creator. They need him to handle second-side actions, attack soft coverage, and keep the offense from sputtering out when the initial play breaks down. So far, he’s been able to do that without forcing possessions or turning the game into isolation-heavy offense.

```{r playtype-volume, fig.height=4.7, fig.width=7.4}
play_volume_long_plot <- play_volume_long %>%
  filter(!playtype_clean %in% c("Isolation", "Handoffs"))

ggplot(play_volume_long_plot, aes(x = Time, y = playtype_clean, fill = Split)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.62) +
  geom_text(
    aes(label = fmt_pct(Time)),
    position = position_dodge(width = 0.72),
    hjust = -0.08,
    size = 2.9
  ) +
  scale_fill_manual(
    values = c(
      "Regular Season" = wolves_gray,
      "Postseason" = wolves_navy
    )
  ) +
  scale_x_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "The Usage Shift Confirms the Eye Test",
    subtitle = "Ayo is carrying more ballhandler offense, not just finishing plays",
    x = "Share of Possessions",
    y = NULL,
    caption = "Data: Synergy / SportsRadar | Viz: @Rambzee_"
  ) +
  theme_ayo() +
  theme(panel.grid.major.y = element_blank())
```

The volume shift is just as important as the efficiency. Transition and spot-ups are still a major part of his value, but the postseason role has included more direct creation, especially as a pick-and-roll ballhandler. That was the part people were less sure he had in him. Through this series, he has shown he can handle more responsibility without moving too far away from what already made him effective.

```{r playtype-table, results='asis'}
playtype_table <- play_compare %>%
  filter(playtype_clean %in% c("Transition", "PnR Ballhandler", "Spot Up")) %>%
  arrange(desc(po_poss)) %>%
  transmute(
    Playtype = playtype_clean,
    `RS Poss` = rs_poss,
    `RS Time%` = fmt_pct(rs_time),
    `RS PPP` = fmt_num(rs_ppp, 3),
    `PO Poss` = po_poss,
    `PO Time%` = fmt_pct(po_time),
    `PO PPP` = fmt_num(po_ppp, 3),
    `PPP Δ` = fmt_plus(ppp_change, 3)
  )

style_kable(
  playtype_table,
  "Ayo Dosunmu core offensive playtypes with the Timberwolves: regular season vs postseason",
  align = "lccccccc",
  font_size = 8.3
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(4, bold = TRUE) %>%
  column_spec(7, bold = TRUE) %>%
  column_spec(8, bold = TRUE)
```

# League Context

The improvement is not just internal. Relative to the rest of the playoff field, Dosunmu has held up well as both a shooter and a pick-and-roll ballhandler.

```{r pnr-table, results='asis'}
ayo_row <- which(pnr_tbl$Player == "Ayo Dosunmu")

style_kable(
  pnr_tbl,
  "Top playoff pick-and-roll scorers by PPP (Min 4 GP, 4 Poss/G)",
  align = "clcc",
  escape = TRUE
) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(ayo_row, bold = TRUE, color = "white", background = wolves_navy)
```

```{r three-table, results='asis'}
ayo_row <- which(three_tbl$Player == "Ayo Dosunmu")

style_kable(
  three_tbl,
  "Top playoff three-point shooters by percentage (Min 4 GP, 3+ 3PA/G)",
  align = "clcc",
  escape = TRUE
) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(ayo_row, bold = TRUE, color = "white", background = wolves_navy)
```

# The Game 4 Swing

```{r game-four-bench-40-table, results='asis'}
ayo_row <- which(bench_40_tbl$Player == "Ayo Dosunmu")

style_kable(
  bench_40_tbl,
  "40+ point playoff games off the bench in NBA history",
  align = "clcc",
  escape = TRUE
) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(ayo_row, bold = TRUE, color = "white", background = wolves_navy)
```

Game 4 was the clearest proof of concept. Edwards and DiVincenzo went down. The Wolves were staring down a team that is more talented, more offensively explosive, powered by one of the best offensive duos ever, and they cut off the faucet defensively while Ayo took over. His heroic 43 point game stamped his name as one of the best bench guards in basketball, and his performance landed in rare bench-company territory.

```{r game-four-efficient-40-table, results='asis'}
ayo_row <- which(efficient_40_tbl$Player == "Ayo Dosunmu")

style_kable(
  efficient_40_tbl,
  "Most efficient 40+ point playoff games in NBA history",
  align = "clccc",
  escape = TRUE
) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(ayo_row, bold = TRUE, color = "white", background = wolves_navy)
```

Beyond the bench angle, his 40-point explosion also ranks among the most efficient playoff scoring games on record.

# Closing

Losing both Edwards and DiVincenzo is a massive hit to Minnesota’s deep playoff hopes. Dosunmu was not brought in to carry this kind of offensive burden, but the injuries have pushed him into a role larger than the Wolves likely expected when they traded for him.

So far, he has given them a real answer. Minnesota’s game plan has been strong, and the execution has mostly matched it, but Denver is not the kind of team you can treat as finished. If the Wolves are going to close out a former championship core, they need steady contributions across the roster. Dosunmu’s efficiency has become one of the more reliable ways they have managed the non-Edwards stretches.

That matters beyond this series as well. If Minnesota advances before Edwards returns, Dosunmu will likely keep carrying more creation responsibility than expected. He does not have to save the offense by himself, but he does have to keep giving the Wolves enough downhill pressure, shooting, and decision-making to give their offense a chance.

For a bench guard, this is already one of the better offensive playoff series on record: real scoring volume, absurd efficiency, and a 43-point game that sits near the top of multiple historical lists. Dosunmu cannot be held to the standard he has set so far, but this stretch has shown why Minnesota targeted him in the first place.

He can fit around stars, but he is not dependent on them. For a Wolves team trying to survive long enough for its franchise player to return, that flexibility has become one of the biggest reasons its season is still alive.
