---
title: "Jaren Jackson Jr.: Stocks > Rebounds and the Rebounding Trade-Off"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
params:
  player_id: 1628991
  player_name: "Jaren Jackson Jr."
  seasons_jjj: ["2018-19","2019-20","2020-21","2021-22","2022-23","2023-24","2024-25","2025-26"]
  # Playoffs included ONLY for Stocks > Rebounds totals/rates and stocks tables.
  include_playoffs_for_stocks: true
  sleep_s: 0.7
  cache_dir: "cache_hoopr"
  export_dir: "exports"
  # Peer REB/36 comparison window (must match the xlsx files below)
  seasons_peer: ["2021-22","2022-23","2023-24","2024-25","2025-26"]
  peer_files:
    - "RebPer36_GreaterThan68_2021_22.xlsx"
    - "RebPer36_GreaterThan68_2022_23.xlsx"
    - "RebPer36_GreaterThan68_2023_24.xlsx"
    - "RebPer36_GreaterThan68_2024_25.xlsx"
    - "RebPer36_GreaterThan68_2025_26.xlsx"
  min_minutes_peer: 900
  min_games_peer: 20
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center"
)

library(dplyr); library(tidyr); library(purrr)
library(stringr); library(readr)
library(ggplot2); library(scales)
library(knitr); library(kableExtra)
library(readxl)
library(hoopR)

# Fonts + showtext
library(showtext)
sysfonts::font_add_google("Poppins", "poppins")
showtext_auto()

# Memphis Grizzlies palette (approx)
GRIZZ_NAVY <- "#12173F"
GRIZZ_BLUE <- "#5D76A9"
GRIZZ_TEAL <- "#00B2A9"
GRIZZ_GOLD <- "#F5B112"
GRIZZ_GRAY <- "#D8D8D8"

# Canvas style 
BG        <- "#f6f4ef"
TEXT_DARK <- "#141414"
GRID      <- "#dedad3"

theme_caleb_grizz <- function()
{
  theme_minimal(base_family = "poppins") +
    theme(
      plot.background  = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 18, face = "bold", hjust = 0.5, color = GRIZZ_NAVY),
      plot.subtitle    = element_text(size = 10, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 6)),
      axis.title       = element_text(size = 12, face = "bold", color = TEXT_DARK),
      axis.text        = element_text(size = 10, color = "#333333"),
      legend.position  = "top",
      legend.title     = element_text(size = 10, face = "bold", color = TEXT_DARK),
      legend.text      = element_text(size = 9, color = TEXT_DARK),
      plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 0, margin = margin(t = 6)),
      plot.margin      = margin(10, 12, 14, 12)
    )
}

to_num <- function(x) suppressWarnings(as.numeric(gsub("[^0-9\\.-]", "", x)))

kbl_caleb <- function(df, caption, font_size = 9, scale_down = TRUE, digits = NULL)
{
  if (is.null(digits)) {
    kb <- kableExtra::kbl(df, format = "latex", booktabs = TRUE, caption = caption)
  } else {
    kb <- kableExtra::kbl(df, format = "latex", booktabs = TRUE, caption = caption, digits = digits)
  }

  latex_options <- c("striped", "hold_position")
  if (isTRUE(scale_down)) latex_options <- c(latex_options, "scale_down")

  kb %>%
    kableExtra::kable_styling(latex_options = latex_options, font_size = font_size)
}

cache_dir <- params$cache_dir
export_dir <- params$export_dir
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)
if (!dir.exists(export_dir)) dir.create(export_dir, recursive = TRUE)
```

# Executive Summary

This report isolates a specific (and unusually frequent) Jaren Jackson Jr. pattern: his stocks exceeding his rebounds. 

For most bigs, "stocks > rebounds" is an outlier. For JJJ, it happens often enough to be an indicator of his defensive profile. He generates unique value through his rim protection and defensive event-creation, but gives up possession-ending value on the glass. 

This report aims to quantify that frequency, and analyze how his deficiencies as a rebounder compare to his peers. 

# Part 1. Stocks vs Rebounds

```{r pull-jjj}
player_id    <- params$player_id
player_name  <- params$player_name
seasons_jjj  <- params$seasons_jjj

get_season_logs <- function(season, season_type = c("Regular Season","Playoffs"), sleep_s = params$sleep_s) {
  season_type <- match.arg(season_type)

  f <- file.path(cache_dir, paste0("jjj_gamelogs_", player_id, "_", season, "_", gsub(" ", "", season_type), ".rds"))
  if (file.exists(f)) return(readRDS(f))

  Sys.sleep(sleep_s)
  out <- nba_playergamelogs(
    player_id   = player_id,
    season      = season,
    season_type = season_type
  )

  df <- out[sapply(out, is.data.frame)][[1]]
  df$SEASON_YEAR <- season
  df$SEASON_TYPE <- season_type

  saveRDS(df, f)
  df
}

# Regular season logs always
jjj_rs <- bind_rows(map(seasons_jjj, ~get_season_logs(.x, "Regular Season")))

# Add playoffs only if requested (used for stocks-only sections)
jjj_po <- if (isTRUE(params$include_playoffs_for_stocks)) {
  bind_rows(map(seasons_jjj, ~get_season_logs(.x, "Playoffs")))
} else {
  NULL
}

jjj_games_all <- bind_rows(jjj_rs, jjj_po) %>%
  mutate(
    game_date  = as.Date(GAME_DATE),
    MIN        = round(parse_number(MIN), 1),
    REB        = as.numeric(REB),
    STL        = as.numeric(STL),
    BLK        = as.numeric(BLK),
    stocks     = STL + BLK
  ) %>%
  filter(!is.na(game_date)) %>%
  arrange(game_date)

# Regular season only dataset for all rebounding bucket work
jjj_games_rs <- jjj_games_all %>% filter(SEASON_TYPE == "Regular Season")

meta_tbl <- tibble(
  Games_All = nrow(jjj_games_all),
  Games_RS  = nrow(jjj_games_rs),
  Seasons   = n_distinct(jjj_games_all$SEASON_YEAR),
  Types_All = paste(sort(unique(jjj_games_all$SEASON_TYPE)), collapse = ", ")
)
```

## Overall rate

```{r stocks-overall}
jjj_stocks_gt_reb <- jjj_games_all %>%
  filter(!is.na(stocks), !is.na(REB), stocks > REB) %>%
  transmute(
    Date = game_date,
    Season = SEASON_YEAR,
    Type = SEASON_TYPE,
    Matchup = MATCHUP,
    MIN, REB, STL, BLK,
    Stocks = stocks
  ) %>%
  arrange(Date)

overall_stocks <- tibble(
  Games_All = nrow(jjj_games_all),
  Games_Stocks_GT_REB = nrow(jjj_stocks_gt_reb),
  Rate = nrow(jjj_stocks_gt_reb) / nrow(jjj_games_all)
) %>%
  mutate(Rate = percent(Rate, accuracy = 0.1))

overall_stocks %>%
  kbl_caleb(
    caption = "Overall rate of games where Stocks > Rebounds (regular season + playoffs).",
    font_size = 9,
    scale_down = FALSE
  ) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, bold = TRUE)
```

## Season rates (Regular Season vs Playoffs)

```{r stocks-by-season}
stocks_by_season <- jjj_games_all %>%
  group_by(SEASON_YEAR, SEASON_TYPE) %>%
  summarise(
    Games = n(),
    Stocks_GT_REB = sum(stocks > REB, na.rm = TRUE),
    Rate = Stocks_GT_REB / Games,
    .groups = "drop"
  ) %>%
  arrange(SEASON_YEAR, SEASON_TYPE) %>%
  mutate(Rate = percent(Rate, accuracy = 0.1))

stocks_by_season %>%
  rename(Season = SEASON_YEAR, Type = SEASON_TYPE) %>%
  kbl_caleb(
    caption = "Stocks > Rebounds by season type (playoffs included only here).",
    font_size = 9,
    scale_down = FALSE
  ) %>%
  column_spec(1, bold = TRUE)
```

```{r plot-stocks-rate, fig.cap="Share of games where Stocks > Rebounds by season and season type."}
type_cols <- c("Regular Season" = GRIZZ_BLUE, "Playoffs" = GRIZZ_GOLD)

stocks_plot <- jjj_games_all %>%
  group_by(SEASON_YEAR, SEASON_TYPE) %>%
  summarise(
    Games = n(),
    Rate  = sum(stocks > REB, na.rm = TRUE) / Games,
    .groups = "drop"
  ) %>%
  mutate(SEASON_YEAR = factor(SEASON_YEAR, levels = params$seasons_jjj))

ggplot() +
  # Regular season: line + points
  geom_line(
    data = dplyr::filter(stocks_plot, SEASON_TYPE == "Regular Season"),
    aes(x = SEASON_YEAR, y = Rate, group = 1),
    linewidth = 1.2,
    color = type_cols["Regular Season"]
  ) +
  geom_point(
    data = dplyr::filter(stocks_plot, SEASON_TYPE == "Regular Season"),
    aes(x = SEASON_YEAR, y = Rate),
    size = 2.6,
    color = type_cols["Regular Season"]
  ) +
  # Playoffs: points only (no connecting line)
  geom_point(
    data = dplyr::filter(stocks_plot, SEASON_TYPE == "Playoffs"),
    aes(x = SEASON_YEAR, y = Rate),
    size = 2.8,
    color = type_cols["Playoffs"]
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Stocks > Rebounds Rate",
    subtitle = "Regular season shown as a trend line. Playoffs shown as points only.",
    x = NULL,
    y = "Rate",
    caption = "Data: NBA API via hoopR | Viz: @Rambzee_"
  ) +
  theme_caleb_grizz() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r stocks-tables, results='asis'}
season_summary <- jjj_stocks_gt_reb %>%
  group_by(Season) %>%
  summarise(
    Total = n(),
    Regular_Season = sum(Type == "Regular Season"),
    Playoffs = sum(Type == "Playoffs"),
    .groups = "drop"
  ) %>%
  arrange(Season)

season_summary_with_total <- bind_rows(
  season_summary,
  tibble(
    Season = "Career",
    Total = sum(season_summary$Total, na.rm = TRUE),
    Regular_Season = sum(season_summary$Regular_Season, na.rm = TRUE),
    Playoffs = sum(season_summary$Playoffs, na.rm = TRUE)
  )
)

kbl_caleb(
  season_summary_with_total,
  caption = "Count of games where Stocks > Rebounds by season (RS + playoffs).",
  font_size = 9,
  scale_down = FALSE
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:4, bold = TRUE)
```

```{r stocks-log, results='asis'}
n_show <- 10

stocks_log_show <- jjj_stocks_gt_reb %>%
  arrange(desc(Stocks), desc(REB), Date) %>%
  slice_head(n = n_show)

kbl_caleb(
  stocks_log_show,
  caption = paste0("Top ", n_show, " games where Stocks > Rebounds (sorted by stocks)."),
  font_size = 7,
  scale_down = TRUE
) %>%
  column_spec(1, bold = TRUE)
```

# Part 2. Rebounding buckets (regular season only)

Buckets:

\begin{itemize}
  \item REB < 5
  \item REB >= 5
  \item REB < 7
  \item REB >= 7
  \item REB < 3
\end{itemize}


## Overall bucket rates

```{r reb-buckets-overall, results='asis'}
jjj_reb_buckets_overall <- jjj_games_rs %>%
  filter(!is.na(REB)) %>%
  summarise(
    Games = n(),
    `P(REB < 5)` = mean(REB < 5),
    `P(REB >= 5)` = mean(REB >= 5),
    `P(REB < 7)` = mean(REB < 7),
    `P(REB >= 7)` = mean(REB >= 7),
    `P(REB <= 3)` = mean(REB <= 3)
  ) %>%
  mutate(across(starts_with("P("), ~percent(.x, accuracy = 0.1)))

kbl_caleb(
  jjj_reb_buckets_overall,
  caption = "JJJ rebounding bucket rates (regular season only).",
  font_size = 9,
  scale_down = FALSE
) %>%
  column_spec(1, bold = TRUE)
```

Across his career, Jackson’s rebounding profile is defined less by steadiness and more by sharp swings from game to game. In roughly 37% of his appearances, he finishes with fewer than five rebounds, and in nearly 67% he fails to reach seven. An unusually high frequency for a player of his size. Though, there are nights where the production looks more conventional. He clears five or more boards in about 63% of his games and reaches seven-plus roughly once every three outings. But those stretches are counterbalanced by a strikingly low floor. In about 1/4 of his career games, Jackson records three rebounds or fewer, highlighting how often his defensive value shows up through disruption rather than control of the glass.

## Bucket rates by season (regular season only)

```{r reb-buckets-by-season, results='asis'}
jjj_reb_buckets_by_season <- jjj_games_rs %>%
  filter(!is.na(REB)) %>%
  group_by(SEASON_YEAR) %>%
  summarise(
    Games = n(),
    `P(REB < 5)` = mean(REB < 5),
    `P(REB >= 5)` = mean(REB >= 5),
    `P(REB < 7)` = mean(REB < 7),
    `P(REB >= 7)` = mean(REB >= 7),
    `P(REB <= 3)` = mean(REB <= 3),
    .groups = "drop"
  ) %>%
  arrange(SEASON_YEAR) %>%
  mutate(across(starts_with("P("), ~percent(.x, accuracy = 0.1))) %>%
  rename(Season = SEASON_YEAR)

kbl_caleb(
  jjj_reb_buckets_by_season,
  caption = "JJJ rebounding buckets by season (regular season only).",
  font_size = 8,
  scale_down = FALSE
) %>%
  column_spec(1, bold = TRUE)
```

```{r plot-reb-buckets, fig.cap="JJJ rebounding bucket rates by season (regular season only)."}
jjj_reb_plot <- jjj_games_rs %>%
  filter(!is.na(REB)) %>%
  group_by(SEASON_YEAR) %>%
  summarise(
    REB_le_3 = mean(REB <= 3),
    REB_ge_5 = mean(REB >= 5),
    REB_ge_7 = mean(REB >= 7),
    .groups = "drop"
  ) %>%
  pivot_longer(-SEASON_YEAR, names_to = "Bucket", values_to = "Rate")

bucket_cols <- c(
  "REB ≥ 7" = GRIZZ_GOLD,
  "REB ≥ 5" = GRIZZ_TEAL,
  "REB ≤ 3" = GRIZZ_NAVY
)

jjj_reb_plot <- jjj_games_rs %>%
  filter(!is.na(REB)) %>%
  group_by(SEASON_YEAR) %>%
  summarise(
    REB_le_3 = mean(REB <= 3),
    REB_ge_5 = mean(REB >= 5),
    REB_ge_7 = mean(REB >= 7),
    .groups = "drop"
  ) %>%
  pivot_longer(-SEASON_YEAR, names_to = "Bucket", values_to = "Rate") %>%
  mutate(
    Bucket = recode(
      Bucket,
      REB_le_3 = "REB ≤ 3",
      REB_ge_5 = "REB ≥ 5",
      REB_ge_7 = "REB ≥ 7"
    )
  )

ggplot(jjj_reb_plot, aes(SEASON_YEAR, Rate, color = Bucket, group = Bucket)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.6) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = bucket_cols) +
  labs(
    title = "JJJ Rebounding Buckets by Season",
    subtitle = "Regular season only",
    x = NULL,
    y = "Share of games",
    color = NULL,
    caption = "Data: NBA API via hoopR | Viz: @Rambzee_"
  ) +
  theme_caleb_grizz() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
Season by season, JJJ has made real strides on the glass relative to his early career, with a clear peak during his 2022–23 DPOY season. That year stands out as a positive outlier, when he cleared five rebounds in over 80% of games and seven-plus in more than half. 

Since then, his rebounding has regressed and stabilized at a lower level. Over the past three seasons, he has typically reached five rebounds in roughly 55–65% of games, while seven-plus rebound performances have hovered closer to 25–35%. More concerning is the floor, low-end numbers remain common, with roughly one-fifth to one-quarter of recent games ending with three rebounds or fewer, a disconcerting frequency for a 6'10 big.

# Part 3. Peer context: REB/36 vs 6'8\"+ peers

```{r load-peers}
seasons_peer <- params$seasons_peer
peer_files <- params$peer_files
stopifnot(length(seasons_peer) == length(peer_files))

read_peer_file <- function(season, file) {
  df <- readxl::read_excel(file)
  df %>% mutate(SEASON_YEAR = season)
}

peers_all <- map2_dfr(seasons_peer, peer_files, read_peer_file)
```

```{r standardize-peers}
peers_std <- peers_all %>% rename_with(~toupper(.x))

reb36_candidates <- c("REB_PER36","REB_PER_36","REB36","REB/36")
reb36_col <- reb36_candidates[reb36_candidates %in% names(peers_std)][1]
if (is.na(reb36_col)) stop("No REB/36 column found. Columns: ", paste(names(peers_std), collapse = ", "))

name_candidates <- c("PLAYER_NAME","PLAYER")
name_col <- name_candidates[name_candidates %in% names(peers_std)][1]
if (is.na(name_col)) stop("No player name column found. Columns: ", paste(names(peers_std), collapse = ", "))

team_col <- if ("TEAM" %in% names(peers_std)) "TEAM" else if ("TEAM_ABBREVIATION" %in% names(peers_std)) "TEAM_ABBREVIATION" else NA_character_
gp_col   <- if ("GP" %in% names(peers_std)) "GP" else NA_character_
min_col  <- if ("MIN" %in% names(peers_std)) "MIN" else NA_character_

plot_df <- peers_std %>%
  transmute(
    SEASON_YEAR = as.character(SEASON_YEAR),
    PLAYER_NAME = as.character(.data[[name_col]]),
    TEAM = if (!is.na(team_col)) as.character(.data[[team_col]]) else NA_character_,
    GP = if (!is.na(gp_col)) as.integer(.data[[gp_col]]) else NA_integer_,
    MIN = if (!is.na(min_col)) as.numeric(.data[[min_col]]) else NA_real_,
    REB_PER36 = as.numeric(.data[[reb36_col]])
  ) %>%
  filter(!is.na(PLAYER_NAME), !is.na(REB_PER36)) %>%
  mutate(SEASON_YEAR = factor(SEASON_YEAR, levels = seasons_peer))

# Optional filters (xlsx usually already filtered)
if (any(!is.na(plot_df$MIN))) plot_df <- plot_df %>% filter(is.na(MIN) | MIN >= params$min_minutes_peer)
if (any(!is.na(plot_df$GP)))  plot_df <- plot_df %>% filter(is.na(GP)  | GP  >= params$min_games_peer)

jjj_peer_df <- plot_df %>% filter(str_detect(tolower(PLAYER_NAME), "jaren jackson"))

peer_meta <- tibble(
  Seasons = n_distinct(plot_df$SEASON_YEAR),
  Players = n_distinct(plot_df$PLAYER_NAME),
  JJJ_Rows = nrow(jjj_peer_df)
)
```

```{r plot-peers, fig.cap="REB/36 distribution for 6'8\"+ peers by season with JJJ highlighted."}
ggplot(plot_df, aes(x = SEASON_YEAR, y = REB_PER36)) +
  geom_jitter(width = 0.18, alpha = 0.22, color = GRIZZ_BLUE) +
  geom_point(data = jjj_peer_df, size = 3.2, color = GRIZZ_NAVY) +
  geom_line(data = jjj_peer_df, aes(group = 1), linewidth = 1.2, color = GRIZZ_NAVY) +
  labs(
    title = "REB/36 vs 6'8\"+ Peers",
    subtitle = "Each season distribution with JJJ highlighted and connected.",
    x = NULL,
    y = "REB per 36",
    caption = "Data: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb_grizz()
```

```{r peer-percentiles, results='asis'}
peer_pct <- plot_df %>%
  group_by(SEASON_YEAR) %>%
  mutate(pct = percent_rank(REB_PER36)) %>%
  ungroup()

jjj_pct <- peer_pct %>%
  filter(str_detect(tolower(PLAYER_NAME), "jaren jackson")) %>%
  group_by(SEASON_YEAR) %>%
  summarise(
    REB_per36 = first(REB_PER36),
    Percentile = first(pct),
    Peer_N = n_distinct(peer_pct$PLAYER_NAME[peer_pct$SEASON_YEAR == first(SEASON_YEAR)]),
    .groups = "drop"
  ) %>%
  arrange(SEASON_YEAR) %>%
  mutate(Percentile = percent(Percentile, accuracy = 1))

kbl_caleb(
  jjj_pct,
  caption = "JJJ percentile rank in REB/36 among 6'8\"+ peers by season.",
  font_size = 9,
  scale_down = FALSE
) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, bold = TRUE)
```

```{r peer-hists, fig.cap="Peer REB/36 distributions by season (JJJ marked with a vertical line)."}
plot_hist_season <- function(season) {
  df_s <- plot_df %>% filter(SEASON_YEAR == season)
  j_s  <- jjj_peer_df %>% filter(SEASON_YEAR == season)

  # If JJJ missing (injury / mins filter), skip gracefully
  if (nrow(j_s) == 0) return(NULL)

  ggplot(df_s, aes(x = REB_PER36)) +
    geom_histogram(bins = 30, fill = GRIZZ_BLUE, alpha = 0.35, color = NA) +
    geom_vline(data = j_s, aes(xintercept = REB_PER36), color = GRIZZ_NAVY, linewidth = 1.1) +
    labs(
      title = paste0("REB/36 Distribution (", as.character(season), ")"),
      subtitle = "Vertical line = JJJ | Among 6'8\"+ peers",
      x = "REB per 36",
      y = "Count"
    ) +
    theme_caleb_grizz()
}

# Print one histogram per season (only seasons where JJJ appears in peer file)
walk(levels(plot_df$SEASON_YEAR), ~print(plot_hist_season(.x)))
```
When placed alongside his peers (players 6'8"+ with at least 900 minutes in a season) JJJ’s standing as a rebounder becomes clear. His highest mark came in 2022–23, his DPOY season, but even then his rebounding profile graded out as mediocre to lower-tier relative to similarly sized players. Since then, he has consistently fallen near the bottom of the distribution, ranking in roughly the bottom fifth among his peers, including around the 20th percentile this season.

## Takeaways
* Stocks vs rebounds over time: JJJ’s tendency to generate more stocks than rebounds has softened slightly in recent seasons, but it remains a defining feature of his game. The pattern is persistent enough to be descriptive, not anecdotal, and reflects how his defensive impact most often shows up.

* Playoff context: While playoffs were excluded from the core rebounding analysis due to small sample size, they reinforce the identity. Even with a higher average (6.5 RPG across 27 playoff games), the same volatility appears, with nearly half of those games ending at five rebounds or fewer.

* Defensive role link: Jackson’s elite stocks production aligns with a help-side, contest-first defensive role, where rim deterrence and disruption take priority over consistent box-outs and possession-ending rebounds.

* Rebounding buckets (regular season): His rebounding clearly peaked in 2022–23, his DPOY season, when he cleared five rebounds in roughly 70% of games. Since then, his profile has stabilized at a lower level, with seven-plus rebound games hovering around ~20% and a concerning share of nights falling at the very bottom of the distribution. The frequency of REB ≤ 3 games (around 10–25% depending on season) matters because it shows his non-involvement on the glass for a 6'10 big, not just below-average output.

* REB/36 vs peers: Relative to other 6'8"+ players with meaningful minutes, Jackson consistently grades out below average as a rebounder. Even in his 2022–23 peak, he sat in the lower half of the distribution, before slipping into roughly the bottom 20% in subsequent seasons. The year-by-year histograms show this isn’t driven by a single outlier season, he regularly occupies the lower tail rather than falling around the league median.

* Contextual factors: Environment and role also play a part. Jackson has often played alongside strong rebounders like Steven Adams and Zach Edey, allowing him to roam as a helper rather than function as a box-out anchor. Early in his career, his responsibilities were different, but even as his role has expanded, the fundamental trade-off remains: elite defensive event generation paired with consistently weak rebounding for his size.