---
title: "Utah’s Defensive Bet: Jaren Jackson Jr. and the Jazz Core"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{ragged2e}
params:
  # --- Focus ---
  season_focus: "2025-26"
  team_id_jazz: 1610612762

  # --- Core players ---
  player_id_jjj: 1628991
  player_name_jjj: "Jaren Jackson Jr."
  player_id_kessler: 1631117
  player_name_kessler: "Walker Kessler"
  player_id_keyonte: 1641730
  player_name_keyonte: "Keyonte George"
  player_id_lauri: 1628374
  player_name_lauri: "Lauri Markkanen"

  include_keyonte_appendix: true

  # --- Appendix: stocks ---
  seasons_jjj: ["2018-19","2019-20","2020-21","2021-22","2022-23","2023-24","2024-25","2025-26"]
  include_playoffs_for_stocks: true

  # --- Peer file inputs (REB/36) ---
  seasons_peer: ["2021-22","2022-23","2023-24","2024-25","2025-26"]
  peer_files:
    - "RebPer36_GreaterThan68_2021_22.xlsx"
    - "RebPer36_GreaterThan68_2022_23.xlsx"
    - "RebPer36_GreaterThan68_2023_24.xlsx"
    - "RebPer36_GreaterThan68_2024_25.xlsx"
    - "RebPer36_GreaterThan68_2025_26.xlsx"
  min_minutes_peer: 900
  min_games_peer: 20

  # --- Caching / exports ---
  sleep_s: 0.7
  cache_dir: "cache_hoopr"
  export_dir: "exports"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  dpi = 300,
  out.width = "100%"
)

library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(readr)
library(janitor)
library(ggplot2)
library(scales)
library(forcats)
library(knitr)
library(kableExtra)
library(readxl)
library(hoopR)
library(glue)
library(ggrepel)

# Fonts
library(showtext)
sysfonts::font_add_google("Poppins", "poppins")
showtext_auto()

# Utah palette
UTAH_PURPLE <- "#5A2A82"
UTAH_BLACK  <- "#0B0B0F"
UTAH_SKY    <- "#6EC1E4"
UTAH_STONE  <- "#D9D6CF"

# Canvas style
BG        <- "#f6f4ef"
TEXT_DARK <- "#141414"
GRID      <- "#dedad3"

theme_caleb_jazz <- function() {
  theme_minimal(base_family = "poppins") +
    theme(
      plot.background  = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 18, face = "bold", hjust = 0.5, color = UTAH_BLACK),
      plot.subtitle    = element_text(size = 10, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 6)),
      axis.title       = element_text(size = 12, face = "bold", color = TEXT_DARK),
      axis.text        = element_text(size = 10, color = "#333333"),
      legend.position  = "top",
      legend.title     = element_text(size = 10, face = "bold", color = TEXT_DARK),
      legend.text      = element_text(size = 9, color = TEXT_DARK),
      plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 0, margin = margin(t = 6)),
      plot.margin      = margin(10, 12, 14, 12)
    )
}

kbl_caleb <- function(df, caption, font_size = 9, scale_down = TRUE, digits = NULL) {
  kb <- if (is.null(digits)) {
    kableExtra::kbl(df, format = "latex", booktabs = TRUE, caption = caption)
  } else {
    kableExtra::kbl(df, format = "latex", booktabs = TRUE, caption = caption, digits = digits)
  }

  latex_options <- c("striped", "hold_position")
  if (isTRUE(scale_down)) latex_options <- c(latex_options, "scale_down")

  kb %>%
    kableExtra::kable_styling(
      latex_options = latex_options,
      font_size = font_size
    )
}

cache_dir <- params$cache_dir
export_dir <- params$export_dir
if (!dir.exists(cache_dir)) dir.create(cache_dir, recursive = TRUE)
if (!dir.exists(export_dir)) dir.create(export_dir, recursive = TRUE)

to_num <- function(x) readr::parse_number(as.character(x))
```

# Executive Summary

Jaren Jackson Jr.’s acquisition marks a clear pivot for the Utah Jazz toward a defined two-way identity as they position themselves to re-enter the playoff hunt, even as the short-term priority remains preserving their top-eight protected first-round pick. He brings elite rim protection, defensive event creation, and lineup versatility, while also slotting in as a reliable third option who can space the floor and score without heavy usage beside Keyonte George and Lauri Markkanen. At his best, Jackson profiles as an All-Star caliber player whose impact is felt on both ends, combining high-level interior defense with a complementary offensive package.

# 1. Trade Context and Timeline Alignment

Memphis has committed to a rebuild and reset its timeline. Utah has done the opposite, consolidating assets to place a defensive backbone behind a young, offense-driven roster. By establishing a credible defensive frontcourt, the Jazz can preserve their explosive offensive identity while becoming more competitive and better positioned to capitalize on recent draft success, smart trades, and emerging star talent.

# 2. How JJJ Fits the Jazz Offense

Offensively, Jaren Jackson Jr. fits best as a connector. He can score efficiently without dominating the ball, using a combination of spacing, screening, and touch in the paint to keep possessions moving. His development as a post scorer, with a reliable floater and strong finishing, allows Utah to punish switches and late help without tilting the offense away from its primary creators.

## 2.1 Spacing and Shooting Gravity

```{r jjj-offense-context, results='asis'}
# Pull base and advanced dashboards for a stable top-line offensive snapshot.
dash_base <- hoopR::nba_leaguedashplayerstats(
  season = params$season_focus,
  season_type = "Regular Season",
  measure_type = "Base"
)$LeagueDashPlayerStats %>%
  janitor::clean_names()

dash_adv <- hoopR::nba_leaguedashplayerstats(
  season = params$season_focus,
  season_type = "Regular Season",
  measure_type = "Advanced"
)$LeagueDashPlayerStats %>%
  janitor::clean_names()

jjj_base <- dash_base %>% filter(player_name == params$player_name_jjj)
jjj_adv  <- dash_adv  %>% filter(player_name == params$player_name_jjj)

jjj_off_tbl <- jjj_base %>%
  transmute(
    Season = params$season_focus,
    GP = as.numeric(gp),
    MPG = as.numeric(min) / GP,
    PPG = as.numeric(pts) / GP,
    `3PA/g` = as.numeric(fg3a) / GP,
    `3P%` = as.numeric(fg3_pct),
    `TS%` = as.numeric(pts) / (2 * (as.numeric(fga) + (0.44 * as.numeric(fta)))),
    `FTA/g` = as.numeric(fta) / GP
  ) %>%
  mutate(
    `3P%` = percent(`3P%`, 0.1),
    `FTA/g` = round(`FTA/g`, 2),
    MPG = round(MPG, 1),
    PPG = round(PPG, 1),
    `3PA/g` = round(`3PA/g`, 1),
    `TS%` = percent(`TS%`, 0.1)
  )

jjj_adv_tbl <- jjj_adv %>%
  transmute(
    Season = params$season_focus,
    `USG%` = as.numeric(usg_pct),
    `AST%` = as.numeric(ast_pct),
    `AST:TO` = as.numeric(ast_to)
  ) %>%
  mutate(
    `USG%` = percent(`USG%`, 0.1),
    `AST%` = percent(`AST%`, 0.1),
    `AST:TO` = round(`AST:TO`, 1)
  )

kbl_caleb(jjj_off_tbl, "JJJ offensive context", font_size = 9, scale_down = FALSE)
kbl_caleb(jjj_adv_tbl, "JJJ advanced profile", font_size = 9, scale_down = FALSE)
```

This season, Jackson is averaging 19 points per game on roughly league-average efficiency, offering real spacing with five threes per game at 36% while getting to the line at a solid rate. His 25% usage is slightly high, but largely reflects his role as a primary option in an unstable Memphis offense. Passing remains his biggest limitation: a 0.9 AST-to-TO ratio and roughly 10% assist rate show limited playmaking beyond basic reads. Utah’s spacing, guard creation, and Will Hardy’s structure should allow Jackson to maximize his scoring impact without placing undue decision-making demands on him.

## 2.2 Shot Diet: Rim vs. Paint vs. 3 vs. Midrange

```{r jjj-shotmix, results='asis'}
shots_jjj <- hoopR::nba_shotchartdetail(
  season = params$season_focus,
  season_type = "Regular Season",
  player_id = params$player_id_jjj,
  context_measure_simple = "FGA"
)$Shot_Chart_Detail %>%
  janitor::clean_names()

shot_mix <- shots_jjj %>%
  mutate(
    zone = case_when(
      shot_zone_basic == "Restricted Area" ~ "Rim",
      shot_zone_basic == "Mid-Range" ~ "Mid-range",
      str_detect(shot_zone_basic, "3") ~ "3PT",
      TRUE ~ "Other"
    )
  ) %>%
  count(zone) %>%
  mutate(pct = n / sum(n)) %>%
  arrange(desc(pct)) %>%
  mutate(pct = percent(pct, 0.1))

kbl_caleb(shot_mix, "JJJ shot mix by zone (ShotChartDetail)", font_size = 9, scale_down = FALSE)
```

```{r jjj-shotmix-plot, fig.cap="JJJ shot mix by zone (season focus)."}
shot_mix_plot <- shots_jjj %>%
  mutate(
    zone = case_when(
      shot_zone_basic == "Restricted Area" ~ "Rim",
      shot_zone_basic == "Mid-Range" ~ "Mid-range",
      shot_zone_basic == "In The Paint (Non-RA)" ~ "Paint (Non-RA)",
      str_detect(shot_zone_basic, "3") ~ "3PT",
      TRUE ~ "Other"
    )
  ) %>%
  count(zone) %>%
  mutate(
    pct = n / sum(n),
    zone = fct_reorder(zone, pct)
  )

ggplot(shot_mix_plot, aes(zone, pct)) +
  geom_col(fill = UTAH_PURPLE, alpha = 0.90) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(1)) +
  labs(
    title = "Shot Profile Fit",
    subtitle = glue("JJJ shot mix by zone ({params$season_focus})"),
    x = NULL,
    y = "% of FGA",
    caption = "Data: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb_jazz()
```

Jaren Jackson Jr. does a good job of maximizing his strengths, focusing his scoring on rim and paint finishes and three-point shooting, with fewer than 10% of his attempts coming from inefficient midrange shots.

## 2.3 Slotting into the Offensive Hierarchy

Utah’s offense will be most effective with clearly defined roles: Keyonte George running the offense, Lauri Markkanen commanding defensive attention as a high-level scorer, and Jaren Jackson Jr. functioning as a secondary option while still holding up in isolation when needed. That structure keeps Jackson in space, limits inefficient possessions, and allows the offense to stay organized while supporting the development of the rest of the young core.

# 3. How JJJ Fits the Jazz Defense

Utah is focused on building a defense it can trust every night. With Jaren Jackson Jr. alongside Walker Kessler, the Jazz immediately form one of the league’s strongest defensive frontcourts. The perimeter defense remains a weakness, but elite rim protection raises the floor and keeps the unit from collapsing, giving players like Lauri and the young guards room to hopefully be passable defensively.

## 3.1 Rim Protection: DFG% vs. Expected (<6 ft), Last 5 Seasons

```{r rim-protection-lastN, results='asis'}
safe_shotdef <- function(season, player_id) {
  tryCatch(
    hoopR::nba_playerdashptshotdefend(
      season = season,
      season_type = "Regular Season",
      player_id = player_id
    ),
    error = function(e) NULL
  )
}

extract_defending_shots <- function(sd_obj) {
  if (is.null(sd_obj)) return(NULL)

  if ("DefendingShots" %in% names(sd_obj)) {
    tbl <- sd_obj[["DefendingShots"]]
    if (is.data.frame(tbl)) return(tbl)
  }

  tbls <- sd_obj[vapply(sd_obj, is.data.frame, logical(1))]
  if (length(tbls) == 0) return(NULL)
  tbls[[1]]
}

clean_shotdef_tbl <- function(sd_obj, player_label, season_label) {
  tbl <- extract_defending_shots(sd_obj)
  if (is.null(tbl) || !is.data.frame(tbl) || nrow(tbl) == 0) return(NULL)

  out <- tbl %>%
    janitor::clean_names()

  need <- c("defense_category", "d_fga", "d_fgm", "normal_fg_pct", "d_fg_pct", "pct_plusminus")
  if (!all(need %in% names(out))) return(NULL)

  out %>%
    mutate(
      player = player_label,
      season = season_label,
      across(c(gp, g, d_fgm, d_fga), ~ suppressWarnings(as.integer(.x))),
      across(c(freq, d_fg_pct, normal_fg_pct, pct_plusminus), ~ suppressWarnings(as.numeric(.x)))
    )
}

last_n_seasons <- function(season_focus, n = 5) {
  y0 <- as.integer(stringr::str_sub(season_focus, 1, 4))
  yrs <- (y0 - (n - 1)):y0
  paste0(yrs, "-", stringr::str_sub(yrs + 1, 3, 4))
}

seasons_rim <- last_n_seasons(params$season_focus, n = 5)

pull_rim_series <- function(player_id, player_label) {
  purrr::map_dfr(
    seasons_rim,
    function(s) {
      sd <- safe_shotdef(s, player_id)
      clean_shotdef_tbl(sd, player_label, s)
    }
  )
}

rim_jjj <- pull_rim_series(params$player_id_jjj, "JJJ")
rim_kes <- pull_rim_series(params$player_id_kessler, "Kessler")

rim_all <- dplyr::bind_rows(rim_jjj, rim_kes) %>%
  filter(defense_category == "Less Than 6 Ft") %>%
  filter(!is.na(d_fga), !is.na(d_fgm)) %>%
  mutate(
    d_fga = as.numeric(d_fga),
    d_fgm = as.numeric(d_fgm)
  ) %>%
  filter(d_fga > 0)

if (nrow(rim_all) == 0) {

  kbl_caleb(
    tibble::tibble(Note = "No <6 ft shot defense rows returned from nba_playerdashptshotdefend for this season window."),
    caption = glue::glue("Rim Protection: DFG% vs. Expected - last 5 seasons"),
    font_size = 9,
    scale_down = FALSE
  )

} else {

  rim_agg <- rim_all %>%
    group_by(player) %>%
    summarise(
      Seasons = paste(sort(unique(season)), collapse = ", "),
      DFA = sum(d_fga, na.rm = TRUE),
      DFM = sum(d_fgm, na.rm = TRUE),
      DFG = DFM / DFA,
      EXP = weighted.mean(normal_fg_pct, w = d_fga, na.rm = TRUE),
      DIFF = DFG - EXP,
      .groups = "drop"
    ) %>%
    mutate(
      DFA = as.integer(round(DFA)),
      DFM = as.integer(round(DFM)),
      `DFG%` = scales::percent(DFG, accuracy = 0.1),
      `Exp%` = scales::percent(EXP, accuracy = 0.1),
      `Diff` = scales::percent(DIFF, accuracy = 0.1)
    ) %>%
    select(Player = player, Seasons, DFA, DFM, `DFG%`, `Exp%`, `Diff`) %>%
    arrange(match(Player, c("JJJ", "Kessler")))

  kbl_caleb(
    rim_agg,
    caption = glue::glue("Rim Protection: DFG% vs. Expected - last 5 seasons"),
    font_size = 9,
    scale_down = FALSE
  ) %>%
    kableExtra::column_spec(1, bold = TRUE) %>%
    kableExtra::column_spec(5:7, bold = TRUE)

}
```

```{r rim-top5-last5, results='asis'}
rim_top5 <- tibble::tribble(
  ~Player, ~`Shots Defended (<6ft)`, ~`DFG% (<6ft)`,
  "Kristaps Porzingis", 1630, 0.522,
  "Jaren Jackson Jr.", 1810, 0.523,
  "Walker Kessler", 1380, 0.523,
  "Rudy Gobert", 2374, 0.524,
  "Giannis Antetokounmpo", 1387, 0.528
) %>%
  mutate(
    `DFG% (<6ft)` = scales::percent(`DFG% (<6ft)`, accuracy = 0.1)
  )

kbl_caleb(
  rim_top5,
  caption = glue::glue("Rim DFG% Leaders, last 5 seasons (min 1350 shots defended)"),
  font_size = 9,
  scale_down = FALSE
) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::column_spec(2, bold = TRUE) %>%
  kableExtra::column_spec(3, bold = TRUE)
```

Since entering the league, both Jaren Jackson Jr. and Walker Kessler have established themselves as elite rim protectors. Over the past five seasons, they rank second and third in rim DFG%, even with Kessler missing significant time and only playing parts of three seasons in that span. Both also sit near the top of the league in rim field-goal percentage differential. Pairings like this are rare. While there have been strong interior duos, even recently with players like Mobley and Allen, it is uncommon to see two defenders who can independently reshape a team’s defense sharing the same frontcourt.

## 3.2 Defensive Event Creation

Rim deterrence is only part of the defensive equation. Event creation is the other. Jaren Jackson Jr.’s blocks and steals do more than pad the stat sheet; they force offenses into worse shots and create extra possessions. That disruption builds margin for error, giving Utah a chance to compete even on nights when the offense struggles.

```{=latex}
\par\justifying
```

# 4. Support Areas

## 4.1 Rebounding: REB/36 Peer Context (JJJ vs. Kessler)

```{r load-peers}
seasons_peer <- params$seasons_peer
peer_files <- params$peer_files
stopifnot(length(seasons_peer) == length(peer_files))

read_peer_file <- function(season, file) {
  df <- readxl::read_excel(file)
  df %>% mutate(SEASON_YEAR = season)
}

peers_all <- map2_dfr(seasons_peer, peer_files, read_peer_file)
```

```{r standardize-peers}
peers_std <- peers_all %>% rename_with(~toupper(.x))

reb36_candidates <- c("REB_PER36", "REB_PER_36", "REB36", "REB/36")
reb36_col <- reb36_candidates[reb36_candidates %in% names(peers_std)][1]
if (is.na(reb36_col)) stop("No REB/36 column found. Columns: ", paste(names(peers_std), collapse = ", "))

name_candidates <- c("PLAYER_NAME", "PLAYER")
name_col <- name_candidates[name_candidates %in% names(peers_std)][1]
if (is.na(name_col)) stop("No player name column found. Columns: ", paste(names(peers_std), collapse = ", "))

team_col <- if ("TEAM" %in% names(peers_std)) "TEAM" else if ("TEAM_ABBREVIATION" %in% names(peers_std)) "TEAM_ABBREVIATION" else NA_character_
gp_col <- if ("GP" %in% names(peers_std)) "GP" else NA_character_
min_col <- if ("MIN" %in% names(peers_std)) "MIN" else NA_character_

plot_df <- peers_std %>%
  transmute(
    SEASON_YEAR = as.character(SEASON_YEAR),
    PLAYER_NAME = as.character(.data[[name_col]]),
    TEAM = if (!is.na(team_col)) as.character(.data[[team_col]]) else NA_character_,
    GP = if (!is.na(gp_col)) as.integer(.data[[gp_col]]) else NA_integer_,
    MIN = if (!is.na(min_col)) as.numeric(.data[[min_col]]) else NA_real_,
    REB_PER36 = as.numeric(.data[[reb36_col]])
  ) %>%
  filter(!is.na(PLAYER_NAME), !is.na(REB_PER36)) %>%
  mutate(SEASON_YEAR = factor(SEASON_YEAR, levels = seasons_peer))

if (any(!is.na(plot_df$MIN))) plot_df <- plot_df %>% filter(is.na(MIN) | MIN >= params$min_minutes_peer)
if (any(!is.na(plot_df$GP)))  plot_df <- plot_df %>% filter(is.na(GP)  | GP  >= params$min_games_peer)

add_tag <- function(df, pattern, label) {
  df %>%
    mutate(
      tag = dplyr::case_when(
        !is.na(tag) ~ tag,
        stringr::str_detect(stringr::str_to_lower(PLAYER_NAME), pattern) ~ label,
        TRUE ~ NA_character_
      )
    )
}

tagged <- plot_df %>%
  mutate(tag = NA_character_) %>%
  add_tag("jaren jackson", "JJJ") %>%
  add_tag("walker kessler", "Kessler") %>%
  mutate(
    tag = dplyr::if_else(is.na(tag), "Peer", tag),
    tag = factor(tag, levels = c("Peer", "JJJ", "Kessler"))
  )
```

```{r plot-reb36-peers, fig.cap="REB/36 distribution for 6'8+ peers by season with JJJ and Kessler highlighted."}
tag_cols <- c(
  "Peer" = UTAH_SKY,
  "JJJ" = UTAH_PURPLE,
  "Kessler" = UTAH_BLACK
)

ggplot(tagged, aes(x = SEASON_YEAR, y = REB_PER36)) +
  geom_jitter(
    data = tagged %>% filter(tag == "Peer"),
    width = 0.18,
    alpha = 0.3,
    color = UTAH_SKY
  ) +
  geom_line(
    data = tagged %>% filter(tag != "Peer"),
    aes(group = tag, color = tag),
    linewidth = 1.10,
    alpha = 0.95
  ) +
  geom_point(
    data = tagged %>% filter(tag != "Peer"),
    aes(color = tag),
    size = 3.2
  ) +
  geom_text(
    data = tagged %>% filter(tag != "Peer"),
    aes(label = tag, color = tag),
    nudge_y = 0.5,
    size = 2,
    show.legend = FALSE
  ) +
  scale_color_manual(values = tag_cols) +
  labs(
    title = "Rebounding Context (REB/36)",
    subtitle = "Peer (6'8+) distribution by season | Kessler and JJJ are complementary",
    x = NULL,
    y = "REB per 36",
    caption = "Data: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb_jazz() +
  theme(legend.position = "none")
```

This plot is meant to keep the rebounding conversation in perspective. Jaren Jackson Jr. is not being added to solve rebounding, but to control possessions defensively. While he ranks among the weaker rebounders in the league for his size, Walker Kessler is among the best when he is on the floor, giving Utah a natural counterbalance that allows the team to absorb the trade-off through scheme and collective effort.

# 5. What This Means for Utah’s Future

## 5.1 Development Accelerant

If Keyonte George is going to be the long-term offensive engine, Utah needs a defense that absorbs mistakes. The goal is not to hide him, but to let him play aggressively without every breakdown turning into points at the rim. A JJJ-Kessler back line lowers the cost of growing pains and gives young guards room to develop without the defense collapsing into a bottom-five unit.

```{r keyonte-appendix, eval=params$include_keyonte_appendix, results='asis'}
kg_base <- dash_base %>% filter(player_name == params$player_name_keyonte)

if (nrow(kg_base) > 0) {
  kg_tbl <- kg_base %>%
    transmute(
      Season = params$season_focus,
      GP = as.numeric(gp),
      MPG = as.numeric(min) / GP,
      PPG = as.numeric(pts) / GP,
      `3PA/g` = as.numeric(fg3a) / GP,
      `3P%` = as.numeric(fg3_pct),
      `TS%` = as.numeric(pts) / (2 * (as.numeric(fga) + (0.44 * as.numeric(fta)))),
      `FTA/g` = as.numeric(fta) / GP,
      AST = as.numeric(ast) / GP,
      TOV = as.numeric(tov) / GP
    ) %>%
    mutate(
      `TS%` = percent(`TS%`, 0.1),
      `3P%` = percent(`3P%`, 0.1),
      MPG = round(MPG, 1),
      PPG = round(PPG, 1),
      AST = round(AST, 1),
      TOV = round(TOV, 1),
      `3PA/g` = round(`3PA/g`, 1),
      `FTA/g` = round(`FTA/g`, 1)
    )

  kbl_caleb(kg_tbl, "Keyonte George: quick season snapshot", font_size = 9, scale_down = FALSE)
}
```

## 5.2 Lineup Scalability

The upside is flexibility. Utah can lean into size with two rim protectors, downsize with JJJ at the five to add shooting and speed, and still keep the defense intact. That versatility gives the Jazz multiple ways to stay competitive as the roster develops, rather than being locked into a single lineup or play style.

# Final Summary

This trade makes sense for Utah. Control the paint, force tougher shots, and let the roster handle the trade-offs. If the offense holds around its creators, a JJJ-Kessler frontcourt gives the Jazz a real defensive floor and a path to playoff contention as early as next season, with a young Jaren Jackson Jr. who fits both the timeline and the cap.
