---
title: "Rebound After Slow Starts (2025–26)"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
params:
  season: "2025-26"
  season_type: "Regular Season"
  players:
    - "Cam Spencer"
    - "Cam Johnson"
    - "Miles McBride"
    - "Derrick White"
    - "Tyus Jones"
    - "Isaiah Stewart"
    - "Jay Huff"
    - "Anthony Black"
    - "Shaedon Sharpe"
---

```{r setup, include=FALSE}
# ------------------------------------------------------------
# Packages
# ------------------------------------------------------------

library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(lubridate)
library(ggplot2)
library(scales)
library(ggrepel)
library(knitr)
library(showtext)
library(sysfonts)
library(hoopR)

# ------------------------------------------------------------
# Knit options
# ------------------------------------------------------------

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  dpi = 300,
  out.width = "100%",
  fig.width = 7.6,
  fig.height = 4.9
)

# ------------------------------------------------------------
# Fonts
# ------------------------------------------------------------

sysfonts::font_add_google("Poppins", "poppins")
showtext::showtext_auto()

# ------------------------------------------------------------
# Constants
# ------------------------------------------------------------

SEASON <- params$season
SEASON_TYPE <- params$season_type
PLAYERS <- params$players

SHOT_CACHE_DIR <- "rebound_after_slow_starts/data/processed/shots"

BG <- "#f6f4ef"
TEXT_DARK <- "#141414"
GRID <- "#dedad3"

TEAM_COLORS <- c(
  BOS = "#007A33",
  DET = "#C8102E",
  IND = "#002D62",
  MEM = "#5D76A9",
  NYK = "#006BB6",
  ORL = "#0077C0"
)

# ------------------------------------------------------------
# Theme
# ------------------------------------------------------------

theme_caleb <- theme_minimal(base_family = "poppins") +
  theme(
    plot.background = element_rect(fill = BG, color = NA),
    panel.background = element_rect(fill = BG, color = NA),
    panel.grid.major = element_line(color = GRID, linewidth = 0.30),
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = TEXT_DARK),
    plot.subtitle = element_text(
      size = 12,
      hjust = 0.5,
      color = "#2f2f2f",
      margin = margin(b = 10)
    ),
    axis.title = element_text(size = 13, face = "bold", color = TEXT_DARK),
    axis.text = element_text(size = 10, color = "#333333"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    plot.caption = element_text(
      size = 8,
      color = "#6e6e6e",
      hjust = 1,
      margin = margin(t = 8)
    ),
    plot.margin = margin(10, 14, 10, 14)
  )

theme_zone <- theme_minimal(base_family = "poppins") +
  theme(
    plot.background = element_rect(fill = BG, color = NA),
    panel.background = element_rect(fill = BG, color = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

# ------------------------------------------------------------
# General helpers
# ------------------------------------------------------------

to_num <- function(x) suppressWarnings(as.numeric(x))

parse_nba_date <- function(x) {
  suppressWarnings(
    lubridate::parse_date_time(
      x,
      orders = c("Ymd", "ymd", "mdY", "b d, Y", "B d, Y", "m/d/Y")
    )
  ) %>%
    as.Date()
}

fmt_delta <- function(x, digits = 1) {
  out <- round(x, digits)
  ifelse(out > 0, paste0("+", out), as.character(out))
}

fmt_pct_delta <- function(x, accuracy = 0.1) {
  sign <- ifelse(x > 0, "+", "")
  paste0(sign, scales::percent(x, accuracy = accuracy))
}

calc_ts <- function(points, fga, fta) {
  dplyr::if_else(
    (fga + 0.44 * fta) > 0,
    points / (2 * (fga + 0.44 * fta)),
    NA_real_
  )
}
```

```{r data-pull, cache=TRUE}
# ------------------------------------------------------------
# Game logs
# ------------------------------------------------------------

get_player_gamelogs <- function(season = SEASON, season_type = SEASON_TYPE) {
  hoopR::nba_leaguegamelog(
    season = season,
    season_type = season_type,
    player_or_team = "P"
  ) %>%
    purrr::pluck("LeagueGameLog") %>%
    transmute(
      player_name = PLAYER_NAME,
      player_id = PLAYER_ID,
      team = TEAM_ABBREVIATION,
      game_date = parse_nba_date(GAME_DATE),
      minutes = to_num(MIN),
      points = to_num(PTS),
      fgm = to_num(FGM),
      fga = to_num(FGA),
      fg3m = to_num(FG3M),
      fg3a = to_num(FG3A),
      ftm = to_num(FTM),
      fta = to_num(FTA),
      tov = to_num(TOV),
      ast = to_num(AST),
      reb = to_num(REB),
      ts_pct = calc_ts(points, fga, fta)
    ) %>%
    filter(!is.na(game_date)) %>%
    arrange(player_name, game_date)
}

games <- get_player_gamelogs()
```

```{r analysis-helpers}
# ------------------------------------------------------------
# Window helpers
# ------------------------------------------------------------

split_halves <- function(df, player) {
  df_p <- df %>%
    filter(player_name == player) %>%
    arrange(game_date)

  n <- nrow(df_p)

  if (n < 6) {
    return(NULL)
  }

  n_early <- floor(n / 2)

  list(
    early = df_p %>% slice_head(n = n_early),
    recent = df_p %>% slice_tail(n = n - n_early)
  )
}

split_halves_dates <- function(games_player) {
  dates <- games_player %>%
    arrange(game_date) %>%
    pull(game_date)

  n <- length(dates)
  n_early <- floor(n / 2)

  list(
    early_dates = dates[1:n_early],
    recent_dates = dates[(n_early + 1):n]
  )
}

summarize_window <- function(df) {
  df %>%
    summarise(
      gp = n(),
      mpg = mean(minutes, na.rm = TRUE),
      ppg = mean(points, na.rm = TRUE),
      ast_pg = mean(ast, na.rm = TRUE),
      reb_pg = mean(reb, na.rm = TRUE),
      ts = mean(ts_pct, na.rm = TRUE),
      fga_pg = mean(fga, na.rm = TRUE),
      three_rate = mean(fg3a / pmax(fga, 1), na.rm = TRUE),
      ft_rate = mean(fta / pmax(fga, 1), na.rm = TRUE),
      tov_pg = mean(tov, na.rm = TRUE)
    )
}

make_deltas <- function(early, recent) {
  tibble(
    gp_early = early$gp,
    gp_recent = recent$gp,
    mpg_early = early$mpg,
    mpg_recent = recent$mpg,
    ppg_early = early$ppg,
    ppg_recent = recent$ppg,
    ast_pg_early = early$ast_pg,
    ast_pg_recent = recent$ast_pg,
    reb_pg_early = early$reb_pg,
    reb_pg_recent = recent$reb_pg,
    ts_early = early$ts,
    ts_recent = recent$ts,
    fga_pg_early = early$fga_pg,
    fga_pg_recent = recent$fga_pg,
    three_rate_early = early$three_rate,
    three_rate_recent = recent$three_rate,
    ft_rate_early = early$ft_rate,
    ft_rate_recent = recent$ft_rate,
    tov_pg_early = early$tov_pg,
    tov_pg_recent = recent$tov_pg
  ) %>%
    mutate(
      ppg_delta = ppg_recent - ppg_early,
      ast_pg_delta = ast_pg_recent - ast_pg_early,
      reb_pg_delta = reb_pg_recent - reb_pg_early,
      ts_delta = ts_recent - ts_early,
      mpg_delta = mpg_recent - mpg_early,
      fga_pg_delta = fga_pg_recent - fga_pg_early,
      three_rate_delta = three_rate_recent - three_rate_early,
      ft_rate_delta = ft_rate_recent - ft_rate_early,
      tov_pg_delta = tov_pg_recent - tov_pg_early
    )
}

make_halves_table <- function(early_sum, recent_sum) {
  tibble(
    Window = c("Early", "Recent"),
    GP = c(early_sum$gp, recent_sum$gp),
    MPG = round(c(early_sum$mpg, recent_sum$mpg), 1),
    PPG = round(c(early_sum$ppg, recent_sum$ppg), 1),
    `AST/G` = round(c(early_sum$ast_pg, recent_sum$ast_pg), 1),
    `REB/G` = round(c(early_sum$reb_pg, recent_sum$reb_pg), 1),
    `TS%` = scales::percent(c(early_sum$ts, recent_sum$ts), accuracy = 0.1),
    `FGA/G` = round(c(early_sum$fga_pg, recent_sum$fga_pg), 1),
    `3PA Share` = scales::percent(c(early_sum$three_rate, recent_sum$three_rate), accuracy = 1),
    `FT Rate` = scales::percent(c(early_sum$ft_rate, recent_sum$ft_rate), accuracy = 1),
    `TOV/G` = round(c(early_sum$tov_pg, recent_sum$tov_pg), 2)
  )
}

build_player_window_summary <- function(player) {
  sp <- split_halves(games, player)

  if (is.null(sp)) {
    return(NULL)
  }

  team <- games %>%
    filter(player_name == player) %>%
    slice_tail(n = 1) %>%
    pull(team) %>%
    first()

  bind_rows(
    summarize_window(sp$early) %>%
      mutate(player = player, window = "Early", team = team),
    summarize_window(sp$recent) %>%
      mutate(player = player, window = "Recent", team = team)
  )
}

# ------------------------------------------------------------
# Shot helpers
# ------------------------------------------------------------

get_player_shots <- function(player_id, season = SEASON, season_type = SEASON_TYPE) {
  hoopR::nba_shotchartdetail(
    season = season,
    season_type = season_type,
    player_id = player_id,
    team_id = 0,
    context_measure_simple = "FGA"
  ) %>%
    purrr::pluck("Shot_Chart_Detail") %>%
    transmute(
      game_date = parse_nba_date(GAME_DATE),
      x = to_num(LOC_X),
      y = to_num(LOC_Y),
      made = as.integer(SHOT_MADE_FLAG)
    ) %>%
    filter(!is.na(game_date))
}

get_cached_shots <- function(player_id, season = SEASON) {
  dir.create(SHOT_CACHE_DIR, recursive = TRUE, showWarnings = FALSE)

  path <- file.path(SHOT_CACHE_DIR, paste0("shots_", player_id, ".csv"))

  if (file.exists(path)) {
    return(readr::read_csv(path, show_col_types = FALSE))
  }

  shots <- get_player_shots(player_id = player_id, season = season)
  readr::write_csv(shots, path)

  shots
}

build_player_shots_split <- function(games_df, shots_df, player_name) {
  player_games <- games_df %>%
    filter(player_name == .env$player_name) %>%
    arrange(game_date)

  windows <- split_halves_dates(player_games)

  shots_df %>%
    mutate(
      window = case_when(
        game_date %in% windows$early_dates ~ "Early",
        game_date %in% windows$recent_dates ~ "Recent",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(window))
}

add_zone <- function(df) {
  df %>%
    mutate(
      r = sqrt(x^2 + y^2),
      corner = abs(x) > 220 & y < 92,
      zone = case_when(
        r <= 48 ~ "Rim",
        r <= 120 ~ "Paint (non-rim)",
        r <= 220 ~ "Midrange",
        corner ~ "Corner 3",
        r > 220 ~ "Above-the-break 3",
        TRUE ~ "Other"
      ),
      zone = factor(
        zone,
        levels = c(
          "Rim",
          "Paint (non-rim)",
          "Midrange",
          "Corner 3",
          "Above-the-break 3",
          "Other"
        )
      )
    )
}

zone_summary <- function(shots_split) {
  shots_split %>%
    add_zone() %>%
    group_by(window, zone) %>%
    summarise(
      fga = n(),
      fgm = sum(made, na.rm = TRUE),
      fg_pct = fgm / fga,
      .groups = "drop"
    ) %>%
    filter(fga >= 5)
}

zone_delta <- function(zone_df) {
  zone_df %>%
    select(window, zone, fg_pct, fga) %>%
    tidyr::pivot_wider(names_from = window, values_from = c(fg_pct, fga)) %>%
    mutate(
      fg_delta = fg_pct_Recent - fg_pct_Early,
      fga_delta = fga_Recent - fga_Early
    ) %>%
    arrange(desc(abs(fg_delta)))
}

# ------------------------------------------------------------
# Display helpers
# ------------------------------------------------------------

plot_zone_eff <- function(zone_df, title = NULL) {
  ggplot(zone_df, aes(x = zone, y = window, fill = fg_pct)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(
      aes(label = paste0(round(fg_pct * 100), "%\n(", fga, ")")),
      size = 3.1,
      lineheight = 0.95
    ) +
    scale_fill_gradient2(
      low = "red3",
      mid = "orange",
      high = "chartreuse4",
      midpoint = 0.50,
      labels = scales::percent_format(accuracy = 1)
    ) +
    labs(
      title = title,
      subtitle = "FG% by zone. Cell shows FG% and attempts (FGA).",
      x = NULL,
      y = NULL,
      fill = "FG%"
    ) +
    theme_zone
}

write_player_paragraph <- function(player, deltas_row, zone_deltas) {
  top <- zone_deltas %>% slice_head(n = 2)

  z1 <- if (nrow(top) >= 1) {
    paste0(as.character(top$zone[1]), " (", fmt_pct_delta(top$fg_delta[1], 1), ")")
  } else {
    "N/A"
  }

  z2 <- if (nrow(top) >= 2) {
    paste0(as.character(top$zone[2]), " (", fmt_pct_delta(top$fg_delta[2], 1), ")")
  } else {
    "N/A"
  }

  paste0(
    "Across the second half of games played, ", player, " increased production by ",
    fmt_delta(deltas_row$ppg_delta, 1), " PPG with TS% up ",
    fmt_pct_delta(deltas_row$ts_delta, 0.1), ". ",
    "Minutes changed by ", fmt_delta(deltas_row$mpg_delta, 1),
    " MPG and shot volume changed by ",
    fmt_delta(deltas_row$fga_pg_delta, 1), " FGA/G. ",
    "The biggest zone-level FG% swings were ", z1, " and ", z2, "."
  )
}
```

# Executive Summary

This report flags players who improved after poor early-season results by comparing each player's first half of games played to their most recent half of games played.

The goal is to separate different types of rebounds. Some players improved because their role changed. Some improved because their shot diet changed. Others simply started making the same shots at a more normal rate.

# Data and Methods

The analysis uses 2025-26 NBA regular season player game logs and shot chart data from NBA.com through `hoopR`.

Windows are player-specific. For each player, the first half of games played is compared against the second half of games played. The core outcomes are points per game, true shooting percentage, minutes per game, and field-goal attempts per game.

The diagnostic indicators are 3PA share, free-throw rate, turnovers, and shot-zone efficiency.

# Overview Plot

```{r overview-plot, fig.width=6.2, fig.height=4.2}
plot_df <- purrr::map_dfr(PLAYERS, build_player_window_summary)

ggplot(
  plot_df,
  aes(x = three_rate, y = ts, group = player, color = team, shape = window)
) +
  geom_line(linewidth = 0.9, alpha = 0.7) +
  geom_point(size = 3.0) +
  ggrepel::geom_text_repel(
    data = filter(plot_df, window == "Recent"),
    aes(label = player),
    size = 3.0,
    box.padding = 0.35,
    point.padding = 0.25,
    min.segment.length = 0,
    segment.size = 0.3,
    segment.alpha = 0.6,
    max.overlaps = 12,
    force = 1,
    force_pull = 0.2,
    seed = 42,
    show.legend = FALSE
  ) +
  scale_color_manual(values = TEAM_COLORS) +
  scale_shape_manual(values = c(Early = 1, Recent = 16)) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.12))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    expand = expansion(mult = c(0.03, 0.10))
  ) +
  labs(
    title = "Rebound After Slow Starts",
    subtitle = "Early vs. recent half-season splits",
    x = "3PA Share (3PA / FGA)",
    y = "True Shooting %",
    caption = "Data: NBA game logs via hoopR | Windows split by games played | Plot: @Rambzee_"
  ) +
  theme_caleb
```

# Summary Tables

```{r summary-tables}
summary_tbl <- purrr::map_dfr(PLAYERS, function(player) {
  sp <- split_halves(games, player)

  if (is.null(sp)) {
    return(NULL)
  }

  early_sum <- summarize_window(sp$early)
  recent_sum <- summarize_window(sp$recent)

  make_deltas(early_sum, recent_sum) %>%
    mutate(player = player)
}) %>%
  select(
    player,
    gp_early, gp_recent,
    mpg_early, mpg_recent,
    ppg_early, ppg_recent, ppg_delta,
    ast_pg_early, ast_pg_recent, ast_pg_delta,
    reb_pg_early, reb_pg_recent, reb_pg_delta,
    ts_early, ts_recent, ts_delta,
    fga_pg_early, fga_pg_recent, fga_pg_delta,
    three_rate_early, three_rate_recent, three_rate_delta,
    ft_rate_early, ft_rate_recent, ft_rate_delta,
    tov_pg_early, tov_pg_recent, tov_pg_delta
  ) %>%
  arrange(desc(ppg_delta))

summary_core <- summary_tbl %>%
  transmute(
    Player = player,
    `GP (E→R)` = paste0(gp_early, "→", gp_recent),
    `MPG (E→R)` = paste0(round(mpg_early, 1), "→", round(mpg_recent, 1)),
    `PPG Δ` = fmt_delta(ppg_delta, 1),
    `AST/G Δ` = fmt_delta(ast_pg_delta, 1),
    `REB/G Δ` = fmt_delta(reb_pg_delta, 1),
    `TS% Δ` = fmt_pct_delta(ts_delta, 0.1),
    `FGA/G Δ` = fmt_delta(fga_pg_delta, 1)
  )

knitr::kable(
  summary_core,
  booktabs = TRUE,
  caption = "Core rebound signals: first half vs. second half of games played"
)

summary_diet <- summary_tbl %>%
  transmute(
    Player = player,
    `3PA Share (E→R)` = paste0(
      scales::percent(three_rate_early, accuracy = 1),
      "→",
      scales::percent(three_rate_recent, accuracy = 1)
    ),
    `3PA Share Δ` = fmt_pct_delta(three_rate_delta, 1),
    `FT Rate (E→R)` = paste0(
      scales::percent(ft_rate_early, accuracy = 1),
      "→",
      scales::percent(ft_rate_recent, accuracy = 1)
    ),
    `FT Rate Δ` = fmt_pct_delta(ft_rate_delta, 1),
    `TOV/G (E→R)` = paste0(round(tov_pg_early, 2), "→", round(tov_pg_recent, 2)),
    `TOV/G Δ` = fmt_delta(tov_pg_delta, 2)
  )

knitr::kable(
  summary_diet,
  booktabs = TRUE,
  caption = "Shot diet and role proxies: first half vs. second half"
)
```

# How to Interpret the Diagnostics

If 3PA share and FT rate are stable but true shooting jumps, the rebound is likely driven by shot-making regression.

If 3PA share rises, the player's role may have shifted toward spacing, catch-and-shoot usage, or a cleaner shot profile.

If free-throw rate rises, the player may be getting more rim pressure, more physical usage, or a stronger on-ball role.

If turnovers rise with production, the player may be carrying more on-ball responsibility.

# Player Pages

```{r player-pages, results='asis', fig.width=8.5, fig.height=3.8}
for (player in PLAYERS) {
  sp <- split_halves(games, player)

  if (is.null(sp)) {
    next
  }

  early_sum <- summarize_window(sp$early)
  recent_sum <- summarize_window(sp$recent)

  halves_tbl <- make_halves_table(early_sum, recent_sum)
  deltas_row <- make_deltas(early_sum, recent_sum)

  delta_tbl <- deltas_row %>%
    transmute(
      `PPG Δ` = fmt_delta(ppg_delta, 1),
      `AST/G Δ` = fmt_delta(ast_pg_delta, 1),
      `REB/G Δ` = fmt_delta(reb_pg_delta, 1),
      `TS% Δ` = fmt_pct_delta(ts_delta, 0.1),
      `MPG Δ` = fmt_delta(mpg_delta, 1),
      `FGA/G Δ` = fmt_delta(fga_pg_delta, 1),
      `3PA Share Δ` = fmt_pct_delta(three_rate_delta, 1),
      `FT Rate Δ` = fmt_pct_delta(ft_rate_delta, 1),
      `TOV/G Δ` = fmt_delta(tov_pg_delta, 2)
    )

  player_id <- games %>%
    filter(player_name == player) %>%
    slice_tail(n = 1) %>%
    pull(player_id) %>%
    first()

  shots <- get_cached_shots(player_id = player_id, season = SEASON)

  shots_split <- build_player_shots_split(
    games_df = games,
    shots_df = shots,
    player_name = player
  )

  zone_df <- zone_summary(shots_split)
  zone_deltas <- zone_delta(zone_df)

  cat("\n\n\\newpage\n")
  cat("\n\n# ", player, "\n\n", sep = "")
  cat(write_player_paragraph(player, deltas_row, zone_deltas), "\n\n")

  cat("## First Half vs. Second Half\n\n")
  cat(knitr::kable(halves_tbl, booktabs = TRUE), sep = "\n")

  cat("\n\n## Change Summary\n\n")
  cat(knitr::kable(delta_tbl, booktabs = TRUE), sep = "\n")

  cat("\n\n## Shot Zone Efficiency\n\n")

  print(
    plot_zone_eff(
      zone_df,
      title = paste0(player, " shot zone efficiency: Early vs. Recent")
    )
  )
}
```
