---
title: "Scoring Translation in the NBA Playoffs (Since 2000)"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
geometry: margin=0.8in
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(tidyverse)
library(hoopR)
library(janitor)
library(scales)
library(kableExtra)
library(showtext)
library(sysfonts)
library(forcats)

font_add_google("Inter", "inter")
showtext_auto()
options(scipen = 999)

BG <- "#f6f4ef"
TEXT_DARK <- "#141414"
TEXT_LIGHT <- "#6b6b6b"
GRID <- "#dedad3"
ACCENT <- "#1d4ed8"
HILITE <- "#e8f0ff"
RISE <- "#15803d"
DROP <- "#b91c1c"

CAPTION_LINE <- "Data: NBA.com via hoopR | Viz: @Rambzee_"

theme_caleb <- function() {
  theme_minimal(base_family = "inter") +
    theme(
      plot.background = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = TEXT_DARK),
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11.5, color = TEXT_LIGHT, hjust = 0.5),
      axis.title = element_text(size = 11.5, face = "bold"),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 10),
      plot.caption = element_text(size = 8, color = "#6e6e6e", hjust = 1, margin = margin(t = 8))
    )
}

style_kable <- function(df, align = NULL, font_size = 9.5, full_width = FALSE) {
  if (is.null(align)) {
    align <- strrep("c", ncol(df))
  }

  kbl(
    df,
    booktabs = TRUE,
    align = align,
    format = "latex",
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = full_width,
      font_size = font_size,
      stripe_color = "#fbfaf7"
    )
}

safe_percent_change <- function(new, old) {
  ifelse(is.na(old) | old == 0, NA_real_, 100 * (new - old) / old)
}

calc_ts <- function(pts, fga, fta) {
  denom <- 2 * (fga + 0.44 * fta)
  ifelse(is.na(denom) | denom == 0, NA_real_, pts / denom)
}
```

```{r parameters-and-helpers, include=FALSE}
START_SEASON <- 2000
END_SEASON   <- 2024

MIN_RS_GAMES <- 40
MIN_PO_GAMES <- 8
MIN_RS_PPG   <- 10

MIN_CAREER_RS_GAMES <- 250
MIN_CAREER_PO_GAMES <- 50
MIN_LEADERBOARD_PO_GAMES <- 50

TOP_N <- 12

season_labels <- function(start_year, end_year) {
  yrs <- start_year:end_year
  paste0(yrs, "-", stringr::str_sub(yrs + 1, 3, 4))
}

modern_seasons <- season_labels(START_SEASON, END_SEASON)

pull_player_per_game <- function(season, season_type = "Regular Season") {
  out <- nba_leaguedashplayerstats(
    season = season,
    season_type = season_type,
    per_mode = "PerGame"
  )

  if (is.null(out$LeagueDashPlayerStats) || nrow(out$LeagueDashPlayerStats) == 0) {
    return(tibble())
  }

  out$LeagueDashPlayerStats %>%
    clean_names()
}

pull_team_per_game <- function(season, season_type = "Regular Season") {
  out <- nba_leaguedashteamstats(
    season = season,
    season_type = season_type,
    per_mode = "PerGame"
  )

  if (is.null(out$LeagueDashTeamStats) || nrow(out$LeagueDashTeamStats) == 0) {
    return(tibble())
  }

  out$LeagueDashTeamStats %>%
    clean_names()
}

league_ts <- tibble(
  season = c(
    "2000-01", "2001-02", "2002-03", "2003-04", "2004-05", "2005-06",
    "2006-07", "2007-08", "2008-09", "2009-10", "2010-11", "2011-12",
    "2012-13", "2013-14", "2014-15", "2015-16", "2016-17", "2017-18",
    "2018-19", "2019-20", "2020-21", "2021-22", "2022-23", "2023-24", "2024-25"
  ),
  league_rs_ts = c(
    51.8, 52.0, 51.9, 51.6, 52.9, 53.6,
    54.1, 54.0, 54.4, 54.3, 54.1, 52.7,
    53.5, 54.1, 53.4, 54.1, 55.2, 55.6,
    56.0, 56.5, 57.2, 56.6, 58.1, 58.0, 57.6
  ),
  league_po_ts = c(
    51.1, 51.4, 52.5, 50.0, 53.4, 54.7,
    53.0, 53.2, 54.4, 54.3, 52.9, 52.0,
    52.9, 54.8, 52.6, 53.4, 56.3, 55.5,
    55.1, 57.4, 57.1, 56.7, 56.6, 56.6, 56.6
  )
)

read_or_build <- function(path, builder) {
  if (file.exists(path)) {
    readRDS(path)
  } else {
    obj <- builder()
    saveRDS(obj, path)
    obj
  }
}

get_ppg_split <- function(season) {
  rs <- pull_player_per_game(season, "Regular Season")
  po <- pull_player_per_game(season, "Playoffs")

  if (nrow(rs) == 0 || nrow(po) == 0) {
    return(tibble())
  }

  rs_tbl <- rs %>%
    transmute(
      player_id = as.character(player_id),
      player_name,
      team_abbreviation,
      season = season,
      rs_games = as.numeric(gp),
      rs_pts_pg = as.numeric(pts),
      rs_ast_pg = as.numeric(ast),
      rs_reb_pg = as.numeric(reb),
      rs_fga_pg = as.numeric(fga),
      rs_fta_pg = as.numeric(fta),
      rs_points = as.numeric(pts) * as.numeric(gp),
      rs_fga = as.numeric(fga) * as.numeric(gp),
      rs_fta = as.numeric(fta) * as.numeric(gp),
      rs_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(
      rs_ts = calc_ts(rs_points, rs_fga, rs_fta)
    )

  po_tbl <- po %>%
    transmute(
      player_id = as.character(player_id),
      season = season,
      po_games = as.numeric(gp),
      po_pts_pg = as.numeric(pts),
      po_ast_pg = as.numeric(ast),
      po_reb_pg = as.numeric(reb),
      po_fga_pg = as.numeric(fga),
      po_fta_pg = as.numeric(fta),
      po_points = as.numeric(pts) * as.numeric(gp),
      po_fga = as.numeric(fga) * as.numeric(gp),
      po_fta = as.numeric(fta) * as.numeric(gp),
      po_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(
      po_ts = calc_ts(po_points, po_fga, po_fta)
    )

  rs_tbl %>%
    inner_join(po_tbl, by = c("player_id", "season")) %>%
    left_join(league_ts, by = "season") %>%
    mutate(
      ppg_diff = po_pts_pg - rs_pts_pg,
      apg_diff = po_ast_pg - rs_ast_pg,
      rpg_diff = po_reb_pg - rs_reb_pg,
      pct_change = safe_percent_change(po_pts_pg, rs_pts_pg),
      fga_change = po_fga_pg - rs_fga_pg,
      load_change = po_load - rs_load,
      rs_rts = 100 * (rs_ts - league_rs_ts / 100),
      po_rts = 100 * (po_ts - league_po_ts / 100),
      rts_change = po_rts - rs_rts
    )
}

season_level_ppg <- read_or_build(
  "season_level_ppg.rds",
  function() map_dfr(modern_seasons, get_ppg_split)
)

season_level_filtered <- season_level_ppg %>%
  filter(
    rs_games >= MIN_RS_GAMES,
    po_games >= MIN_PO_GAMES,
    rs_pts_pg >= MIN_RS_PPG
  )

modern_playoff_raw <- read_or_build(
  "modern_playoff_raw.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      po <- tryCatch(
        pull_player_per_game(season, "Playoffs"),
        error = function(e) NULL
      )

      if (is.null(po) || nrow(po) == 0) return(tibble())

      po %>%
        transmute(
          player_id = as.character(player_id),
          player_name,
          po_games = as.numeric(gp),
          po_points = as.numeric(pts) * as.numeric(gp),
          po_assists = as.numeric(ast) * as.numeric(gp),
          po_rebounds = as.numeric(reb) * as.numeric(gp)
        )
    })
  }
)

modern_playoff <- modern_playoff_raw %>%
  group_by(player_id, player_name) %>%
  summarise(
    po_games = sum(po_games, na.rm = TRUE),
    total_points = sum(po_points, na.rm = TRUE),
    total_assists = sum(po_assists, na.rm = TRUE),
    total_rebounds = sum(po_rebounds, na.rm = TRUE),
    playoff_ppg = total_points / po_games,
    playoff_apg = total_assists / po_games,
    playoff_rpg = total_rebounds / po_games,
    .groups = "drop"
  ) %>%
  filter(po_games >= MIN_LEADERBOARD_PO_GAMES)

modern_ppg_leaders <- modern_playoff %>%
  arrange(desc(playoff_ppg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff PPG` = round(playoff_ppg, 1)
  ) %>%
  slice_head(n = 10)

modern_apg_leaders <- modern_playoff %>%
  arrange(desc(playoff_apg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff APG` = round(playoff_apg, 1)
  ) %>%
  slice_head(n = 10)

modern_rpg_leaders <- modern_playoff %>%
  arrange(desc(playoff_rpg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff RPG` = round(playoff_rpg, 1)
  ) %>%
  slice_head(n = 10)

career_translation <- season_level_ppg %>%
  group_by(player_id, player_name) %>%
  summarise(
    career_rs_games = sum(rs_games, na.rm = TRUE),
    career_po_games = sum(po_games, na.rm = TRUE),
    career_rs_points = sum(rs_points, na.rm = TRUE),
    career_po_points = sum(po_points, na.rm = TRUE),
    career_rs_fga = sum(rs_fga, na.rm = TRUE),
    career_po_fga = sum(po_fga, na.rm = TRUE),
    career_rs_fta = sum(rs_fta, na.rm = TRUE),
    career_po_fta = sum(po_fta, na.rm = TRUE),
    career_rs_ppg = career_rs_points / career_rs_games,
    career_po_ppg = career_po_points / career_po_games,
    career_rs_ts = calc_ts(career_rs_points, career_rs_fga, career_rs_fta),
    career_po_ts = calc_ts(career_po_points, career_po_fga, career_po_fta),
    career_rs_rts = weighted.mean(rs_rts, w = rs_games, na.rm = TRUE),
    career_po_rts = weighted.mean(po_rts, w = po_games, na.rm = TRUE),
    career_ppg_diff = career_po_ppg - career_rs_ppg,
    career_pct_change = safe_percent_change(career_po_ppg, career_rs_ppg),
    career_rts_change = career_po_rts - career_rs_rts,
    career_rs_load = weighted.mean(rs_load, w = rs_games, na.rm = TRUE),
    career_po_load = weighted.mean(po_load, w = po_games, na.rm = TRUE),
    career_load_change = career_po_load - career_rs_load,
    .groups = "drop"
  ) %>%
  filter(
    career_rs_games >= MIN_CAREER_RS_GAMES,
    career_po_games >= MIN_CAREER_PO_GAMES
  )

career_risers <- career_translation %>%
  arrange(desc(career_ppg_diff), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1),
    `Load Δ` = round(career_load_change, 1)
  ) %>%
  slice_head(n = TOP_N)

career_maintainers <- career_translation %>%
  filter(abs(career_ppg_diff) <= 1) %>%
  arrange(desc(career_po_ppg), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1),
    `Load Δ` = round(career_load_change, 1)
  ) %>%
  slice_head(n = TOP_N)

career_droppers <- career_translation %>%
  filter(career_load_change >= -1.5) %>%
  arrange(career_ppg_diff, career_rts_change) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1),
    `Load Δ` = round(career_load_change, 1)
  ) %>%
  slice_head(n = TOP_N)

biggest_risers_season <- season_level_filtered %>%
  arrange(desc(ppg_diff), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `Load Δ` = round(load_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

biggest_droppers_season <- season_level_filtered %>%
  filter(load_change >= -1.5) %>%
  arrange(ppg_diff, rts_change) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `Load Δ` = round(load_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

maintainers_season <- season_level_filtered %>%
  filter(dplyr::between(ppg_diff, -1, 1)) %>%
  arrange(desc(po_pts_pg), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `Load Δ` = round(load_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

league_scoring <- read_or_build(
  "league_scoring_modern.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      rs <- tryCatch(pull_team_per_game(season, "Regular Season"), error = function(e) NULL)
      po <- tryCatch(pull_team_per_game(season, "Playoffs"), error = function(e) NULL)

      bind_rows(
        if (!is.null(rs) && nrow(rs) > 0) {
          tibble(
            season = season,
            season_start = as.numeric(substr(season, 1, 4)),
            context = "Regular Season",
            ppg = mean(as.numeric(rs$pts), na.rm = TRUE)
          )
        },
        if (!is.null(po) && nrow(po) > 0) {
          tibble(
            season = season,
            season_start = as.numeric(substr(season, 1, 4)),
            context = "Playoffs",
            ppg = mean(as.numeric(po$pts), na.rm = TRUE)
          )
        }
      )
    }) %>%
      filter(!is.na(ppg))
  }
)
```

# Project Overview

Playoff basketball is a different environment. Scoring drops, possessions tighten, and defenses are built to take away what a player does best.

That makes the idea of “rising in the playoffs” more complicated than it sounds. A scoring jump in a worse environment means something very different than it would in the regular season.

This analysis looks at how scoring actually changes once players reach the playoffs, with a focus on whether production, efficiency, and scoring load carry over.

## Definitions

This study focuses on scoring translation: how a player’s playoff scoring compares to their regular-season averages.

### Pretext

Regular-season production is measured only from seasons in which the player reached the playoffs.

This keeps the comparison focused on a player’s playoff-era role and avoids early-career or non-playoff seasons from distorting the baseline. Controlling this variability is essential to interpreting how players truly played relative to the level they were at.

### Scoring Change

PPG Difference (Diff) = Playoff PPG – Regular Season PPG

Positive values = increase in scoring. Negative values = decline.

### Efficiency 

rTS Change measures how a player’s scoring efficiency changes relative to league average between the regular season and playoffs.

It is calculated as a player’s playoff rTS minus their regular-season rTS.

A negative value indicates the player became less efficient relative to league average, even if overall league efficiency also declined. The change should be read alongside playoff rTS itself, as a player can decline relative to their own average while still remaining highly efficient overall.

### Load

Load estimates a player’s scoring responsibility, calculated as:

FGA + 0.44 × FTA

It reflects how often a player is involved in scoring plays.

### Load Change

Load Δ = Playoff Load – Regular Season Load

This helps separate role from performance.

If scoring drops along with load, it likely reflects reduced responsibility. If scoring drops while load stays similar, it points to a real decline. The same idea applies to increases in scoring.

### Classification

Players are grouped into three categories:

* Risers: scoring increases in the playoffs
* Maintainers: scoring remains within ±1 PPG
* Droppers: scoring decreases

For droppers, results are filtered to remove most obvious role-loss cases. This helps isolate players whose scoring declined despite similar offensive responsibility, making the drop more meaningful.

Riser seasons are interpreted with more context. Some increases come from higher scoring load, while others reflect genuine shotmaking and efficiency gains.

## Playoff Scoring (Regular Season vs Playoffs - Since 2000) 

```{r league-scoring-plot, fig.height=4.7, fig.width=7.5}
ggplot(
  league_scoring,
  aes(x = season_start, y = ppg, color = context)
) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Playoffs" = DROP, "Regular Season" = ACCENT)) +
  scale_x_continuous(breaks = seq(2000, END_SEASON, by = 4)) +
  labs(
    title = "Scoring consistently dips once the playoffs begin",
    subtitle = "League average team scoring by season since 2000-01",
    x = NULL,
    y = "Points per game",
    caption = CAPTION_LINE
  ) +
  theme_caleb()
```

While scoring has increased across the league since the turn of the century, one trend has held constant: production drops in the playoffs.

That gap is what creates the tension of postseason basketball. Some players find ways to hold their level, a few push beyond it, and many fall off once defenses are built entirely to stop them.

In that context, “rising” takes on a different meaning. In a lower-scoring setting, simply maintaining production is already a triumph. True increases are rare, and when they happen, they need to be understood in context.

# Best Playoff Performers of All-Time and of the Modern Era

Raw playoff production still matters, especially over larger samples. Before getting into translation, it helps to establish the players who have simply produced the most on the playoff stage in the modern era.

## Top playoff scorers

```{r ppg-leaders}
modern_ppg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

## Top playoff playmakers

```{r apg-leaders}
modern_apg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

## Top playoff rebounders

```{r rpg-leaders}
modern_rpg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

Those tables tell us who has produced the most in the playoffs. They do not tell us whose games actually translate best.

The more revealing question is what happens when playoff scoring is compared to a player's regular-season baseline, then layered with efficiency and scoring load.

# Playoff-Season Baseline Translation

To keep the comparison cleaner, the regular-season side here uses only seasons in which a player actually reached the playoffs. That avoids early-career and non-playoff years from dragging down the baseline and keeps the comparison closer to a player's playoff-era version of himself.

This is not trying to answer who had the bigger career arc. It is trying to answer who held up best once they were actually in playoff environments.

## Quick takeaways

The riser table is not as clean as it looks if efficiency falls at the same time. Some players do find another gear, but many “risers” are really trading a bit of efficiency for more responsibility.

The maintainer table is the real signal. In a postseason environment where most scorers lose something, the players who can keep their volume and avoid a big efficiency drop are usually the ones whose games travel best.

The dropper table is filtered to remove most obvious role-loss cases. If scoring falls while load stays roughly stable, that is much more meaningful than a player simply getting phased into a smaller offensive role.

## Biggest career risers

```{r career-risers}
career_risers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

## Best career maintainers

```{r career-maintainers}
career_maintainers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

## Biggest career droppers

```{r career-droppers}
career_droppers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

# Since 2000: Who Actually Scales?

Career summaries smooth everything out. Single postseasons show the sharper edges: role changes, matchup advantages, injury context, and the rare cases where a player's game fully scales into playoff basketball.

These tables are noisier than the career ones by nature, but they are also where the clearest single-run examples show up.

## Biggest playoff risers by season

Some of these jumps reflect real postseason scaling. Others come with a bigger scoring load or a slightly shakier efficiency profile. That is why the scoring change and the rTS change need to be read together.

```{r season-risers}
biggest_risers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

## Biggest playoff droppers by season

This is the cleaner version of the dropper idea. The table is filtered to strip out most obvious role-loss seasons, so what remains is closer to true scoring slippage: players whose production fell even without a major reduction in scoring load.

```{r season-droppers}
biggest_droppers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

## Most important: the maintainers

Most players drop. These players do not.

This is probably the clearest signal in the whole piece. Holding your scoring in the playoffs is already hard. Doing it while keeping load and efficiency reasonably intact is what separates a star regular-season scorer from someone whose game really survives playoff basketball.

```{r season-maintainers}
maintainers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

# What This Suggests

The playoffs do not just test talent. They test translation.

League-wide scoring comes down, possessions get tighter, and defenses spend entire series attacking the weak points of a player's offensive game. That is why playoff production by itself is not enough. The more useful question is what happens to a player's scoring, efficiency, and load once the environment turns against him.

Most players decline. A few rise. But the rarest outcome is simpler than that: maintaining star-level production in a setting built to take it away.

That is the group worth paying attention to.

\bigskip

\small
\textit{* RS PPG in the career tables is based only on seasons in which the player reached the playoffs, so the baseline reflects the player’s playoff-era regular seasons rather than every season of his career.}
