---
title: "Scoring Translation in the NBA Playoffs (Since 2000)"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
geometry: margin=0.8in
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(tidyverse)
library(hoopR)
library(janitor)
library(scales)
library(kableExtra)
library(showtext)
library(sysfonts)
library(forcats)

font_add_google("Inter", "inter")
showtext_auto()
options(scipen = 999)

BG <- "#f6f4ef"
TEXT_DARK <- "#141414"
TEXT_LIGHT <- "#6b6b6b"
GRID <- "#dedad3"
ACCENT <- "#1d4ed8"
HILITE <- "#e8f0ff"
RISE <- "#15803d"
DROP <- "#b91c1c"

CAPTION_LINE <- "Data: NBA.com via hoopR | Viz: @Rambzee_"

theme_caleb <- function() {
  theme_minimal(base_family = "inter") +
    theme(
      plot.background = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = TEXT_DARK),
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11.5, color = TEXT_LIGHT, hjust = 0.5),
      axis.title = element_text(size = 11.5, face = "bold"),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 10),
      plot.caption = element_text(size = 8, color = "#6e6e6e", hjust = 1, margin = margin(t = 8))
    )
}

style_kable <- function(df, align = NULL, font_size = 9.5, full_width = FALSE) {
  if (is.null(align)) {
    align <- strrep("c", ncol(df))
  }

  kbl(
    df,
    booktabs = TRUE,
    align = align,
    format = "latex",
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = full_width,
      font_size = font_size,
      stripe_color = "#fbfaf7"
    )
}

safe_percent_change <- function(new, old) {
  ifelse(is.na(old) | old == 0, NA_real_, 100 * (new - old) / old)
}

calc_ts <- function(pts, fga, fta) {
  denom <- 2 * (fga + 0.44 * fta)
  ifelse(is.na(denom) | denom == 0, NA_real_, pts / denom)
}
```

```{r parameters-and-helpers, include=FALSE}
START_SEASON <- 2000
END_SEASON   <- 2024

MIN_RS_GAMES <- 40
MIN_PO_GAMES <- 8
MIN_RS_PPG   <- 10

MIN_CAREER_RS_GAMES <- 250
MIN_CAREER_PO_GAMES <- 50
MIN_LEADERBOARD_PO_GAMES <- 50

TOP_N <- 12

CACHE_REBUILD <- FALSE
VALIDATE_DATA <- TRUE

season_labels <- function(start_year, end_year) {
  yrs <- start_year:end_year
  paste0(yrs, "-", stringr::str_sub(yrs + 1, 3, 4))
}

modern_seasons <- season_labels(START_SEASON, END_SEASON)

pull_player_per_game <- function(season, season_type = "Regular Season", max_attempts = 3) {
  for (i in seq_len(max_attempts)) {
    out <- tryCatch(
      nba_leaguedashplayerstats(
        season = season,
        season_type = season_type,
        per_mode = "PerGame"
      ),
      error = function(e) NULL
    )

    if (!is.null(out$LeagueDashPlayerStats) && nrow(out$LeagueDashPlayerStats) > 0) {
      return(out$LeagueDashPlayerStats %>% clean_names())
    }

    message(sprintf(
      "EMPTY pull_player_per_game: %s | %s | attempt %s",
      season, season_type, i
    ))

    Sys.sleep(2 * i)
  }

  tibble()
}

pull_player_totals <- function(season, season_type = "Regular Season") {
  out <- tryCatch(
    nba_leaguedashplayerstats(
      season = season,
      season_type = season_type,
      per_mode = "Totals"
    ),
    error = function(e) {
      message(sprintf("FAILED pull_player_totals: %s | %s | %s", season, season_type, e$message))
      return(NULL)
    }
  )

  if (is.null(out) || is.null(out$LeagueDashPlayerStats) || nrow(out$LeagueDashPlayerStats) == 0) {
    message(sprintf("EMPTY pull_player_totals: %s | %s", season, season_type))
    return(tibble())
  }

  out$LeagueDashPlayerStats %>%
    clean_names()
}

pull_team_per_game <- function(season, season_type = "Regular Season") {
  out <- tryCatch(
    nba_leaguedashteamstats(
      season = season,
      season_type = season_type,
      per_mode = "PerGame"
    ),
    error = function(e) {
      message(sprintf("FAILED pull_team_per_game: %s | %s | %s", season, season_type, e$message))
      return(NULL)
    }
  )

  if (is.null(out) || is.null(out$LeagueDashTeamStats) || nrow(out$LeagueDashTeamStats) == 0) {
    message(sprintf("EMPTY pull_team_per_game: %s | %s", season, season_type))
    return(tibble())
  }

  out$LeagueDashTeamStats %>%
    clean_names()
}

league_ts <- tibble(
  season = c(
    "2000-01", "2001-02", "2002-03", "2003-04", "2004-05", "2005-06",
    "2006-07", "2007-08", "2008-09", "2009-10", "2010-11", "2011-12",
    "2012-13", "2013-14", "2014-15", "2015-16", "2016-17", "2017-18",
    "2018-19", "2019-20", "2020-21", "2021-22", "2022-23", "2023-24", "2024-25"
  ),
  league_rs_ts = c(
    51.8, 52.0, 51.9, 51.6, 52.9, 53.6,
    54.1, 54.0, 54.4, 54.3, 54.1, 52.7,
    53.5, 54.1, 53.4, 54.1, 55.2, 55.6,
    56.0, 56.5, 57.2, 56.6, 58.1, 58.0, 57.6
  ),
  league_po_ts = c(
    51.1, 51.4, 52.5, 50.0, 53.4, 54.7,
    53.0, 53.2, 54.4, 54.3, 52.9, 52.0,
    52.9, 54.8, 52.6, 53.4, 56.3, 55.5,
    55.1, 57.4, 57.1, 56.7, 56.6, 56.6, 56.6
  )
)

read_or_build <- function(path, builder, rebuild = FALSE, allow_empty = FALSE) {
  if (file.exists(path) && !rebuild) {
    return(readRDS(path))
  }

  obj <- builder()

  if (!allow_empty) {
    if (is.null(obj)) {
      stop(sprintf("Builder returned NULL for: %s", path))
    }

    if (inherits(obj, "data.frame") && nrow(obj) == 0) {
      stop(sprintf("Builder returned 0 rows for: %s", path))
    }
  }

  saveRDS(obj, path)
  obj
}

audit_season_coverage <- function(df, seasons = modern_seasons) {
  df %>%
    count(season, name = "rows") %>%
    tidyr::complete(season = seasons, fill = list(rows = 0)) %>%
    arrange(season)
}

check_missing_seasons <- function(df, seasons = modern_seasons, dataset_name = deparse(substitute(df))) {
  audit_tbl <- audit_season_coverage(df, seasons)

  missing_tbl <- audit_tbl %>%
    filter(rows == 0)

  if (nrow(missing_tbl) > 0) {
    stop(
      sprintf(
        "Missing season data in %s: %s",
        dataset_name,
        paste(missing_tbl$season, collapse = ', ')
      )
    )
  }

  audit_tbl
}

audit_player_presence <- function(player_search, df = season_level_ppg) {
  df %>%
    filter(
      stringr::str_detect(
        stringr::str_to_lower(player_name),
        stringr::str_to_lower(player_search)
      ) |
        player_id == as.character(player_search)
    ) %>%
    arrange(player_name, season) %>%
    select(
      player_id, player_name, season,
      rs_games, po_games,
      rs_pts_pg, po_pts_pg,
      rs_ts, po_ts
    )
}

get_ppg_split <- function(season) {
  rs <- pull_player_per_game(season, "Regular Season")
  po <- pull_player_per_game(season, "Playoffs")

  if (nrow(rs) == 0 || nrow(po) == 0) {
    stop(sprintf("Missing RS or PO data for season %s", season))
  }

  rs_tbl <- rs %>%
    transmute(
      player_id = as.character(player_id),
      player_name,
      team_abbreviation,
      season = season,
      rs_games = as.numeric(gp),
      rs_mpg = as.numeric(min),
      rs_pts_pg = as.numeric(pts),
      rs_ast_pg = as.numeric(ast),
      rs_reb_pg = as.numeric(reb),
      rs_fga_pg = as.numeric(fga),
      rs_fta_pg = as.numeric(fta),
      rs_points = as.numeric(pts) * as.numeric(gp),
      rs_fga = as.numeric(fga) * as.numeric(gp),
      rs_fta = as.numeric(fta) * as.numeric(gp),
      rs_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(rs_ts = calc_ts(rs_points, rs_fga, rs_fta))

  po_tbl <- po %>%
    transmute(
      player_id = as.character(player_id),
      season = season,
      po_games = as.numeric(gp),
      po_mpg = as.numeric(min),
      po_pts_pg = as.numeric(pts),
      po_ast_pg = as.numeric(ast),
      po_reb_pg = as.numeric(reb),
      po_fga_pg = as.numeric(fga),
      po_fta_pg = as.numeric(fta),
      po_points = as.numeric(pts) * as.numeric(gp),
      po_fga = as.numeric(fga) * as.numeric(gp),
      po_fta = as.numeric(fta) * as.numeric(gp),
      po_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(po_ts = calc_ts(po_points, po_fga, po_fta))

  rs_tbl %>%
    inner_join(po_tbl, by = c("player_id", "season")) %>%
    left_join(league_ts, by = "season") %>%
    mutate(
      ppg_diff = po_pts_pg - rs_pts_pg,
      apg_diff = po_ast_pg - rs_ast_pg,
      rpg_diff = po_reb_pg - rs_reb_pg,
      pct_change = safe_percent_change(po_pts_pg, rs_pts_pg),
      fga_change = po_fga_pg - rs_fga_pg,
      load_change = po_load - rs_load,
      mpg_change = po_mpg - rs_mpg,
      rs_rts = 100 * (rs_ts - league_rs_ts / 100),
      po_rts = 100 * (po_ts - league_po_ts / 100),
      rts_change = po_rts - rs_rts
    )
}

season_level_list <- lapply(modern_seasons, function(s) {
  message("Building: ", s)
  get_ppg_split(s)
})

season_level_ppg <- bind_rows(season_level_list)

season_level_audit <- audit_season_coverage(season_level_ppg)

if (VALIDATE_DATA) {
  missing_seasons_initial <- season_level_audit %>%
    filter(rows == 0) %>%
    pull(season)

  if (length(missing_seasons_initial) > 0) {
    message(
      sprintf(
        "Incomplete cached season_level_ppg detected. Force rebuilding missing/incomplete cache. Missing seasons: %s",
        paste(missing_seasons_initial, collapse = ", ")
      )
    )

    season_level_ppg <- read_or_build(
      "season_level_ppg.rds",
      function() map_dfr(modern_seasons, get_ppg_split),
      rebuild = TRUE
    )

    season_level_audit <- audit_season_coverage(season_level_ppg)
  }

  check_missing_seasons(season_level_ppg, dataset_name = "season_level_ppg")
}

season_level_filtered <- season_level_ppg %>%
  filter(
    rs_games >= MIN_RS_GAMES,
    po_games >= MIN_PO_GAMES,
    rs_pts_pg >= MIN_RS_PPG
  )

modern_playoff_raw <- read_or_build(
  "modern_playoff_raw.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      po <- tryCatch(
        pull_player_per_game(season, "Playoffs"),
        error = function(e) NULL
      )

      if (is.null(po) || nrow(po) == 0) return(tibble())

      po %>%
        transmute(
          player_id = as.character(player_id),
          player_name,
          po_games = as.numeric(gp),
          po_points = as.numeric(pts) * as.numeric(gp),
          po_assists = as.numeric(ast) * as.numeric(gp),
          po_rebounds = as.numeric(reb) * as.numeric(gp)
        )
    })
  }
,
  rebuild = CACHE_REBUILD
)

modern_playoff_audit <- modern_playoff_raw %>%
  count(player_name, name = "rows_captured") %>%
  arrange(desc(rows_captured), player_name)

modern_playoff <- modern_playoff_raw %>%
  group_by(player_id, player_name) %>%
  summarise(
    po_games = sum(po_games, na.rm = TRUE),
    total_points = sum(po_points, na.rm = TRUE),
    total_assists = sum(po_assists, na.rm = TRUE),
    total_rebounds = sum(po_rebounds, na.rm = TRUE),
    playoff_ppg = total_points / po_games,
    playoff_apg = total_assists / po_games,
    playoff_rpg = total_rebounds / po_games,
    .groups = "drop"
  ) %>%
  filter(po_games >= MIN_LEADERBOARD_PO_GAMES)

modern_ppg_leaders <- modern_playoff %>%
  arrange(desc(playoff_ppg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff PPG` = round(playoff_ppg, 1)
  ) %>%
  slice_head(n = 10)

modern_apg_leaders <- modern_playoff %>%
  arrange(desc(playoff_apg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff APG` = round(playoff_apg, 1)
  ) %>%
  slice_head(n = 10)

modern_rpg_leaders <- modern_playoff %>%
  arrange(desc(playoff_rpg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff RPG` = round(playoff_rpg, 1)
  ) %>%
  slice_head(n = 10)

career_translation <- season_level_ppg %>%
  group_by(player_id, player_name) %>%
  summarise(
    career_rs_games = sum(rs_games, na.rm = TRUE),
    career_po_games = sum(po_games, na.rm = TRUE),
    career_rs_points = sum(rs_points, na.rm = TRUE),
    career_po_points = sum(po_points, na.rm = TRUE),
    career_rs_fga = sum(rs_fga, na.rm = TRUE),
    career_po_fga = sum(po_fga, na.rm = TRUE),
    career_rs_fta = sum(rs_fta, na.rm = TRUE),
    career_po_fta = sum(po_fta, na.rm = TRUE),
    career_rs_ppg = career_rs_points / career_rs_games,
    career_po_ppg = career_po_points / career_po_games,
    career_rs_ts = calc_ts(career_rs_points, career_rs_fga, career_rs_fta),
    career_po_ts = calc_ts(career_po_points, career_po_fga, career_po_fta),
    career_rs_rts = weighted.mean(rs_rts, w = rs_games, na.rm = TRUE),
    career_po_rts = weighted.mean(po_rts, w = po_games, na.rm = TRUE),
    career_ppg_diff = career_po_ppg - career_rs_ppg,
    career_pct_change = safe_percent_change(career_po_ppg, career_rs_ppg),
    career_rts_change = career_po_rts - career_rs_rts,
    career_rs_load = weighted.mean(rs_load, w = rs_games, na.rm = TRUE),
    career_po_load = weighted.mean(po_load, w = po_games, na.rm = TRUE),
    career_load_change = career_po_load - career_rs_load,
    career_rs_mpg = weighted.mean(rs_mpg, w = rs_games, na.rm = TRUE),
    career_po_mpg = weighted.mean(po_mpg, w = po_games, na.rm = TRUE),
    career_mpg_change = career_po_mpg - career_rs_mpg,
    .groups = "drop"
  ) %>%
  filter(
    career_rs_games >= MIN_CAREER_RS_GAMES,
    career_po_games >= MIN_CAREER_PO_GAMES
  )

career_filtered <- career_translation %>%
  filter(
    career_po_games >= 50,
    career_rs_mpg >= 20,
    career_po_mpg >= 20
  )

career_risers <- career_filtered %>%
  filter(
    career_ppg_diff >= 1,
    career_po_rts >= 0,
    career_rts_change >= -8
  ) %>%
  arrange(desc(career_ppg_diff), desc(career_po_rts), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

career_maintainers <- career_filtered %>%
  filter(
    dplyr::between(career_ppg_diff, -1, 1.5),
    career_po_rts >= 1,
    career_rts_change >= -2
  ) %>%
  arrange(desc(career_po_ppg), desc(career_po_rts), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

career_droppers <- career_filtered %>%
  filter(
    career_rs_ppg >= 15,
    career_ppg_diff <= -1,
    career_rts_change <= -1.5,
    career_po_rts <= 0
  ) %>%
  arrange(career_ppg_diff, desc(career_rs_ppg), career_rts_change) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

biggest_risers_season <- season_level_filtered %>%
  filter(
    ppg_diff >= 1,
    po_rts >= 0,
    rts_change >= -8
  ) %>%
  arrange(desc(ppg_diff), desc(po_rts), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

biggest_droppers_season <- season_level_filtered %>%
  filter(
    mpg_change >= -5,
    ppg_diff <= -1,
    rts_change <= -1.5,
    po_rts <= 2
  ) %>%
  arrange(ppg_diff, rts_change, po_rts) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

maintainers_season <- season_level_filtered %>%
  filter(
    dplyr::between(ppg_diff, -1, 1),
    po_rts >= 1,
    rts_change >= -2
  ) %>%
  arrange(desc(po_pts_pg), desc(po_rts), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

league_scoring <- read_or_build(
  "league_scoring_modern.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      rs <- tryCatch(pull_team_per_game(season, "Regular Season"), error = function(e) NULL)
      po <- tryCatch(pull_team_per_game(season, "Playoffs"), error = function(e) NULL)

      bind_rows(
        if (!is.null(rs) && nrow(rs) > 0) {
          tibble(
            season = season,
            season_start = as.numeric(substr(season, 1, 4)),
            context = "Regular Season",
            ppg = mean(as.numeric(rs$pts), na.rm = TRUE)
          )
        },
        if (!is.null(po) && nrow(po) > 0) {
          tibble(
            season = season,
            season_start = as.numeric(substr(season, 1, 4)),
            context = "Playoffs",
            ppg = mean(as.numeric(po$pts), na.rm = TRUE)
          )
        }
      )
    }) %>%
      filter(!is.na(ppg))
  }
,
  rebuild = CACHE_REBUILD
)

league_scoring_audit <- league_scoring %>%
  count(season, name = "rows") %>%
  tidyr::complete(
    season = modern_seasons,
    fill = list(rows = 0)
  ) %>%
  arrange(season)

if (VALIDATE_DATA) {
  bad_league_scoring <- league_scoring_audit %>%
    filter(rows != 2)

  if (nrow(bad_league_scoring) > 0) {
    stop(
      sprintf(
        "League scoring is missing a regular-season or playoff row for: %s",
        paste(bad_league_scoring$season, collapse = ", ")
      )
    )
  }
}
```

# Project Overview

The playoffs change everything.

Scoring drops. Possessions slow down. Defenses stop reacting and start targeting. Every weakness gets tested over and over again until something breaks.

That is what makes playoff performance so difficult to evaluate. Raw production still matters, but it does not tell the full story. A player scoring 27 points per game in April is not the same as scoring 27 in June.

This project looks at that gap.

Not just who produces in the playoffs, but whose scoring actually translates when the environment turns against them.

## Definitions

This study focuses on scoring translation: how a player’s playoff scoring compares to their regular-season averages.

### Pretext

Regular-season production is measured only from seasons in which the player reached the playoffs.

This keeps the comparison focused on a player’s playoff-era version of himself, rather than being dragged down by early-career or non-playoff seasons.

### Scoring Change

PPG Difference (Diff) = Playoff PPG – Regular Season PPG

Positive values = increase in scoring. Negative values = decline.

### Efficiency 

rTS Change measures how a player’s scoring efficiency shifts relative to league average from the regular season to the playoffs.

It is calculated as playoff rTS minus regular-season rTS.

A negative value means the player became less efficient relative to the league, even if overall league efficiency also declined. That change should always be read alongside playoff rTS itself.

### Classification

Players are grouped into three categories:

* Risers: scoring increases in the playoffs
* Maintainers: scoring remains within ±1 PPG
* Droppers: scoring decreases

For droppers, results are filtered to remove most obvious role-loss cases. This helps isolate players whose scoring declined despite similar offensive responsibility, making the drop more meaningful.

Riser seasons are interpreted with more context. Some increases come from higher scoring load, while others reflect genuine shotmaking and efficiency gains.

## Playoff Scoring (Regular Season vs Playoffs - Since 2000) 

```{r league-scoring-plot, fig.height=4.7, fig.width=7.5}
ggplot(
  league_scoring,
  aes(x = season_start, y = ppg, color = context)
) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Playoffs" = DROP, "Regular Season" = ACCENT)) +
  scale_x_continuous(breaks = seq(2000, END_SEASON, by = 4)) +
  labs(
    title = "Scoring consistently dips once the playoffs begin",
    subtitle = "League average team scoring by season since 2000-01",
    x = NULL,
    y = "Points per game",
    caption = CAPTION_LINE
  ) +
  theme_caleb()
```

While scoring has increased across the league since the turn of the century, one trend has held constant: production drops in the playoffs.

That gap is what creates the tension of postseason basketball. Some players find ways to hold their level, a few push beyond it, and many fall off once defenses are built entirely to stop them.

In that context, “rising” takes on a different meaning. In a lower-scoring setting, simply maintaining production is already a triumph. True increases are rare, and when they happen, they need to be understood in context.

# Best Playoff Performers of All-Time and of the Modern Era

Before looking at how scoring changes, it helps to establish a baseline: who has actually produced the most in playoff settings.

Because translation only matters if the production exists in the first place.

## All-Time Leaders

These are the most productive playoff performers in NBA history, filtered for players with at least 50 postseason games.

```{r all-time-leaders}
# ------------------------------------------------------------
# All-time playoff production leaders (NBA only)
# Minimum 50 playoff games
# ------------------------------------------------------------

all_time_ppg <- tibble(
  Rank = 1:13,
  Player = c(
    "Michael Jordan", "Luka Dončić*", "Allen Iverson", "Kevin Durant*",
    "Jerry West", "Donovan Mitchell*", "LeBron James*", "Nikola Jokić*",
    "Elgin Baylor", "Giannis Antetokounmpo*", "George Gervin",
    "Stephen Curry*", "Anthony Davis*"
  ),
  GP = c(179, 55, 71, 170, 153, 65, 293, 94, 134, 84, 59, 155, 60),
  PPG = c(33.45, 30.85, 29.73, 29.32, 29.13, 28.38, 28.35, 27.39, 27.04, 27.02, 26.98, 26.75, 26.10)
) %>%
  mutate(PPG = round(PPG, 2))

all_time_rpg <- tibble(
  Rank = 1:13,
  Player = c(
    "Bill Russell", "Wilt Chamberlain", "Wes Unseld", "Bob Pettit",
    "Dave Cowens", "George Mikan", "Moses Malone", "Nate Thurmond",
    "Elvin Hayes", "Elgin Baylor", "Charles Barkley", "Nikola Jokić*",
    "Dolph Schayes"
  ),
  GP = c(165, 160, 119, 88, 89, 60, 94, 81, 96, 134, 123, 94, 97),
  RPG = c(24.87, 24.46, 14.93, 14.82, 14.44, 13.85, 13.78, 13.59, 12.96, 12.87, 12.86, 12.36, 12.22)
) %>%
  mutate(RPG = round(RPG, 2))

all_time_apg <- tibble(
  Rank = 1:13,
  Player = c(
    "Magic Johnson", "John Stockton", "Oscar Robertson", "Kevin Johnson",
    "Isiah Thomas", "Steve Nash", "Bob Cousy", "Rajon Rondo",
    "Chris Paul", "Norm Nixon", "Jason Kidd", "Luka Dončić*", "Nikola Jokić*"
  ),
  GP = c(190, 182, 86, 105, 111, 120, 109, 134, 149, 58, 158, 55, 94),
  APG = c(12.35, 10.10, 8.94, 8.90, 8.89, 8.84, 8.60, 8.48, 8.28, 8.02, 7.99, 7.76, 7.61)
) %>%
  mutate(APG = round(APG, 2))

style_leaderboard <- function(df, caption_text = NULL, align = "c") {
  align_vec <- c("c", "l", "c", "c")
  
  kbl(
    df,
    format = "latex",
    booktabs = TRUE,
    align = align_vec,
    caption = caption_text,
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = FALSE,
      font_size = 9.5,
      stripe_color = "#fbfaf7"
    )
}

style_leaderboard <- function(df, stat_col) {
  df_fmt <- df
  df_fmt[[stat_col]] <- ifelse(
    df_fmt$Rank <= 3,
    cell_spec(df_fmt[[stat_col]], bold = TRUE, format = "latex"),
    as.character(df_fmt[[stat_col]])
  )
  
  df_fmt$Player <- ifelse(
    df_fmt$Rank <= 3,
    cell_spec(df_fmt$Player, bold = TRUE, format = "latex"),
    df_fmt$Player
  )
  
  kbl(
    df_fmt,
    format = "latex",
    escape = FALSE,
    booktabs = TRUE,
    align = c("c", "l", "c", "c"),
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = FALSE,
      font_size = 9.5,
      stripe_color = "#fbfaf7"
    )
}
```

*Minimum 50 playoff games | star = active player | Data: Basketball-Reference*

### Scoring

```{r all-time-ppg}
all_time_ppg %>% style_leaderboard("PPG")
```

### Rebounding

```{r all-time-rpg}
all_time_rpg %>% style_leaderboard("RPG")
```

### Playmaking

```{r all-time-apg}
all_time_apg %>% style_leaderboard("APG")
```

These tables highlight two different ways playoff averages get built: overwhelming dominance and sustained longevity. To narrow the lens, it helps to look at playoff production again in the modern era.

## Modern Leaders

Looking only at the modern game helps remove some of the stylistic and era differences that shape all-time leaderboards.

### Top playoff scorers

```{r ppg-leaders}
modern_ppg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

### Top playoff rebounders

```{r rpg-leaders}
modern_rpg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

### Top playoff playmakers

```{r apg-leaders}
modern_apg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

These tables show who produces.

They do not show whose game survives.

Because playoff basketball is not just about output. It is about what happens when that output is put under pressure.

That is where translation matters.

```{r leaders-for-run}
# ------------------------------------------------------------
# Biggest single-run playoff totals (modern era)
# ------------------------------------------------------------

modern_single_run_raw <- read_or_build(
  "modern_single_run_raw.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      po <- tryCatch(
        pull_player_totals(season, "Playoffs"),
        error = function(e) NULL
      )

      if (is.null(po) || nrow(po) == 0) return(tibble())

      po %>%
        transmute(
          player_id = as.character(player_id),
          player_name,
          season = season,
          team_abbreviation,
          po_games = as.numeric(gp),
          total_points = as.numeric(pts),
          total_rebounds = as.numeric(reb),
          total_assists = as.numeric(ast),
          total_steals = as.numeric(stl),
          total_blocks = as.numeric(blk)
        )
    })
  }
,
  rebuild = CACHE_REBUILD
)

single_run_audit <- audit_season_coverage(modern_single_run_raw)

if (VALIDATE_DATA) {
  check_missing_seasons(modern_single_run_raw, dataset_name = "modern_single_run_raw")
}

single_run_points <- modern_single_run_raw %>%
  arrange(desc(total_points)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Points = total_points
  ) %>%
  slice_head(n = 5)

single_run_rebounds <- modern_single_run_raw %>%
  arrange(desc(total_rebounds)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Rebounds = total_rebounds
  ) %>%
  slice_head(n = 5)

single_run_assists <- modern_single_run_raw %>%
  arrange(desc(total_assists)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Assists = total_assists
  ) %>%
  slice_head(n = 5)

single_run_steals <- modern_single_run_raw %>%
  arrange(desc(total_steals)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Steals = total_steals
  ) %>%
  slice_head(n = 5)

single_run_blocks <- modern_single_run_raw %>%
  arrange(desc(total_blocks)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Blocks = total_blocks
  ) %>%
  slice_head(n = 5)
```

## Biggest Single-Run Playoff Totals

Per-game production captures consistency, but it does not fully reflect the scale of a single postseason run.  
These totals capture the biggest modern playoff workloads in one run.

### Points
```{r single-run-points}
single_run_points %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Rebounds
```{r single-run-rebounds}
single_run_rebounds %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Assists
```{r single-run-assists}
single_run_assists %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Steals
```{r single_run_steals}
single_run_steals%>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Blocks
```{r single_run_blocks}
single_run_blocks %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

# Playoff-Season Baseline Translation

To keep the comparison cleaner, the regular-season side here uses only seasons in which a player actually reached the playoffs. That avoids early-career and non-playoff years from dragging down the baseline and keeps the comparison closer to a player's playoff-era version of himself.

This is not trying to answer who had the bigger career arc. It is trying to answer who held up best once they were actually in playoff environments.

## Largest career playoff scoring lifts

```{r career-risers}
career_risers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

This group is more nuanced than it looks.

Some players here genuinely elevate — increasing both scoring and maintaining strong efficiency. Others are taking on more responsibility, trading a bit of efficiency for volume.

That distinction matters.

A player adding four points per game at the cost of efficiency is not the same as one who maintains or improves it. The table captures both, but they are not equal forms of “rising.”

## Best career maintainers

```{r career-maintainers}
career_maintainers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

This is the most important group in the entire analysis.

Most players decline in the playoffs. That is the baseline.

So maintaining production is not neutral. It is a signal.

These are the players whose scoring holds up even when defenses are built specifically to stop them. That is what real playoff scalability looks like.

## Biggest career droppers

```{r career-droppers}
career_droppers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

Not all drops are the same.

Some players lose role. Others lose efficiency. The table is filtered to remove most obvious role-loss cases, which means what remains is closer to true scoring decline.

These are players whose production fell even when their offensive responsibility stayed relatively stable.

# Since 2000: Who Actually Scales?

Career summaries smooth everything out. Single postseasons show the sharper edges: role changes, matchup advantages, injury context, and the rare cases where a player's game fully scales into playoff basketball.

These tables are noisier than the career ones by nature, but they are also where the clearest single-run examples show up.

## Biggest playoff risers by season

Some of these jumps reflect real postseason scaling. Others come with a bigger scoring load or a slightly shakier efficiency profile. That is why the scoring change and the rTS change need to be read together.

```{r season-risers}
biggest_risers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

## Most important: the maintainers

Most players drop. These players do not.

This is probably the clearest signal in the whole piece. Holding your scoring in the playoffs is already hard. Doing it while keeping load and efficiency reasonably intact is what separates a star regular-season scorer from someone whose game really survives playoff basketball.

```{r season-maintainers}
maintainers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

## Biggest playoff droppers by season

This is the cleaner version of the dropper idea. The table is filtered to strip out most obvious role-loss seasons, so what remains is closer to true scoring slippage: players whose production fell even without a major reduction in scoring load.

```{r season-droppers}
biggest_droppers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```
