---
title: "Scoring Translation in the NBA Playoffs (Since 2000)"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
geometry: margin=0.8in
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(tidyverse)
library(hoopR)
library(janitor)
library(scales)
library(kableExtra)
library(showtext)
library(sysfonts)
library(forcats)

font_add_google("Inter", "inter")
showtext_auto()
options(scipen = 999)

BG <- "#f6f4ef"
TEXT_DARK <- "#141414"
TEXT_LIGHT <- "#6b6b6b"
GRID <- "#dedad3"
ACCENT <- "#1d4ed8"
HILITE <- "#e8f0ff"
RISE <- "#15803d"
DROP <- "#b91c1c"

CAPTION_LINE <- "Data: NBA.com via hoopR | Viz: @Rambzee_"

theme_caleb <- function() {
  theme_minimal(base_family = "inter") +
    theme(
      plot.background = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = TEXT_DARK),
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11.5, color = TEXT_LIGHT, hjust = 0.5),
      axis.title = element_text(size = 11.5, face = "bold"),
      axis.text = element_text(size = 10),
      axis.ticks = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 10),
      plot.caption = element_text(size = 8, color = "#6e6e6e", hjust = 1, margin = margin(t = 8))
    )
}

style_kable <- function(df, align = NULL, font_size = 9.5, full_width = FALSE) {
  if (is.null(align)) {
    align <- strrep("c", ncol(df))
  }

  kbl(
    df,
    booktabs = TRUE,
    align = align,
    format = "latex",
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = full_width,
      font_size = font_size,
      stripe_color = "#fbfaf7"
    )
}

safe_percent_change <- function(new, old) {
  ifelse(is.na(old) | old == 0, NA_real_, 100 * (new - old) / old)
}

calc_ts <- function(pts, fga, fta) {
  denom <- 2 * (fga + 0.44 * fta)
  ifelse(is.na(denom) | denom == 0, NA_real_, pts / denom)
}
```

```{r parameters-and-helpers, include=FALSE}
START_SEASON <- 2000
END_SEASON   <- 2024

MIN_RS_GAMES <- 40
MIN_PO_GAMES <- 8
MIN_RS_PPG   <- 10

MIN_CAREER_RS_GAMES <- 250
MIN_CAREER_PO_GAMES <- 50
MIN_LEADERBOARD_PO_GAMES <- 50

TOP_N <- 10

CACHE_REBUILD <- FALSE
VALIDATE_DATA <- FALSE

season_labels <- function(start_year, end_year) {
  yrs <- start_year:end_year
  paste0(yrs, "-", stringr::str_sub(yrs + 1, 3, 4))
}

modern_seasons <- season_labels(START_SEASON, END_SEASON)

pull_player_per_game <- function(season, season_type = "Regular Season", max_attempts = 3) {
  for (i in seq_len(max_attempts)) {
    out <- tryCatch(
      nba_leaguedashplayerstats(
        season = season,
        season_type = season_type,
        per_mode = "PerGame"
      ),
      error = function(e) NULL
    )

    if (!is.null(out$LeagueDashPlayerStats) && nrow(out$LeagueDashPlayerStats) > 0) {
      return(out$LeagueDashPlayerStats %>% clean_names())
    }

    message(sprintf(
      "EMPTY pull_player_per_game: %s | %s | attempt %s",
      season, season_type, i
    ))

    Sys.sleep(2 * i)
  }

  tibble()
}

pull_player_totals <- function(season, season_type = "Regular Season", max_attempts = 3) {
  for (i in seq_len(max_attempts)) {
    out <- tryCatch(
      nba_leaguedashplayerstats(
        season = season,
        season_type = season_type,
        per_mode = "Totals"
      ),
      error = function(e) NULL
    )

    if (!is.null(out$LeagueDashPlayerStats) && nrow(out$LeagueDashPlayerStats) > 0) {
      return(out$LeagueDashPlayerStats %>% clean_names())
    }

    message(sprintf(
      "EMPTY pull_player_totals: %s | %s | attempt %s",
      season, season_type, i
    ))

    Sys.sleep(2 * i)
  }

  tibble()
}

pull_team_per_game <- function(season, season_type = "Regular Season", max_attempts = 5) {
  for (i in seq_len(max_attempts)) {
    out <- tryCatch(
      nba_leaguedashteamstats(
        season = season,
        season_type = season_type,
        per_mode = "PerGame"
      ),
      error = function(e) NULL
    )

    if (!is.null(out$LeagueDashTeamStats) && nrow(out$LeagueDashTeamStats) > 0) {
      return(out$LeagueDashTeamStats %>% clean_names())
    }

    message(sprintf(
      "EMPTY pull_team_per_game: %s | %s | attempt %s",
      season, season_type, i
    ))

    Sys.sleep(3 * i)
  }

  tibble()
}

league_ts <- tibble(
  season = c(
    "2000-01", "2001-02", "2002-03", "2003-04", "2004-05", "2005-06",
    "2006-07", "2007-08", "2008-09", "2009-10", "2010-11", "2011-12",
    "2012-13", "2013-14", "2014-15", "2015-16", "2016-17", "2017-18",
    "2018-19", "2019-20", "2020-21", "2021-22", "2022-23", "2023-24", "2024-25"
  ),
  league_rs_ts = c(
    51.8, 52.0, 51.9, 51.6, 52.9, 53.6,
    54.1, 54.0, 54.4, 54.3, 54.1, 52.7,
    53.5, 54.1, 53.4, 54.1, 55.2, 55.6,
    56.0, 56.5, 57.2, 56.6, 58.1, 58.0, 57.6
  ),
  league_po_ts = c(
    51.1, 51.4, 52.5, 50.0, 53.4, 54.7,
    53.0, 53.2, 54.4, 54.3, 52.9, 52.0,
    52.9, 54.8, 52.6, 53.4, 56.3, 55.5,
    55.1, 57.4, 57.1, 56.7, 56.6, 56.6, 56.6
  )
)

read_or_build <- function(path, builder, rebuild = FALSE, allow_empty = FALSE) {
  if (file.exists(path) && !rebuild) {
    return(readRDS(path))
  }

  obj <- builder()

  if (!allow_empty) {
    if (is.null(obj)) {
      stop(sprintf("Builder returned NULL for: %s", path))
    }

    if (inherits(obj, "data.frame") && nrow(obj) == 0) {
      stop(sprintf("Builder returned 0 rows for: %s", path))
    }
  }

  saveRDS(obj, path)
  obj
}

audit_season_coverage <- function(df, seasons = modern_seasons) {
  df %>%
    count(season, name = "rows") %>%
    tidyr::complete(season = seasons, fill = list(rows = 0)) %>%
    arrange(season)
}

check_missing_seasons <- function(df, seasons = modern_seasons, dataset_name = deparse(substitute(df))) {
  audit_tbl <- audit_season_coverage(df, seasons)

  missing_tbl <- audit_tbl %>%
    filter(rows == 0)

  if (nrow(missing_tbl) > 0) {
    stop(
      sprintf(
        "Missing season data in %s: %s",
        dataset_name,
        paste(missing_tbl$season, collapse = ', ')
      )
    )
  }

  audit_tbl
}

audit_player_presence <- function(player_search, df = season_level_ppg) {
  df %>%
    filter(
      stringr::str_detect(
        stringr::str_to_lower(player_name),
        stringr::str_to_lower(player_search)
      ) |
        player_id == as.character(player_search)
    ) %>%
    arrange(player_name, season) %>%
    select(
      player_id, player_name, season,
      rs_games, po_games,
      rs_pts_pg, po_pts_pg,
      rs_ts, po_ts
    )
}

get_ppg_split <- function(season, max_attempts = 5) {
  rs <- tibble()
  po <- tibble()

  for (i in seq_len(max_attempts)) {
    if (nrow(rs) == 0) rs <- pull_player_per_game(season, "Regular Season")
    if (nrow(po) == 0) po <- pull_player_per_game(season, "Playoffs")

    if (nrow(rs) > 0 && nrow(po) > 0) break

    message(sprintf(
      "Retrying get_ppg_split for %s | attempt %s | RS rows: %s | PO rows: %s",
      season, i, nrow(rs), nrow(po)
    ))

    Sys.sleep(2 * i)
  }

  if (nrow(rs) == 0 || nrow(po) == 0) {
    stop(sprintf("Missing RS or PO data for season %s", season))
  }

  rs_tbl <- rs %>%
    transmute(
      player_id = as.character(player_id),
      player_name,
      team_abbreviation,
      season = season,
      rs_games = as.numeric(gp),
      rs_mpg = as.numeric(min),
      rs_pts_pg = as.numeric(pts),
      rs_ast_pg = as.numeric(ast),
      rs_reb_pg = as.numeric(reb),
      rs_fga_pg = as.numeric(fga),
      rs_fta_pg = as.numeric(fta),
      rs_points = as.numeric(pts) * as.numeric(gp),
      rs_assists = as.numeric(ast) * as.numeric(gp),
      rs_rebounds = as.numeric(reb) * as.numeric(gp),
      rs_fga = as.numeric(fga) * as.numeric(gp),
      rs_fta = as.numeric(fta) * as.numeric(gp),
      rs_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(rs_ts = calc_ts(rs_points, rs_fga, rs_fta))

  po_tbl <- po %>%
    transmute(
      player_id = as.character(player_id),
      season = season,
      po_games = as.numeric(gp),
      po_mpg = as.numeric(min),
      po_pts_pg = as.numeric(pts),
      po_ast_pg = as.numeric(ast),
      po_reb_pg = as.numeric(reb),
      po_fga_pg = as.numeric(fga),
      po_fta_pg = as.numeric(fta),
      po_points = as.numeric(pts) * as.numeric(gp),
      po_assists = as.numeric(ast) * as.numeric(gp),
      po_rebounds = as.numeric(reb) * as.numeric(gp),
      po_fga = as.numeric(fga) * as.numeric(gp),
      po_fta = as.numeric(fta) * as.numeric(gp),
      po_load = as.numeric(fga) + 0.44 * as.numeric(fta)
    ) %>%
    mutate(po_ts = calc_ts(po_points, po_fga, po_fta))

  rs_tbl %>%
    inner_join(po_tbl, by = c("player_id", "season")) %>%
    left_join(league_ts, by = "season") %>%
    mutate(
      ppg_diff = po_pts_pg - rs_pts_pg,
      apg_diff = po_ast_pg - rs_ast_pg,
      rpg_diff = po_reb_pg - rs_reb_pg,
      pct_change = safe_percent_change(po_pts_pg, rs_pts_pg),
      fga_change = po_fga_pg - rs_fga_pg,
      load_change = po_load - rs_load,
      mpg_change = po_mpg - rs_mpg,
      rs_rts = 100 * (rs_ts - league_rs_ts / 100),
      po_rts = 100 * (po_ts - league_po_ts / 100),
      rts_change = po_rts - rs_rts
    )
}

season_level_ppg <- read_or_build(
  "season_level_ppg.rds",
  function() {
    map_dfr(modern_seasons, function(s) {
      message("Rebuilding season-level split: ", s)

      out <- tryCatch(
        get_ppg_split(s),
        error = function(e) {
          message(sprintf("FAILED get_ppg_split for %s: %s", s, e$message))
          tibble()
        }
      )

      out
    })
  },
  rebuild = CACHE_REBUILD
)

season_level_audit <- audit_season_coverage(season_level_ppg)

season_level_filtered <- season_level_ppg %>%
  filter(
    rs_games >= MIN_RS_GAMES,
    po_games >= MIN_PO_GAMES,
    rs_pts_pg >= MIN_RS_PPG
  )

modern_playoff_raw <- read_or_build(
  "modern_playoff_raw.rds",
  function() {
    season_level_ppg %>%
      transmute(
        season,
        player_id,
        player_name,
        po_games,
        po_points,
        po_assists,
        po_rebounds
      )
  },
  rebuild = CACHE_REBUILD
)

modern_playoff_audit <- modern_playoff_raw %>%
  count(season, name = "rows") %>%
  tidyr::complete(season = modern_seasons, fill = list(rows = 0)) %>%
  arrange(season)

if (VALIDATE_DATA) {
  bad_modern_playoff <- modern_playoff_audit %>%
    filter(rows == 0)

  if (nrow(bad_modern_playoff) > 0) {
    message(
      sprintf(
        "Incomplete cached modern_playoff_raw detected. Rebuilding cache. Missing seasons: %s",
        paste(bad_modern_playoff$season, collapse = ", ")
      )
    )

    modern_playoff_raw <- read_or_build(
      "modern_playoff_raw.rds",
      function() {
        map_dfr(modern_seasons, function(season) {
          message("Modern playoff totals rebuild pull: ", season)

          po <- tryCatch(
            pull_player_totals(season, "Playoffs"),
            error = function(e) NULL
          )

          if (is.null(po) || nrow(po) == 0) {
            return(tibble())
          }

          po %>%
            transmute(
              season = season,
              player_id = as.character(player_id),
              player_name,
              po_games = as.numeric(gp),
              po_points = as.numeric(pts),
              po_assists = as.numeric(ast),
              po_rebounds = as.numeric(reb)
            )
        })
      },
      rebuild = T
    )

    modern_playoff_audit <- modern_playoff_raw %>%
      count(season, name = "rows") %>%
      tidyr::complete(season = modern_seasons, fill = list(rows = 0)) %>%
      arrange(season)
  }

  bad_modern_playoff <- modern_playoff_audit %>%
    filter(rows == 0)

  if (nrow(bad_modern_playoff) > 0) {
    stop(
      sprintf(
        "modern_playoff_raw is missing seasons: %s",
        paste(bad_modern_playoff$season, collapse = ", ")
      )
    )
  }
}

modern_playoff <- modern_playoff_raw %>%
  group_by(player_id, player_name) %>%
  summarise(
    po_games = sum(po_games, na.rm = TRUE),
    total_points = sum(po_points, na.rm = TRUE),
    total_assists = sum(po_assists, na.rm = TRUE),
    total_rebounds = sum(po_rebounds, na.rm = TRUE),
    playoff_ppg = total_points / po_games,
    playoff_apg = total_assists / po_games,
    playoff_rpg = total_rebounds / po_games,
    .groups = "drop"
  ) %>%
  filter(po_games >= MIN_LEADERBOARD_PO_GAMES)

modern_ppg_leaders <- modern_playoff %>%
  arrange(desc(playoff_ppg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff PPG` = round(playoff_ppg, 1)
  ) %>%
  slice_head(n = 10)

modern_apg_leaders <- modern_playoff %>%
  arrange(desc(playoff_apg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff APG` = round(playoff_apg, 1)
  ) %>%
  slice_head(n = 10)

modern_rpg_leaders <- modern_playoff %>%
  arrange(desc(playoff_rpg)) %>%
  transmute(
    Player = player_name,
    `PO G` = po_games,
    `Playoff RPG` = round(playoff_rpg, 1)
  ) %>%
  slice_head(n = 10)

career_translation <- season_level_ppg %>%
  group_by(player_id, player_name) %>%
  summarise(
    career_rs_games = sum(rs_games, na.rm = TRUE),
    career_po_games = sum(po_games, na.rm = TRUE),
    career_rs_points = sum(rs_points, na.rm = TRUE),
    career_po_points = sum(po_points, na.rm = TRUE),
    career_rs_fga = sum(rs_fga, na.rm = TRUE),
    career_po_fga = sum(po_fga, na.rm = TRUE),
    career_rs_fta = sum(rs_fta, na.rm = TRUE),
    career_po_fta = sum(po_fta, na.rm = TRUE),
    career_rs_ppg = career_rs_points / career_rs_games,
    career_po_ppg = career_po_points / career_po_games,
    career_rs_ts = calc_ts(career_rs_points, career_rs_fga, career_rs_fta),
    career_po_ts = calc_ts(career_po_points, career_po_fga, career_po_fta),
    career_rs_rts = weighted.mean(rs_rts, w = rs_games, na.rm = TRUE),
    career_po_rts = weighted.mean(po_rts, w = po_games, na.rm = TRUE),
    career_ppg_diff = career_po_ppg - career_rs_ppg,
    career_pct_change = safe_percent_change(career_po_ppg, career_rs_ppg),
    career_rts_change = career_po_rts - career_rs_rts,
    career_rs_load = weighted.mean(rs_load, w = rs_games, na.rm = TRUE),
    career_po_load = weighted.mean(po_load, w = po_games, na.rm = TRUE),
    career_load_change = career_po_load - career_rs_load,
    career_rs_mpg = weighted.mean(rs_mpg, w = rs_games, na.rm = TRUE),
    career_po_mpg = weighted.mean(po_mpg, w = po_games, na.rm = TRUE),
    career_mpg_change = career_po_mpg - career_rs_mpg,
    .groups = "drop"
  ) %>%
  filter(
    career_rs_games >= MIN_CAREER_RS_GAMES,
    career_po_games >= MIN_CAREER_PO_GAMES
  )

career_filtered <- career_translation %>%
  filter(
    career_po_games >= 50,
    career_rs_mpg >= 20,
    career_po_mpg >= 20
  )

career_risers <- career_filtered %>%
  filter(
    career_ppg_diff >= 1,
    career_po_rts >= 0,
    career_rts_change >= -8
  ) %>%
  arrange(desc(career_ppg_diff), desc(career_po_rts), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = 12)

career_maintainers <- career_filtered %>%
  filter(
    dplyr::between(career_ppg_diff, -1, 1),
    career_po_rts >= 0,
    career_rts_change >= -2
  ) %>%
  arrange(desc(career_po_ppg), desc(career_po_rts), desc(career_rts_change)) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = 12)

career_droppers <- career_filtered %>%
  filter(
    career_rs_ppg >= 14,
    career_ppg_diff <= -1,
    career_rts_change <= -1.5,
    career_po_rts <= 1
  ) %>%
  arrange(career_ppg_diff, desc(career_rs_ppg), career_rts_change) %>%
  transmute(
    Player = player_name,
    `RS G` = career_rs_games,
    `PO G` = career_po_games,
    `RS PPG*` = round(career_rs_ppg, 1),
    `PO PPG` = round(career_po_ppg, 1),
    Diff = round(career_ppg_diff, 1),
    `% Change` = round(career_pct_change, 1),
    `rTS Change` = round(career_rts_change, 1),
    `PO rTS` = round(career_po_rts, 1)
  ) %>%
  slice_head(n = 12)

biggest_risers_season <- season_level_filtered %>%
  filter(
    ppg_diff >= 1,
    po_rts >= 0,
    rts_change >= -8
  ) %>%
  arrange(desc(ppg_diff), desc(po_rts), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

biggest_droppers_season <- season_level_filtered %>%
  filter(
    mpg_change >= -5,
    ppg_diff <= -1,
    rts_change <= -1.5,
    po_rts <= 2
  ) %>%
  arrange(ppg_diff, rts_change, po_rts) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

maintainers_season <- season_level_filtered %>%
  filter(
    dplyr::between(ppg_diff, -1, 1),
    po_rts >= 0,
    rts_change >= -2
  ) %>%
  arrange(desc(po_pts_pg), desc(po_rts), desc(rts_change)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `RS G` = rs_games,
    `PO G` = po_games,
    `RS PPG` = round(rs_pts_pg, 1),
    `PO PPG` = round(po_pts_pg, 1),
    Diff = round(ppg_diff, 1),
    `% Change` = round(pct_change, 1),
    `rTS Change` = round(rts_change, 1),
    `PO rTS` = round(po_rts, 1)
  ) %>%
  slice_head(n = TOP_N)

league_scoring <- read_or_build(
  "league_scoring.rds",
  function() {
    map_dfr(modern_seasons, function(season) {
      message("League scoring pull: ", season)

      rs <- tibble()
      po <- tibble()

      for (i in seq_len(6)) {
        if (nrow(rs) == 0) rs <- pull_team_per_game(season, "Regular Season")
        if (nrow(po) == 0) po <- pull_team_per_game(season, "Playoffs")

        if (nrow(rs) > 0 && nrow(po) > 0) break

        message(sprintf(
          "Retrying league scoring for %s | attempt %s | RS rows: %s | PO rows: %s",
          season, i, nrow(rs), nrow(po)
        ))

        Sys.sleep(2 * i)
      }

      bind_rows(
        if (nrow(rs) > 0) tibble(
          season = season,
          season_start = as.numeric(substr(season, 1, 4)),
          context = "Regular Season",
          ppg = mean(as.numeric(rs$pts), na.rm = TRUE)
        ),
        if (nrow(po) > 0) tibble(
          season = season,
          season_start = as.numeric(substr(season, 1, 4)),
          context = "Playoffs",
          ppg = mean(as.numeric(po$pts), na.rm = TRUE)
        )
      )
    })
  },
  rebuild = CACHE_REBUILD
)

league_scoring_audit <- league_scoring %>%
  count(season, name = "rows") %>%
  tidyr::complete(
    season = modern_seasons,
    fill = list(rows = 0)
  ) %>%
  arrange(season)

if (VALIDATE_DATA) {
  bad_league_scoring <- league_scoring_audit %>%
    filter(rows != 2)

  if (nrow(bad_league_scoring) > 0) {
    message(
      sprintf(
        "Incomplete cached league_scoring detected. Rebuilding cache. Bad seasons: %s",
        paste(bad_league_scoring$season, collapse = ", ")
      )
    )

    league_scoring <- read_or_build(
      "league_scoring.rds",
      function() {
        map_dfr(modern_seasons, function(season) {
          message("League scoring rebuild pull: ", season)

          rs <- pull_team_per_game(season, "Regular Season")
          po <- pull_team_per_game(season, "Playoffs")

          bind_rows(
            if (nrow(rs) > 0) tibble(
              season = season,
              season_start = as.numeric(substr(season, 1, 4)),
              context = "Regular Season",
              ppg = mean(as.numeric(rs$pts), na.rm = TRUE)
            ),
            if (nrow(po) > 0) tibble(
              season = season,
              season_start = as.numeric(substr(season, 1, 4)),
              context = "Playoffs",
              ppg = mean(as.numeric(po$pts), na.rm = TRUE)
            )
          )
        })
      },
      rebuild = T
    )

    league_scoring_audit <- league_scoring %>%
      count(season, name = "rows") %>%
      tidyr::complete(
        season = modern_seasons,
        fill = list(rows = 0)
      ) %>%
      arrange(season)
  }

  bad_league_scoring <- league_scoring_audit %>%
    filter(rows != 2)

  if (nrow(bad_league_scoring) > 0) {
    stop(
      sprintf(
        "League scoring is missing a regular-season or playoff row for: %s",
        paste(bad_league_scoring$season, collapse = ", ")
      )
    )
  }
}
```

# Project Overview

It is a basketball cliché that the playoffs are a different game. But it holds up. Defenses tighten, weak points get hunted, and stars see entire game plans built to make life harder.

A player scoring 27 a night in January is not facing the same game as someone trying to get to 27 in May. The volume might look the same. The difficulty does not.

That is where playoff discourse often misses. Players get labeled as risers or droppers off one stretch, even when the sample is too small to mean much. This looks at the bigger picture: not just who scores, but whose scoring actually translates when defenses are built to take it away.

## Definitions

This analysis focuses on scoring translation: how a player’s playoff scoring compares to their regular-season averages.

### Pretext

Regular-season production is measured only from seasons in which the player reached the playoffs.

This keeps the comparison focused on a player’s playoff-era version of himself, rather than being dragged down by early-career or non-playoff seasons.

### Scoring Change

PPG Difference (Diff) = Playoff PPG – Regular Season PPG

Positive values = increase in scoring. Negative values = decline.

### Efficiency 

rTS Change measures how a player’s scoring efficiency shifts relative to league average from the regular season to the playoffs.

It is calculated as playoff rTS minus regular-season rTS.

A negative value means the player became less efficient relative to the league, even if overall league efficiency also declined. That change should always be read alongside playoff rTS itself.

### Classification

Players are grouped into three categories:

* Risers: scoring increases in the playoffs
* Maintainers: scoring remains within ±1 PPG
* Droppers: scoring decreases

For droppers, results are filtered to remove most obvious role-loss cases. This helps isolate players whose scoring declined despite similar offensive responsibility, making the drop more meaningful.

Riser seasons are interpreted with more context. Some increases come from higher scoring load, while others reflect genuine shotmaking and efficiency gains.

## Playoff Scoring (Regular Season vs Playoffs - Since 2000) 

```{r league-scoring-plot, fig.height=4.7, fig.width=7.5}
ggplot(
  league_scoring,
  aes(x = season_start, y = ppg, color = context)
) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Playoffs" = DROP, "Regular Season" = ACCENT)) +
  scale_x_continuous(breaks = seq(2000, END_SEASON, by = 4)) +
  labs(
    title = "Scoring consistently dips once the playoffs begin",
    subtitle = "League average team scoring by season since 2000-01",
    x = NULL,
    y = "Points per game",
    caption = CAPTION_LINE
  ) +
  theme_caleb()
```

While scoring has climbed across the league since 2000, one trend has stayed the same: offense gets harder in the playoffs.

That changes how production should be read. Holding your scoring is already a win. Pushing beyond it is rare. Falling off is common.

Once the game tightens, some players survive it, a few rise above it, and others get dragged down by the kind of attention and physicality only playoff basketball brings.

# Best Playoff Performers of All-Time and of the Modern Era

Before looking at how scoring changes, it helps to establish who has actually produced the most in playoff settings.

## All-Time Leaders

These are the most productive playoff performers in NBA history, filtered for players with at least 50 postseason games.

```{r all-time-leaders}
# ------------------------------------------------------------
# All-time playoff production leaders (NBA only)
# Minimum 50 playoff games
# ------------------------------------------------------------

all_time_ppg <- tibble(
  Rank = 1:13,
  Player = c(
    "Michael Jordan", "Luka Dončić*", "Allen Iverson", "Kevin Durant*",
    "Jerry West", "Donovan Mitchell*", "LeBron James*", "Nikola Jokić*",
    "Elgin Baylor", "Giannis Antetokounmpo*", "George Gervin",
    "Stephen Curry*", "Anthony Davis*"
  ),
  GP = c(179, 55, 71, 170, 153, 65, 293, 94, 134, 84, 59, 155, 60),
  PPG = c(33.45, 30.85, 29.73, 29.32, 29.13, 28.38, 28.35, 27.39, 27.04, 27.02, 26.98, 26.75, 26.10)
) %>%
  mutate(PPG = round(PPG, 2))

all_time_rpg <- tibble(
  Rank = 1:13,
  Player = c(
    "Bill Russell", "Wilt Chamberlain", "Wes Unseld", "Bob Pettit",
    "Dave Cowens", "George Mikan", "Moses Malone", "Nate Thurmond",
    "Elvin Hayes", "Elgin Baylor", "Charles Barkley", "Nikola Jokić*",
    "Dolph Schayes"
  ),
  GP = c(165, 160, 119, 88, 89, 60, 94, 81, 96, 134, 123, 94, 97),
  RPG = c(24.87, 24.46, 14.93, 14.82, 14.44, 13.85, 13.78, 13.59, 12.96, 12.87, 12.86, 12.36, 12.22)
) %>%
  mutate(RPG = round(RPG, 2))

all_time_apg <- tibble(
  Rank = 1:13,
  Player = c(
    "Magic Johnson", "John Stockton", "Oscar Robertson", "Kevin Johnson",
    "Isiah Thomas", "Steve Nash", "Bob Cousy", "Rajon Rondo",
    "Chris Paul", "Norm Nixon", "Jason Kidd", "Luka Dončić*", "Nikola Jokić*"
  ),
  GP = c(190, 182, 86, 105, 111, 120, 109, 134, 149, 58, 158, 55, 94),
  APG = c(12.35, 10.10, 8.94, 8.90, 8.89, 8.84, 8.60, 8.48, 8.28, 8.02, 7.99, 7.76, 7.61)
) %>%
  mutate(APG = round(APG, 2))

style_leaderboard <- function(df, caption_text = NULL, align = "c") {
  align_vec <- c("c", "l", "c", "c")
  
  kbl(
    df,
    format = "latex",
    booktabs = TRUE,
    align = align_vec,
    caption = caption_text,
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = FALSE,
      font_size = 9.5,
      stripe_color = "#fbfaf7"
    )
}

style_leaderboard <- function(df, stat_col) {
  df_fmt <- df
  df_fmt[[stat_col]] <- ifelse(
    df_fmt$Rank <= 3,
    cell_spec(df_fmt[[stat_col]], bold = TRUE, format = "latex"),
    as.character(df_fmt[[stat_col]])
  )
  
  df_fmt$Player <- ifelse(
    df_fmt$Rank <= 3,
    cell_spec(df_fmt$Player, bold = TRUE, format = "latex"),
    df_fmt$Player
  )
  
  kbl(
    df_fmt,
    format = "latex",
    escape = FALSE,
    booktabs = TRUE,
    align = c("c", "l", "c", "c"),
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "striped"),
      full_width = FALSE,
      font_size = 9.5,
      stripe_color = "#fbfaf7"
    )
}
```

*Minimum 50 playoff games | star = active player | Data: Basketball-Reference*

### Scoring

```{r all-time-ppg}
all_time_ppg %>% style_leaderboard("PPG")
```

### Rebounding

```{r all-time-rpg}
all_time_rpg %>% style_leaderboard("RPG")
```

### Playmaking

```{r all-time-apg}
all_time_apg %>% style_leaderboard("APG")
```

Across eras, the same types of players rise to the top. Scorers who cannot be pushed off their spots. Bigs who control the glass. Primary creators who can still organize offense when everything slows down.

The other thing that stands out is volume. Players like Jordan, LeBron, Magic, and Russell built these numbers over massive playoff samples, essentially adding multiple extra seasons at the highest level.

## Modern Leaders

For all the similarities, there are still real differences across eras. The positions that dominate change. The schemes change. The style of offense changes. And so do some of the names that end up at the top of these leaderboards.

To make the comparison cleaner, it helps to narrow the lens to the modern game. Here are the players who stand out as the playoff leaders of this era.

### Top playoff scorers

```{r ppg-leaders}
modern_ppg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

### Top playoff rebounders

```{r rpg-leaders}
modern_rpg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

### Top playoff playmakers

```{r apg-leaders}
modern_apg_leaders %>%
  style_kable(align = "lcc", font_size = 9.5)
```

The modern leaderboards show who has stood out as the defining playoff producers of the past 25 years.

The names that show up more than once are the biggest story. Luka. Jokić. LeBron. Giannis. Anthony Davis. These are not one-skill playoff stars. They affect the game in too many ways to be cleanly removed from it, most often by pairing elite physical tools with the feel to exploit whatever matchup is in front of them.

That same theme carries over when the lens shifts from per-game production to the biggest single playoff runs.

```{r leaders-for-run}
# ------------------------------------------------------------
# Biggest single-run playoff totals (modern era)
# ------------------------------------------------------------

modern_single_run_raw <- read_or_build(
  "modern_single_run_raw.rds",
  function() {
    totals_pull <- map_dfr(modern_seasons, function(season) {
      po <- tryCatch(
        pull_player_totals(season, "Playoffs"),
        error = function(e) NULL
      )

      if (is.null(po) || nrow(po) == 0) return(tibble())

      po %>%
        transmute(
          player_id = as.character(player_id),
          player_name,
          season = season,
          team_abbreviation,
          po_games = as.numeric(gp),
          total_points = as.numeric(pts),
          total_rebounds = as.numeric(reb),
          total_assists = as.numeric(ast),
          total_steals = as.numeric(stl),
          total_blocks = as.numeric(blk)
        )
    })

    base_from_ppg <- season_level_ppg %>%
      transmute(
        player_id,
        player_name,
        season,
        team_abbreviation,
        po_games,
        total_points = po_points,
        total_rebounds = po_rebounds,
        total_assists = po_assists
      )

    base_from_ppg %>%
      left_join(
        totals_pull %>%
          select(player_id, season, total_steals, total_blocks),
        by = c("player_id", "season")
      ) %>%
      mutate(
        total_steals = coalesce(total_steals, 0),
        total_blocks = coalesce(total_blocks, 0)
      )
  },
  rebuild = CACHE_REBUILD
)

single_run_audit <- audit_season_coverage(modern_single_run_raw)

if (VALIDATE_DATA) {
  check_missing_seasons(modern_single_run_raw, dataset_name = "modern_single_run_raw")
}

single_run_points <- modern_single_run_raw %>%
  arrange(desc(total_points)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Points = total_points
  ) %>%
  slice_head(n = 5)

single_run_rebounds <- modern_single_run_raw %>%
  arrange(desc(total_rebounds)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Rebounds = total_rebounds
  ) %>%
  slice_head(n = 5)

single_run_assists <- modern_single_run_raw %>%
  arrange(desc(total_assists)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Assists = total_assists
  ) %>%
  slice_head(n = 5)

single_run_steals <- modern_single_run_raw %>%
  arrange(desc(total_steals)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Steals = total_steals
  ) %>%
  slice_head(n = 5)

single_run_blocks <- modern_single_run_raw %>%
  arrange(desc(total_blocks)) %>%
  transmute(
    Player = player_name,
    Season = season,
    Team = team_abbreviation,
    `PO G` = po_games,
    Blocks = total_blocks
  ) %>%
  slice_head(n = 5)
```

## Biggest Single-Run Playoff Totals (Since 2000)

These totals capture the biggest modern playoff performances over a single run.

### Points
```{r single-run-points}
single_run_points %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Rebounds
```{r single-run-rebounds}
single_run_rebounds %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Assists
```{r single-run-assists}
single_run_assists %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Steals
```{r single_run_steals}
single_run_steals%>%
  style_kable(align = "lcccc", font_size = 9.2)
```

### Blocks
```{r single_run_blocks}
single_run_blocks %>%
  style_kable(align = "lcccc", font_size = 9.2)
```

Single-run totals show something per-game averages cannot. They capture what it looks like when a player has to sustain elite production over an entire playoff run.

Some of these are burned into playoff history: LeBron dragging Cleveland through the 2018 East, Kawhi carrying Toronto in 2019, Iverson’s absurd 2001 shot diet, Kobe controlling a title run in 2009.

They also show how different dominance can look. Duncan owning the glass and the rim. Dwight overwhelming the paint. Rondo driving deep runs as Boston’s engine.

More than anything, these totals show which players could carry massive responsibility for two months and still hold up.

# Modern Playoff Performers Translation

Beyond averages and single runs, most playoff conversations come down to who rose and who fell. But that only means so much without the right baseline.

Here, regular-season production only comes from seasons where a player actually reached the playoffs, so the comparison stays tied to his real playoff window. This is not a career-value assessment. It is a look at whose scoring still translated once the game got tougher.

## Largest career playoff scoring lifts

```{r career-risers}
career_risers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

This is probably the messiest group of the three, because not every rise means the same thing.

Jimmy is the obvious example. The scoring jump is real, but so is the burden. Once the playoffs start, everything swings further in his direction and the production still climbs.

Jokić is a different kind of riser. The jump is smaller, but it comes while he is still controlling the whole offense and staying elite efficiency-wise. Kawhi fits that mold too. The increase is not huge, but it is on much better efficiency, and that matters more.

Then there are players like Nash, Durant, and Chris Paul. The increases are smaller, but they come while carrying offensive responsibility that usually pulls efficiency down.

That is the line this table is trying to draw. More points does not always mean better playoff translation. Sometimes it just means more shots. The players who really stand out are the ones whose games still scale when the pressure rises.

## Best career maintainers

```{r career-maintainers}
career_maintainers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

This is arguably the most important group. Maintaining may sound boring, but it is one of the hardest things to do in playoff basketball.

Most players drop in the playoffs. Holding your level is not neutral, it is one of the clearest signs your scoring holds up.

Kobe. Wade. AD. Tatum. Players who stayed near their regular-season production even as possessions got harder and defenses loaded up.

## Biggest career droppers

```{r career-droppers}
career_droppers %>%
  style_kable(align = "lcccccccc", font_size = 8.6)
```

This group tells the other side of the story, players whose production consistently fell when expectations and responsibilities rose.

A lot of these names line up with playoff frustrations people already remember. DeRozan in those Cleveland series. Lou Williams in bench-heavy scoring roles that got harder to sustain. Herro in a setting where the weaknesses become easier to target.

That points to the bigger question underneath all of this: what kinds of scoring actually hold up in the playoffs? Some styles survive playoff attention. Others persist as weak points and crack once a defense has a full series to exploit them.

# Modern Single-Run Performers Translation 

Career numbers smooth everything out. Single runs don’t.

They capture what happens when everything clicks, or when everything breaks, over the course of one playoff run.

## Biggest playoff risers by season

```{r season-risers}
biggest_risers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

These are runs where players truly rose to the occasion and played out of their minds. Steve Nash’s ahead-of-his-time scoring explosion in 2005. Duncan trying to drag the Spurs to a title in 2006. Jamal Murray announcing himself with one of the biggest playoff jumps of the modern era. LeBron overwhelming entire runs in 2009 and 2018.

A lot of these are star burdens pushed to the edge. In some cases, the size of the scoring jump is almost a sign that too much was being asked of one player.

Then there are different kinds of rises. Role players becoming lifesavers for a stretch. Reggie Jackson. Mike Bibby. Jameer Nelson. Nick Van Exel. Not full offensive engines, but players who became critical once the postseason demanded more.

That is what makes riser discourse so tricky. Not every rise means the same thing. Some players caught fire for one run. Others reached a level their teams could actually lean on deep into the playoffs. But in every case, this is what a real postseason jump looks like when it happens all at once.

## Most important: the maintainers

```{r season-maintainers}
maintainers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

Looking at the single-run maintainers, this is where a lot of the all-time great postseason runs reside. Not necessarily players dragging an offense by themselves, but stars taking on bigger burdens and still scaling to them.

Kobe in 2001. LeBron in multiple years. Durant. Curry. Dirk. These runs now read as part of their case for greatness. The scoring load was already huge, the playoff attention was already there, and they still held their level through deep runs.

That is what makes these performances so impressive. There is no surprise factor here. No one was catching defenses off guard. These were stars facing full playoff game plans and still getting to their numbers anyway.

That is also where a lot of playoff discourse gets warped. A player does not need some giant scoring spike to have had an incredible postseason. Sometimes the most impressive thing is just holding your level while everything around you gets harder.

## Biggest playoff droppers by season

```{r season-droppers}
biggest_droppers_season %>%
  style_kable(align = "lcccccccccc", font_size = 8.1)
```

On the flip side, this is where some of the most damaging runs live. The kinds of playoff performances that follow players around for years, and for some unlucky names here, show up more than once.

Drops happen for a lot of reasons. Injuries. Bad shooting luck. Tough matchups. Defensive game plans finally catching up. But the worst runs usually expose something deeper. A scoring style that was always unstable. A shot diet that gets harder to survive on. Free-throw reliance that dries up. Regular-season advantages that disappear once every possession is being played with real force and attention.

That is what makes these runs different. They are not just bad stretches. They are often the point where a player’s scoring style stops translating altogether.

And that is why they stick. Not just because the numbers cratered, but because the playoffs made the weakness impossible to ignore.

# Conclusion

The playoffs do not create a player out of nowhere. More often, they reveal what was already there.

They show the value of scalable scoring, dependable ballhandling, and players who can settle a game when everything gets tighter. They can change the way a career is talked about, separating empty production from the kinds of offensive games that actually hold up in May and June.

That is why playoff translation matters. Not because every drop is an indictment or every rise is proof of greatness, but because the postseason is where scoring styles get tested the hardest and judged the clearest.
