---
title: "Michael Porter Jr.: Offensive Value, Constraints, and Warriors Fit"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{placeins}
---

```{r setup, include=FALSE}
# Chunk defaults
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.pos = "H",
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.6,
  dpi = 300,
  out.width = "100%"
)

# Libraries
library(dplyr); library(tidyr); library(tibble)
library(stringr); library(readr); library(purrr)
library(ggplot2); library(scales); library(knitr)
library(kableExtra); library(janitor); library(forcats)
library(showtext); library(sysfonts); library(readxl)

# Fonts
font_add_google("Poppins", "poppins")
showtext_auto()

# Theme
theme_caleb <- function()
  {
  theme_minimal(base_family = "poppins") +
    theme(
      plot.background  = element_rect(fill = "#f6f4ef", color = NA),
      panel.background = element_rect(fill = "#f6f4ef", color = NA),
      panel.grid.major = element_line(color = "#dedad3", linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title       = element_text(size = 18, face = "bold", hjust = 0.5, color = "#141414"),
      plot.subtitle    = element_text(size = 10, hjust = 0.5, color = "#2f2f2f", margin = margin(b = 8)),
      axis.title       = element_text(size = 12, face = "bold", color = "#141414"),
      axis.text        = element_text(size = 10, color = "#333333"),
      plot.caption     = element_text(size = 8, color = "#6e6e6e", hjust = 0, margin = margin(t = 8)),
      plot.margin      = margin(10, 12, 14, 12),
      legend.position  = "top"
    )
  }

# Colors (Nets)
NETS_BLACK <- "#000000"
NETS_WHITE <- "#FFFFFF"
NETS_GRAY  <- "#8A8D8F"
NETS_DARK  <- "#2C2C2C"

COL_RED   <- NETS_BLACK
COL_BLACK <- NETS_BLACK
COL_SILV  <- NETS_GRAY
COL_GOLD  <- NETS_DARK

# Numeric helpers
to_num <- function(x) readr::parse_number(as.character(x))

as_pct01 <- function(x)
  {
  z <- to_num(x)
  dplyr::case_when(
    is.na(z) ~ NA_real_,
    z > 1 ~ z / 100,
    TRUE ~ z
  )
  }

pct01 <- as_pct01
```

```{r paths}
# File paths
crafted_path <- "MPJ_CraftedNBA_OffAndDef_2026.xlsx"
pt_path      <- "MPJ_Playtype_Eff_2026.xlsx"
shoot_path   <- "MPJ_Shooting_Career.xlsx"
```

```{r load-data}
# Helper: read first sheet safely
read_first_sheet <- function(path)
  {
  sheets <- readxl::excel_sheets(path)
  readxl::read_excel(path, sheet = sheets[1])
  }

# Crafted metrics 
crafted_raw <- read_first_sheet(crafted_path)

# Play-type efficiency
pt_sheets <- readxl::excel_sheets(pt_path)
pt_sheet  <- if ("MPJ_Playtype_Eff_2026" %in% pt_sheets)
  {
  "MPJ_Playtype_Eff_2026"
  } else
    {
  pt_sheets[1]
      }

pt_raw <- readxl::read_excel(pt_path, sheet = pt_sheet)

# Career shooting data
shoot_sheets <- readxl::excel_sheets(shoot_path)
shoot_sheet  <- if ("Career Shooting Numbers" %in% shoot_sheets)
  {
  "Career Shooting Numbers"
  } else
  {
  shoot_sheets[1]
    }

shoot_raw <- readxl::read_excel(shoot_path, sheet = shoot_sheet)
```

# Executive Summary
Entering the season, Michael Porter Jr. was widely viewed as a scorer whose value depended heavily on elite creation around him. Many questioned whether his production in Denver was inseparable from Nikola Jokić, while others pointed to injuries and limited self-creation as factors that would cap his impact in an expanded role. Few expected him to thrive as a primary offensive option in Brooklyn.

Instead, Porter has exceeded those expectations across the board. Through 29 games this season, he is averaging 25.9 points, 7.4 rebounds, and 3.4 assists per game, all career highs, while shooting 49% from the field, 40% from three, and 83% from the line (62.4% TS). This production has not come through inefficient volume or short-term variance, but through sustained shot-making at scale.

Porter has proven to be a legitimate high-volume, high-efficiency wing scorer whose offensive value translates beyond a Jokic-centered system. While he is not a primary offensive engine in the traditional sense, his shooting gravity, movement scoring, and ability to get clean looks regardless of coverage make him an offensive amplifier in virtually any offensive environment.

# Player Profile (2025–26 snapshot)
```{r roster-inputs}
# Brooklyn Nets (2025–26 rotation)
nets_2026 <- tibble::tribble(
  ~Player,               ~Pos, ~MP,  ~`3PA`, ~`3P%`, ~PTS, ~AST, ~TRB,
  "Michael Porter Jr.",  "SF", 33.1,   9.4,   0.397, 25.9,  3.4,  7.4,
  "Nic Claxton",         "C",  29.9,   0.4,   0.167, 13.3,  4.1,  7.7,
  "Noah Clowney",        "PF", 28.4,   6.5,   0.325, 13.1,  1.9,  3.9,
  "Terance Mann",        "SG", 26.2,   2.6,   0.313,  8.0,  3.8,  3.5,
  "Cam Thomas",          "SG", 25.4,   5.4,   0.355, 19.6,  2.4,  1.8,
  "Egor Dёmin",          "PG", 24.7,   6.2,   0.389, 10.5,  3.5,  3.3,
  "Ziaire Williams",     "SF", 22.6,   5.1,   0.313,  9.1,  0.7,  2.4,
  "Tyrese Martin",       "SG", 19.5,   4.2,   0.322,  7.4,  2.2,  3.0,
  "Danny Wolf",          "PF", 19.3,   3.0,   0.355,  7.5,  1.9,  4.0,
  "Drake Powell",        "SG", 18.1,   2.1,   0.327,  5.8,  1.7,  1.6,
  "Day'Ron Sharpe",      "C",  17.6,   0.8,   0.214,  7.5,  2.4,  5.9,
  "Nolan Traore",        "PG", 15.7,   2.3,   0.324,  4.3,  2.1,  1.2
)

# Golden State Warriors (2025–26 rotation)
warriors_2026 <- tibble::tribble(
  ~Player,                ~Pos, ~MP,  ~`3PA`, ~`3P%`, ~PTS, ~AST, ~TRB,
  "Stephen Curry",        "PG", 32.2,  12.1,   0.394, 28.7,  4.7,  3.8,
  "Jimmy Butler",         "SF", 31.4,   2.1,   0.384, 19.6,  4.9,  5.5,
  "Brandin Podziemski",   "SG", 27.1,   4.5,   0.395, 11.9,  3.3,  4.5,
  "Draymond Green",       "PF", 27.0,   4.4,   0.340,  8.7,  5.5,  5.8,
  "De'Anthony Melton",    "PG", 20.5,   4.3,   0.300,  9.7,  1.7,  2.4,
  "Al Horford",           "C",  19.4,   4.4,   0.352,  6.9,  1.8,  4.8,
  "Will Richard",         "SG", 19.0,   3.1,   0.383,  7.5,  1.1,  2.7,
  "Quinten Post",         "PF", 18.2,   4.5,   0.343,  8.1,  1.5,  3.9,
  "Buddy Hield",          "SG", 17.4,   4.4,   0.329,  7.5,  1.4,  2.5,
  "Pat Spencer",          "PG", 15.1,   1.8,   0.364,  6.1,  3.2,  2.2,
  "Trayce Jackson-Davis", "C",  12.9,   0.0,   NA,     4.6,  1.0,  3.9,
  "Gary Payton II",       "SG", 12.4,   1.1,   0.275,  4.9,  1.7,  3.3
)
```

```{r roster-tables}
# Roster summary table
knitr::kable(
  nets_2026,
  format   = "latex",
  booktabs = TRUE,
  caption  = "Brooklyn Nets 2026 production snapshot"
) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "HOLD_position", "scale_down"),
    font_size = 8
  )
```

This season, Michael Porter Jr. has played for an underwhelming Brooklyn Nets team largely focused on developing young talent. Despite that context, MPJ has emerged as the team’s only consistent source of offense. In his 961 minutes on the floor, the Nets post a +0.6 net rating (118.3 ORTG, 117.7 DRTG). With him off the court, they crater to –10.7 (105.5 ORTG, 116.2 DRTG).

Porter’s presence lifts team efficiency by roughly +5 percentage points in true shooting, reduces turnover rate, and stabilizes offensive execution without imposing a defensive tax. He is not carrying the offense through on-ball creation, but he is providing a consistent source of shot-making. 

# Offensive Value
```{r crafted-clean-and-off}
# Standardize column names 
crafted <- crafted_raw %>%
  janitor::clean_names()

crafted_cols <- names(crafted)

# Identify the "Stat" label column
stat_col <- dplyr::case_when(
  "unnamed_0" %in% crafted_cols ~ "unnamed_0",
  TRUE ~ crafted_cols[1]
)

# Identify value + percentile columns (fallbacks)
value_col <- dplyr::case_when(
  "value" %in% crafted_cols ~ "value",
  "val"   %in% crafted_cols ~ "val",
  TRUE ~ NA_character_
)

pctile_col <- dplyr::case_when(
  "percentile" %in% crafted_cols ~ "percentile",
  "pctile"     %in% crafted_cols ~ "pctile",
  TRUE ~ NA_character_
)

# Helper: pull a column if it exists, else NA
col_or_na <- function(df, nm) {
  if (!is.na(nm) && nm %in% names(df)) df[[nm]] else NA
}

# Offense slice 
crafted_off <- crafted %>%
  dplyr::filter(if ("off_stat" %in% crafted_cols) off_stat == 1 else TRUE) %>%
  dplyr::transmute(
    Stat       = .data[[stat_col]],
    Value_num  = to_num(col_or_na(., value_col)),
    Percentile = to_num(col_or_na(., pctile_col))
  ) %>%
  dplyr::mutate(
    Percentile = ifelse(is.na(Percentile), NA_integer_, as.integer(round(Percentile))),
    Value = dplyr::case_when(
      stringr::str_detect(Stat, "TS%|3PAr|cTOV%|Rate|%") ~ scales::percent(Value_num / 100, accuracy = 0.1),
      TRUE ~ format(round(Value_num, 1), nsmall = 1)
    )
  ) %>%
  dplyr::select(Stat, Value, Percentile)

# Defense slice
crafted_def <- crafted %>%
  dplyr::filter(if ("def_stat" %in% crafted_cols) def_stat == 1 else TRUE) %>%
  dplyr::transmute(
    Stat       = .data[[stat_col]],
    Value_num  = to_num(col_or_na(., value_col)),
    Percentile = to_num(col_or_na(., pctile_col))
  ) %>%
  dplyr::mutate(
    Percentile = ifelse(is.na(Percentile), NA_integer_, as.integer(round(Percentile))),
    Value = dplyr::case_when(
      stringr::str_detect(Stat, "Rate|%") ~ scales::percent(Value_num / 100, accuracy = 0.1),
      TRUE ~ format(round(Value_num, 1), nsmall = 1)
    )
  ) %>%
  dplyr::select(Stat, Value, Percentile)
```

## Synergy: Play Types

\FloatBarrier

```{r playtype-clean}
# Play-type table (clean + standardize + format)
pt <- pt_raw %>%
  janitor::clean_names()

# Helpers: pick best-matching column name
pick_col <- function(df, candidates) {
  hit <- intersect(candidates, names(df))
  if (length(hit) == 0) {
    stop("Missing expected column. Tried: ", paste(candidates, collapse = ", "))
  }
  hit[1]
}

play_col <- pick_col(pt, c("stat", "play", "play_type", "playtype"))
poss_col <- pick_col(pt, c("poss", "possessions"))
time_col <- pick_col(pt, c("percent_time", "pct_time", "time", "pct"))
ppp_col  <- pick_col(pt, c("ppp"))
tov_col  <- pick_col(pt, c("to_percent", "to_pct", "tov_pct", "tov", "to"))

# Standardize numerics (one pass)
pt_std <- pt %>%
  transmute(
    Play     = stringr::str_squish(as.character(.data[[play_col]])),
    Poss     = to_num(.data[[poss_col]]),
    pct_time = as_pct01(.data[[time_col]]),
    PPP      = to_num(.data[[ppp_col]]),
    TO       = as_pct01(.data[[tov_col]])
  )

# Format for display
pt_tbl <- pt_std %>%
  arrange(dplyr::desc(pct_time)) %>%
  transmute(
    Play,
    Poss,
    `%Time` = scales::percent(pct_time, accuracy = 0.1),
    PPP     = round(PPP, 3),
    `TO%`   = scales::percent(TO, accuracy = 0.1)
  )
```

```{r playtype-table}
kbl(pt_tbl,
    format = "latex",
    booktabs = TRUE,
    caption = "Michael Porter Jr. Play-Type Profile (Synergy, 2025–26)"
) %>%
  kable_styling(latex_options = c("striped","hold_position","scale_down"), font_size = 9) %>%
  column_spec(1, bold = TRUE)
```

```{r playtype-usage-plot, fig.cap="MPJ play-type usage share (2025–26)."}
# Color palette by play type
pt_cols_var <- c(
  "Spot Up"    = "#000000",
  "Handoffs"   = "#1A1A1A",
  "Transition" = "#3A3A3A",
  "Off Screen" = "#8B9096",
  "Cut"        = "#c6cfd4"
)

# Prep data for plotting
pt_plot <- pt_tbl %>%
  mutate(
    pct_time = readr::parse_number(`Avg %Time`) / 100
  ) %>%
  arrange(desc(pct_time)) %>%
  mutate(
    Play  = factor(Play, levels = Play),
    label = scales::percent(pct_time, accuracy = 0.1)
  )

# Plot: play-type usage share
ggplot(pt_plot, aes(x = Play, y = pct_time, fill = Play)) +
  geom_col(width = 0.68, alpha = 0.95) +
  geom_text(
    aes(label = label),
    vjust = -0.35,
    size = 3.6,
    family = "poppins",
    color = "#141414"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.12))
  ) +
  scale_fill_manual(values = pt_cols_var) +
  labs(
    title = "Where MPJ Gets His Possessions",
    subtitle = "Top five play types by usage share (best actions)",
    x = NULL,
    y = "% of tracked possessions",
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb() +
  theme(
    legend.position = "none",
    plot.title = element_text(color = COL_RED),
    axis.text.x = element_text(size = 11)
  )
```

Porter Jr.’s play-type diet (Figure 1) is exactly what you would expect from a high-value off-ball scorer. His usage is concentrated in play-finishing actions, with the bulk of his offense coming from spot-ups and handoffs, supplemented by off-screen movement and transition opportunities. He generates over 1.0 PPP across all of his most-used play types, ranking as an elite transition scorer (78th percentile) and off-screen threat (81st percentile). Across his primary actions, he consistently grades in the upper echelon (60th percentile or better), underscoring both efficiency and portability rather than scheme-dependent production.

```{r shooting-summary}
# Clean raw shooting sheet: normalize column names
shoot <- shoot_raw %>%
  rename_with(~ gsub("\\.0$", "", .x)) %>%
  janitor::clean_names()

# Standardize season column names you want to use downstream
shoot_tbl <- shoot %>%
  rename(
    s2019_20 = x2020,
    s2020_21 = x2021,
    s2021_22 = x2022,
    s2022_23 = x2023,
    s2023_24 = x2024,
    s2024_25 = x2025,
    s2025_26 = x2026
  )

# Format values for display
shoot_tbl_clean <- shoot_tbl %>%
  mutate(is_fg_pct_row = stringr::str_detect(stringr::str_to_lower(year), "fg%")) %>%
  mutate(
    across(
      where(is.numeric),
      ~ if_else(is_fg_pct_row, scales::number(round(.x, 1), accuracy = 0.1),
                scales::number(round(.x, 0), accuracy = 1))
    )
  ) %>%
  select(-is_fg_pct_row)

# FG% table only, with a clean "Zone" label
fg_tbl <- shoot_tbl_clean %>%
  filter(stringr::str_detect(stringr::str_to_lower(year), "fg%")) %>%
  mutate(
    Zone = stringr::str_trim(stringr::str_remove(year, "\\s*fg%"))
  ) %>%
  select(
    Zone,
    s2019_20, s2020_21, s2021_22, s2022_23, s2023_24, s2024_25, s2025_26
  )
```

## Shooting Profile (Career & 2025-26) 

\FloatBarrier

```{r shooting-head-table}
kableExtra::kbl(
  fg_tbl,
  format = "latex",
  booktabs = TRUE,
  caption = "MPJ Shooting Efficiency by Zone (FG\\%, Career by Season)",
  col.names = c(
    "Zone",
    "19--20","20--21","21--22","22--23","23--24","24--25","25--26"
  ),
  align = "lrrrrrrr"
) %>%
  kableExtra::kable_styling(
    latex_options = c("striped","HOLD_position","scale_down"),
    font_size = 8
  ) %>%
  kableExtra::column_spec(1, bold = TRUE) 
```

```{r mpj-zone-fg-trends, fig.height=4.8, fig.width=7.2, fig.cap="MPJ: Zone FG% trends by season (dashed = career baseline for that zone)."}
# Numeric version for calculations
shoot_tbl_num <- shoot_tbl %>%
  mutate(across(starts_with("s"), ~ suppressWarnings(as.numeric(.x))))

# Helper: grab one labeled row as a 1-row tibble of s* columns
get_row <- function(lbl)
  {
  shoot_tbl_num %>%
    filter(stringr::str_to_lower(year) == stringr::str_to_lower(lbl)) %>%
    select(starts_with("s")) %>%
    slice(1)
  }

# rows needed
ra_fgm      <- get_row("RA FGM")
ra_fga      <- get_row("RA FGA")

nonra_fgm   <- get_row("NON-RA PAINT FGM")
nonra_fga   <- get_row("NON-RA PAINT FGA")

mid_fgm     <- get_row("MIDRANGE FGM")
mid_fga     <- get_row("MIDRANGE FGA")

corner_fgm  <- get_row("CORNER 3 FGM")
corner_fga  <- get_row("CORNER 3 FGA")

atb_fgm     <- get_row("ATB3 FGM")
atb_fga     <- get_row("ATB3 FGA")

# Helpers: season ordering and display labels
seasons <- names(ra_fgm)

season_key <- function(s)
  {
  # s2019_20 -> 2019
  as.integer(stringr::str_extract(s, "\\d{4}"))
  }

season_lab <- function(s)
  {
  # s2020_21 -> '20–21
  x <- stringr::str_remove(s, "^s")
  x <- stringr::str_replace(x, "_", "–")
  paste0("'", stringr::str_sub(x, 3))
  }

# Build wide FG% table (safe division)
num <- function(x) as.numeric(x)

fg_wide <- tibble(
  season_id = season_key(seasons),
  season    = season_lab(seasons),

  `Corner 3` = num(corner_fgm[1, ]) / num(corner_fga[1, ]),
  `ATB 3`    = num(atb_fgm[1, ])    / num(atb_fga[1, ]),
  `Overall 3`= (num(corner_fgm[1, ]) + num(atb_fgm[1, ])) /
               (num(corner_fga[1, ]) + num(atb_fga[1, ])),

  `Rim (RA)`      = num(ra_fgm[1, ])    / num(ra_fga[1, ]),
  `Non-RA Paint`  = num(nonra_fgm[1, ]) / num(nonra_fga[1, ]),
  `Midrange`      = num(mid_fgm[1, ])   / num(mid_fga[1, ])
)

# Long format + baselines
fg_long <- fg_wide %>%
  pivot_longer(
    cols = -c(season_id, season),
    names_to = "Zone",
    values_to = "fg"
  ) %>%
  filter(season_id != 2021) %>%
  arrange(season_id) %>%
  group_by(Zone) %>%
  mutate(baseline = mean(fg, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    season = factor(season, levels = unique(season)),
    Zone   = factor(
      Zone,
      levels = c("Corner 3","ATB 3","Overall 3","Rim (RA)","Non-RA Paint","Midrange")
    ),
    team = if_else(season_id <= 2024, "Nuggets", "Nets")  # adjust cutoff if needed
  )

# Plot
ggplot(fg_long, aes(x = season, y = fg, group = 1)) +
  geom_hline(
    aes(yintercept = baseline),
    linetype = "dashed",
    color = "#8B9096",
    linewidth = 0.6
  ) +
  geom_line(color = "#000000", linewidth = 1.05) +
  geom_point(aes(color = team), size = 2.4) +
  facet_wrap(~Zone, ncol = 3, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = c("Nuggets" = "#FEC524", "Nets" = "#000000"), guide = "none") +
  labs(
    title = "MPJ: Zone FG% Trends",
    subtitle = "Dashed line = career baseline for that zone (excluded season removed)",
    x = NULL,
    y = "Field Goal Percentage",
    caption = "Source: NBA.com | Viz: @Rambzee_"
  ) +
  theme_caleb() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 5),
    strip.text = element_text(face = "bold", color = "#141414"),
    plot.title = element_text(color = "#000000")
  )
```

MPJ’s shooting efficiency is broadly in line with the rest of his career. He’s having an excellent finishing season, converting 79% of his shots in the restricted area. The biggest red flag is corner shooting: he’s at just 16% on left-corner threes and 31% on right-corner threes, which would both be career lows. That said, he has only 42 total corner attempts to this point, and many of those looks have come with more degree of difficulty than what he typically saw in Denver. Overall, he’s still hitting 39% from three on the season, which remains an elite mark.

# Strengths / Tradeoffs
```{r crafted-percentile-plot, fig.height=3.8, fig.width=7.0, fig.cap="CraftedNBA percentiles (Offense vs Defense)."}
# Combine into one plot-ready dataset
crafted_plot_df <- dplyr::bind_rows(
  crafted_off %>% dplyr::mutate(side = "Offense"),
  crafted_def %>% dplyr::mutate(side = "Defense")
) %>%
  dplyr::mutate(
    Percentile = as.numeric(Percentile)
  )

# Tier bins for color mapping
crafted_plot_df <- crafted_plot_df %>%
  dplyr::mutate(
    tier = dplyr::case_when(
      Percentile >= 97 ~ "97-100",
      Percentile >= 93 ~ "93-96",
      Percentile >= 88 ~ "88-92",
      Percentile >= 80 ~ "80-87",
      Percentile >= 70 ~ "70-79",
      Percentile >= 60 ~ "60-69",
      Percentile >= 50 ~ "50-59",
      Percentile >= 30 ~ "30-49",
      TRUE             ~ "<30"
    ),
    tier = factor(
      tier,
      levels = c("97-100","93-96","88-92","80-87","70-79","60-69","50-59","30-49","<30")
    )
  )

# Color palette by tier
pct_cols <- c(
  "97-100" = "#0B4F2D",
  "93-96"  = "#146B3A",
  "88-92"  = "#1F8F4C",
  "80-87"  = "#43A95C",
  "70-79"  = "#6FC16F",
  "60-69"  = "#9EDC7D",
  "50-59"  = "#E6E68A",
  "30-49"  = "#E2B07A",
  "<30"    = "#C26A5A"
)

# Plot one side at a time
plot_one_side <- function(df, side_name) {

  df2 <- df %>%
    dplyr::filter(side == side_name) %>%
    dplyr::arrange(Percentile) %>%
    dplyr::mutate(Stat = factor(Stat, levels = unique(Stat)))  # lock order low -> high

  ggplot(df2, aes(x = Percentile, y = Stat, fill = tier)) +
    geom_col(alpha = 0.95) +
    geom_text(
      aes(label = Percentile),
      hjust = -0.2,
      size = 3.1,
      color = "#2B2B2B",
      family = "poppins"
    ) +
    geom_vline(
      xintercept = 50,
      linetype = "dashed",
      color = "#CFCFCF",
      linewidth = 0.4
    ) +
    scale_fill_manual(values = pct_cols, drop = FALSE) +
    scale_x_continuous(limits = c(0, 110), breaks = seq(0, 100, 20)) +
    labs(
      title = paste0("MPJ: CraftedNBA Percentiles (", side_name, ")"),
      subtitle = "Percentile shading reflects relative value",
      x = "Percentile",
      y = NULL,
      fill = NULL,
      caption = "Source: CraftedNBA | Viz: @Rambzee_"
    ) +
    theme_caleb() +
    theme(legend.position = "none")
}

# Build and print both plots
p_off <- plot_one_side(crafted_plot_df, "Offense")
p_def <- plot_one_side(crafted_plot_df, "Defense")

p_off
p_def


```

CraftedNBA context (useful indicators, not gospel) reinforces just how strong Michael Porter Jr.’s offensive profile is. He grades out in the 80th percentile or better across every major play-finishing and portability metric, underscoring his value as a high-level off-ball scorer. Most notably, he sits in the 84th percentile in true shooting (TS%) while carrying a high offensive load (82nd percentile), a difficult efficiency–volume combo.

While he’s averaging a career-high 3.4 assists, Porter is still a below-average passer overall, which is entirely acceptable given his role. His value is not in creation for others, but in elite shot-making and spacing gravity as a complementary offensive weapon.

Defensively, MPJ has been a serviceable rather than impactful. He shows decent versatility and rim protection, but ranks near the bottom of the league in overall defensive impact (18th percentile in CraftedDPM). Context matters, though: Brooklyn’s defensive infrastructure has been poor, which likely suppresses his on/off and impact-based metrics. In a more structured team-defense environment, Porter projects as a competent defender rather than a liability.

# Takeaways
## What MPJ Brings
* Elite off-ball scoring. High-volume, high-efficiency shooter across spot-ups, handoffs, off-screens, and transition.
* Spacing gravity. Defenses stay attached without the ball, creating advantage by presence alone.
* Scalable offense. Efficiency holds or improves next to better creators.

## What MPJ Needs
* Primary offensive initiation to set the table and shift defenses.
* Ball movement and pace to maximize movement shooting.
* High-quality screening to consistently access handoffs and off-screen actions.
* Strong point-of-attack defense to keep him in a team-defense role rather than stress matchups.

# Team Fit: Why MPJ Is a Near-Perfect Match in Golden State
```{r warriors}
kbl(warriors_2026, format="latex", 
    booktabs=TRUE,
    caption="Golden State Warriors 2026 production snapshot") %>% 
  kable_styling(latex_options=c("striped","hold_position","scale_down"), font_size=8)
```

```{r 3pt-context-plot, fig.cap="3-point volume vs accuracy: Nets vs Warriors."}
# Combine team data + enforce numeric types
ctx <- dplyr::bind_rows(
  nets_2026 %>% dplyr::mutate(Team = "Nets"),
  warriors_2026 %>% dplyr::mutate(Team = "Warriors")
) %>%
  dplyr::mutate(
    `3P%` = as.numeric(`3P%`),
    `3PA` = as.numeric(`3PA`)
  )

# Player ID lookup from NBA Stats (just needs to cover these players)
players <- hoopR::nba_leaguedashplayerstats(
  season = "2025-26",
  season_type = "Regular Season",
  measure_type = "Base"
)$LeagueDashPlayerStats %>%
  janitor::clean_names() %>%
  dplyr::transmute(
    player_id   = as.character(player_id),
    player_name = as.character(player_name)
  ) %>%
  dplyr::distinct()

# Name key for joining (strip punctuation & suffixes, lower-case)
name_key <- function(x)
  {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[[:punct:]]", " ") %>%
    stringr::str_replace_all("\\b(jr|sr|ii|iii|iv|v)\\b", "") %>%
    stringr::str_squish()
  }

players_keyed <- players %>%
  dplyr::mutate(name_key = name_key(player_name))

ctx_img <- ctx %>%
  dplyr::mutate(name_key = name_key(Player)) %>%
  dplyr::left_join(
    players_keyed %>% dplyr::select(name_key, player_id),
    by = "name_key"
  ) %>%
  dplyr::mutate(
    headshot = dplyr::if_else(
      !is.na(player_id),
      paste0("https://cdn.nba.com/headshots/nba/latest/1040x760/", player_id, ".png"),
      NA_character_
    )
  )

# Plot
ggplot2::ggplot(ctx_img) +
  ggimage::geom_image(
    ggplot2::aes(x = `3PA`, y = `3P%`, image = headshot),
    size = 0.09,
    asp = 1.1
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::labs(
    title = "Shooting Environment Comparison",
    subtitle = "Nets vs Warriors",
    x = "3PA per game",
    y = "3P%",
    caption = "Source: BBall Reference & NBA headshots | Viz: @Rambzee_"
  ) +
  theme_caleb() +
  ggplot2::theme(plot.title = ggplot2::element_text(color = COL_RED))
```

```{r projected-lineup, fig.cap="Warriors Projected Starting 5 w/ MPJ"}
# Helper: normalize player names for joins
name_key <- function(x)
  {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[[:punct:]]", " ") %>%
    stringr::str_replace_all("\\b(jr|sr|ii|iii|iv|v)\\b", "") %>%
    stringr::str_squish()
  }

# Lineup layout (x positions across, single y row)
lineup <- tibble::tibble(
  Player = c(
    "Stephen Curry",
    "Jimmy Butler",
    "Michael Porter Jr.",
    "Draymond Green",
    "Quinten Post"
  ),
  Pos = c("PG", "SG", "SF", "PF", "C"),
  x   = 1:5,
  y   = 1
) %>%
  dplyr::mutate(name_key = name_key(Player))

# Keyed lookup from existing `players` table 
players_keyed <- players %>%
  dplyr::mutate(name_key = name_key(player_name))

lineup_img <- lineup %>%
  dplyr::left_join(
    players_keyed %>% dplyr::select(name_key, player_id),
    by = "name_key"
  ) %>%
  dplyr::mutate(
    headshot = dplyr::if_else(
      !is.na(player_id),
      paste0("https://cdn.nba.com/headshots/nba/latest/1040x760/", player_id, ".png"),
      NA_character_
    )
  )

# Plot
ggplot2::ggplot(lineup_img, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(
    shape = 21,
    size = 20,
    fill = "#1D428A",
    color = "#f6f4ef",
    stroke = 1.4
  ) +
  ggimage::geom_image(
    ggplot2::aes(image = headshot),
    size = 0.15,
    asp = 1.1
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = Player),
    y = 0.9,
    size = 3.4,
    family = "poppins",
    fontface = "bold"
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = Pos),
    y = 1.1,
    size = 3.6,
    family = "poppins",
    color = "#6e6e6e"
  ) +
  ggplot2::scale_x_continuous(limits = c(0.5, 5.5)) +
  ggplot2::scale_y_continuous(limits = c(0.6, 1.4)) +
  ggplot2::labs(
    title = "Projected Warriors Starting Lineup",
    subtitle = "Steph – Jimmy – MPJ – Draymond – Post",
    caption = "Viz: @Rambzee_"
  ) +
  theme_caleb() +
  ggplot2::theme(
    axis.text  = ggplot2::element_blank(),
    axis.title = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    plot.title = ggplot2::element_text(color = "#1D428A")
  )
```

Stephen Curry: Burden Relief 

* MPJ provides much-needed elite off-ball shooting gravity alongside Steph, reducing help and switch leverage even when Curry is not the primary ball handler.
* Defenses can no longer load up early or cheat off weak-side shooters, preserving Steph’s space both on and off the ball.
* The result is fewer forced possessions, cleaner relocations, and a meaningful reduction in the offensive burden Curry has carried at age 37.

Jimmy Butler: Scoring Pressure Reduction

* MPJ absorbs scoring responsibility without demanding on-ball touches, allowing Butler to attack selectively rather than manufacture offense every trip.
* With a reliable high-efficiency scorer spacing the floor, Butler can conserve energy, punish mismatches, and maximize his impact in closing stretches.
* This aligns directly with Butler’s preferred offensive play-style at this point in his career. 

Draymond Green: Offensive Connectivity 

* MPJ enhances Draymond’s value as a facilitator by cleaning up short-roll reads and discouraging help from the corners.
* His shooting gravity increases the efficiency of handoffs, split actions, and delay offense, making Golden State’s orchestration more decisive.
* Porter also adds a legitimate defensive rebounding presence and functions as a solid team defender within structure, helping end possessions cleanly.

Overall Impact: Stability, Hierarchy, and Fewer Droughts

* MPJ instantly slots in as a clear second or third offensive option, establishing a cleaner hierarchy and reducing the need for secondary players to overextend.
* His presence directly lowers the likelihood of prolonged scoring droughts, an area where Golden State has been among the league’s most volatile offenses.
* This is additive talent meeting an optimized environment: elite, portable offense that stabilizes non-Steph minutes, amplifies Steph-led lineups, and restores the wing gravity the Warriors’ offense has historically depended on.
