---
title: "Keyonte George: 2025–26 Scoring Leap & Play-Type Profile"
author: "Caleb Ramsey"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(knitr)
library(kableExtra)
library(showtext)
library(sysfonts)
library(ggrepel)

font_add_google("Poppins", "poppins")
showtext_auto()

# Utah Jazz palette
JAZZ_NAVY   <- "#002B5C"
JAZZ_YELLOW <- "#F9A01B"
JAZZ_GREEN  <- "#00471B"

BG        <- "#f6f4ef"
TEXT_DARK <- "#141414"
GRID      <- "#dedad3"

theme_caleb_jazz <- function()
{
  theme_minimal(base_family = "poppins") +
    theme(
      plot.background  = element_rect(fill = BG, color = NA),
      panel.background = element_rect(fill = BG, color = NA),
      panel.grid.major = element_line(color = GRID, linewidth = 0.35),
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5, color = TEXT_DARK),
      plot.subtitle = element_text(
        size = 10, hjust = 0.5, color = "#2f2f2f",
        margin = margin(b = 6)
      ),
      axis.title = element_text(size = 12, face = "bold", color = TEXT_DARK),
      axis.text = element_text(size = 10, color = "#333333"),
      legend.position = "none",
      plot.caption = element_text(
        size = 8, color = "#6e6e6e", hjust = 0,
        margin = margin(t = 6)
      ),
      plot.margin = margin(10, 12, 14, 12)
    )
}

playtype_path <- "KeyonteGeorgePlaytypeTable.xlsx"
```

# Executive Summary

**Season context:** 
```{r season-context}
season_ctx <- tibble::tribble(
  ~Season, ~MPG, ~PPG, ~FGA, ~FG_PCT, ~`3PA`, ~`3P_PCT`, ~FTA, ~FT_PCT, ~AST, ~TO,
  "2024–25", 31.5, 16.8, 13.7, 0.391, 7.6, 0.343, 6.1, 0.818, 5.6, 2.7,
  "2025–26", 34.4, 24.6, 16.8, 0.455, 7.0, 0.374, 7.4, 0.903, 6.8, 3.4
)

kbl(
  season_ctx,
  format = "latex",
  booktabs = TRUE,
  digits = 2,
  caption = "Season context: counting stats comparison"
) |>
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position"),
    font_size = 10
  ) |>
  kableExtra::column_spec(1, bold = TRUE) |>
  kableExtra::column_spec(2:ncol(season_ctx), width = "1.2cm")
```

**Scoring spikes:** 9x 30-point games (6 total in first two seasons); 3x 35+ (39, 37, 37).

**What the play-type data says:**

- **Role distribution:** Keyonte’s offense is driven by **PnR Ball Handler volume** with secondary reps in **Transition**, **Spot Up**, and **Isolation**.
- **Efficiency pockets:** His best efficiency comes from **Spot Ups** and **Cuts**. These are high-value scoring channels that scale with star-level creation around him.
- **Engine reps are solid:** **PnR Ball Handler** and **Isolation** are not “free points,” but they are good enough at meaningful volume to support a real scoring leap.

# Data

```{r load-playtype}
pt_raw <- readxl::read_excel(playtype_path) %>%
  janitor::clean_names()

# Convert percents to proportions. Keep PPP numeric.
pt <- pt_raw %>%
  mutate(
    poss = readr::parse_number(poss),
    ppp  = readr::parse_number(ppp),
    percent_time = readr::parse_number(percent_time) / 100,
    across(
      c(
        fg_pct, efg_pct, ts_pct, to_pct, ft_pct,
        sf_pct, score_pct, fg2_pct, fg3_pct
      ),
      ~ readr::parse_number(.x) / 100
    ),
    tsa      = readr::parse_number(tsa),
    fta_fga = readr::parse_number(fta_fga),
    pa3_fga = readr::parse_number(pa3_fga)
  )

# Weighted averages (reference lines)
w_ppp  <- weighted.mean(pt$ppp,  w = pt$poss, na.rm = TRUE)
w_time <- weighted.mean(pt$percent_time, w = pt$poss, na.rm = TRUE)
```

```{r table-snapshot}
pt_small <- pt %>%
  transmute(
    Play = playtype,
    Poss = poss,
    `Time%` = percent_time,
    PPP = round(ppp, 2),
    Rating = ppp_rating,
    `TS%` = ts_pct,
    `TO%` = to_pct,
    `FTA/FGA` = fta_fga,
    `3PA/FGA` = pa3_fga
  ) %>%
  arrange(desc(`Time%`))

kbl(
  pt_small,
  format = "latex",
  booktabs = TRUE,
  caption = "Play-type snapshot (Keyonte George, 2025–26)"
) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 9
  ) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::column_spec(4, bold = TRUE)
```

```{r table-splits}
pt_splits <- pt %>%
  transmute(
    Play = playtype,
    Poss = poss,
    PPP = round(ppp, 2),
    `FG%` = fg_pct,
    `eFG%` = efg_pct,
    `2FG%` = fg2_pct,
    `3P%` = fg3_pct,
    `3PA/FGA` = pa3_fga
  ) %>%
  arrange(desc(Poss))

kbl(
  pt_splits,
  format = "latex",
  booktabs = TRUE,
  caption = "Shot profile by play type"
) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 9
  ) %>%
  kableExtra::column_spec(1, bold = TRUE)
```

# Figure 1: Efficiency by play type (PPP)

```{r plot_ppp}
pt_plot <- pt %>%
  mutate(playtype = forcats::fct_reorder(playtype, ppp))

ggplot(pt_plot, aes(x = ppp, y = playtype)) +
  geom_point(
    aes(size = poss),
    color = JAZZ_GREEN,
    alpha = 0.95
  ) +
  geom_vline(
    xintercept = w_ppp,
    linetype = "dashed",
    color = JAZZ_NAVY,
    linewidth = 0.9
  ) +
  scale_size(range = c(2, 8)) +
  labs(
    title = "Efficiency by play type (PPP)",
    subtitle = glue::glue(
      "Point size = volume | Dashed line = weighted avg PPP ({round(w_ppp, 2)})"
    ),
    x = "Points per possession (PPP)",
    y = NULL,
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb_jazz()
```

# Figure 2: Frequency by play type (%Time)

```{r plot-time}
pt_plot <- pt %>%
  mutate(playtype = forcats::fct_reorder(playtype, percent_time))

ggplot(pt_plot, aes(x = playtype, y = percent_time)) +
  geom_col(
    fill = JAZZ_NAVY,
    alpha = 0.92
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title = "Play-type frequency",
    subtitle = "Share of offensive possessions by play type",
    x = NULL,
    y = "% of possessions",
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb_jazz()
```

# Figure 3: Value map (Usage vs PPP)

```{r plot-value-map}
ggplot(pt, aes(x = percent_time, y = ppp)) +
  geom_vline(
    xintercept = w_time,
    linetype = "dashed",
    color = "#4b4b4b",
    linewidth = 0.7
  ) +
  geom_hline(
    yintercept = w_ppp,
    linetype = "dashed",
    color = "#4b4b4b",
    linewidth = 0.7
  ) +
  geom_point(
    aes(size = poss),
    color = JAZZ_GREEN,
    alpha = 0.75
  ) +
  ggrepel::geom_label_repel(
    aes(label = playtype),
    size = 3.0,
    seed = 42,
    force = 8,
    max.time = 2,
    max.iter = 5000,
    max.overlaps = Inf,
    point.padding = 0.35,
    box.padding = 0.45,
    min.segment.length = 0,
    segment.color = "grey55",
    segment.alpha = 0.7,
    fill = "white",
    label.size = 0.15,
    label.r = unit(0.12, "lines")
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_size(range = c(2, 7)) +
  labs(
    title = "Usage vs Efficiency (Play-Types)",
    subtitle = glue::glue(
      "Size: Volume | Dashed lines: Avg % Time ({scales::percent(w_time, 0.1)}) and PPP ({round(w_ppp, 2)})."
    ),
    x = "%Time",
    y = "PPP",
    caption = "Source: Synergy/SportsRadar | Viz: @Rambzee_"
  ) +
  theme_caleb_jazz()
```

# On/Off Context
```{r on-off}
wowy <- tibble::tribble(
  ~Split, ~MIN, ~OFF_RTG, ~DEF_RTG, ~NET_RTG, ~OFF_TS, ~OFF_TOV, ~OFF_ORB, ~DEF_TS, ~DEF_TOV, ~DEF_ORB,
  "On",   999,  118.7,    125.3,    -6.6,     60.1,   14.9,     30.0,     62.3,   12.0,     26.7,
  "Off",  345,  108.8,    115.7,    -6.9,     55.3,   17.4,     28.3,     57.6,   13.8,     26.8
)

wowy_delta <- tibble::tibble(
  Split   = "On - Off",
  MIN     = NA_real_,
  OFF_RTG = 9.9,
  DEF_RTG = 9.7,
  NET_RTG = 0.2,
  OFF_TS  = 4.8,
  OFF_TOV = -2.6,
  OFF_ORB = 1.7,
  DEF_TS  = 4.7,
  DEF_TOV = -1.8,
  DEF_ORB = -0.1
)

wowy_tbl <- dplyr::bind_rows(wowy, wowy_delta)

knitr::kable(
  wowy_tbl,
  format = "latex",
  booktabs = TRUE,
  digits = 1,
  caption = "Databallr WOWY (2025–26): team performance with Keyonte George ON vs OFF"
) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "hold_position", "scale_down"),
    font_size = 9
  ) %>%
  kableExtra::column_spec(1, bold = TRUE)
```

# Interpretation

## 1) What he runs

- **PnR Ball Handler is the engine** (largest share of possessions).
- Secondary mix is real: **Transition, Spot Up, Isolation** each matter.
- This supports both **creator scoring** and **easy points** buckets.

## 2) Where the efficiency comes from

- **Spot Up** is a major efficiency driver (high PPP and high 3-point rate).
- **Cuts** are extremely efficient in small volume.
- **Handoffs** produce strong PPP, lots of value in motion and DHO actions.

## 3) Impact

- **Offense improves with George on the floor:** +9.9 points/100 and +4.8 TS% (team efficiency spike).
- **Turnovers improve on offense with him on:** offensive TOV% drops by 2.6 points -- cleaner creation than the raw 3.4 TO/g implies.
- **Defense is the problem:** defensive rating worsens by +9.7 points/100 and opponent TS% rises by +4.7 points with him on the floor.
- **Defensive “event” pressure declines:** opponent TOV% drops by 1.8 points with him on -- fewer forced mistakes by the Jazz defense.
- **Net impact is basically neutral in this lens:** +0.2 net because the offensive gains are almost fully canceled by defensive losses.
- **Context note:** WOWY is role + lineup dependent, but the data is consistent with the eye test: offense-driving guard, defensive liability.

# Final Summary
Keyonte George’s 2025–26 leap is driven by legit scoring growth. His top-line production (24.6 PPG on 45.5/37.4/90.3 splits with 6.8 AST) is supported by a play-type mix where high-value buckets pop: Spot Ups and Cuts are areas of elite efficiency in his game, while high-volume creator reps in PnR and Isolation remain solid enough to sustain primary usage. The tradeoff is defensive. On-off splits show a major offensive lift with him on the floor (+9.9 OFF RTG, +4.8 TS%), but an equally severe defensive collapse (+9.7 DEF RTG, +4.7 opponent TS%), leaving the net close to neutral. The All-Star argument is simple: the offense is real; the swing factor is whether Utah can scheme and insulate the defense.
